---
title_meta  : Chapter 1
title       : Transformations in MRS
description : Introduction to transforming variables and subsetting XDF files.
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise key:3f36e6ffa4

## Introduction to Transformations

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220973
```

--- type:VideoExercise key:7f9d201326

## Your First Transformation

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220976
```

--- type:NormalExercise xp:50 key:a89b06f858

## Exercise 1: Transforming a Single Variable
  
Convert engine displacement from cubic inches to liters


*** =instructions

Let's get the hang of the basic transform syntax. In this exercise, you'll
work with a dataset of automobile specifications, including miles per gallon,
number of cylinders, transmission type, and so on.

It's an old dataset, however. For example, the engine displacement
(a measure of engine size) is measured in cubic inches. Nowadays, many people
would prefer liters. In base R, that conversion would be:

`disp * 0.016`

Using `rxDataStep`, create a new variable on the XDF dataset `mtcarsXdf`
called `disp_liters` that converts `disp` from cubic inches to liters.

*** =hint
The basic structure you need is:    

     rxDataStep(
         inData = mtcarsXdf,
         outFile = mtcarsXdf,
         transforms =
             list(______ = _______),
         overwrite = TRUE
     )

*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Convert mtcars to XDF
mtcarsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = cbind(model = rownames(mtcars), mtcars), 
         outFile = mtcarsXdf)

# Create a second copy to see if the file is modified at all
mtcarsXdfOriginal <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = cbind(model = rownames(mtcars), mtcars),
         outFile = mtcarsXdfOriginal)

```

*** =sample_code

```{r,eval=FALSE}
# Examine the variables on the dataset
rxGetVarInfo(mtcarsXdf)
```

*** =solution

```{r,eval=FALSE}
rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsXdf,
           transforms = list(disp_liters = disp * 0.016),
           overwrite = TRUE
)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxDataStep", 
                             not_typed_msg = "You'll have to use the function rxDataStep to make this transformation - start there.")

test_student_typed("mtcarsXdf",
                   not_typed_msg = "Make sure you're reading the data from mtcarsXdf, and writing the modified data back to mtcarsXdf.")

test_student_typed("transforms",
                   not_typed_msg = "Make sure to use the 'transforms' argument in rxDataStep.")

test_student_typed("transforms = list(",
                   not_typed_msg = "Remember: the transforms argument takes a list formatted like: 'list(name_of_new_var = old_var * 20)'.")

test_student_typed("overwrite = TRUE",
                   not_typed_msg = "Be sure to set the argument overwrite to TRUE - otherwise, your transformation won't be written to the XDF file.")


test_that("mtcarsXdf has been modified in some way, even if not totally correct", {
    expect_false(identical(rxDataStep(mtcarsXdf), 
                           rxDataStep(mtcarsXdfOriginal)),
                 failure_msg = "The mtcarsXdf dataset hasn't been changed - be sure to set 'overwrite = TRUE'.")
})

test_that("disp_liters has been created", {
    expect_true("disp_liters" %in% names(mtcarsXdf),
                failure_msg = "It looks like you haven't created a variable called 'disp_liters' on the mtcarsXdf dataset. Try rxGetVarInfo(mtcarsXdf) to see what variables do exist on the dataset.")
})


test_that("disp_liters = disp * 0.016", {
    expect_equivalent(object = rxDataStep(mtcarsXdf)$disp_liters,
                      expected = mtcars$disp * 0.016,
                      failure_msg = "Your conversion of cubic inches to liters doesn't seem right - take another look, and be sure you're using 0.016 as your conversion ratio."
    )
})

success_msg("Well done! You've made a simple transformation.")

```

*** =skills
1, 3

--- type:NormalExercise xp:50 key:241226b296

##  Exercise 2: Transforming a Single Variable
  
Convert car weight from thousands of pounds to thousands of kilograms


*** =instructions

Let's convert another variable to metric. The weight of the cars in `mtcarsXdf`
is recorded in thousands of pounds - for example, the Datsun 710 has a `wt`
value of 3.85 (meaning it weights 3,850 pounds). That's about 1.75 thousand
kilograms. In open-source R, that conversion would look like this:

`wt * 0.454`

Use `rxDataStep` to create a new variable on the XDF dataset `mtcarsXdf`
called `wt_kg` that converts `wt` from thousands-of-pounds to
thousands-of-kilograms.

*** =hint

The basic structure you need is:    

     rxDataStep(inData = mtcarsXdf,
         outFile = mtcarsXdf,
         transforms = 
             list(______ = _______),
         overwrite = TRUE
     )

*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Convert mtcars to XDF
mtcarsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = cbind(model = rownames(mtcars), mtcars), 
         outFile = mtcarsXdf)

# Create a second copy to see if the file is modified at all
mtcarsXdfOriginal <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = cbind(model = rownames(mtcars), mtcars), 
         outFile = mtcarsXdfOriginal)

```

*** =sample_code

```{r,eval=FALSE}
# Examine the variables on the dataset
rxGetVarInfo(mtcarsXdf)
```


*** =solution

```{r,eval=FALSE}

rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsXdf,
           transforms = list(wt_kg = wt * 0.454),
           overwrite = TRUE
)

```


*** =sct

```{r,eval=FALSE}
test_error()

test_student_typed("rxDataStep", 
                             not_typed_msg = "You'll have to use the function rxDataStep to make this transformation - start there.")

test_student_typed("mtcarsXdf",
                   not_typed_msg = "Make sure you're reading the data from mtcarsXdf, and writing the modified data back to mtcarsXdf.")

test_student_typed("transforms",
                   not_typed_msg = "Make sure to use the 'transforms' argument in rxDataStep.")

test_student_typed("transforms = list(",
                   not_typed_msg = "Remember: the transforms argument takes a list formatted like: 'list(name_of_new_var = old_var * 20)'.")

test_student_typed("overwrite = TRUE",
                   not_typed_msg = "Be sure to set the argument overwrite to TRUE - otherwise, your transformation won't be written to the XDF file.")


test_that("mtcarsXdf has been modified in some way, even if not totally correct", {
    expect_false(identical(rxDataStep(mtcarsXdf), rxDataStep(mtcarsXdfOriginal)),
                 failure_msg = "The mtcarsXdf dataset hasn't been changed - be sure to set 'overwrite = TRUE'.")
})

test_that("wt_kg = wt * 0.454", {
    expect_equivalent(object = rxDataStep(mtcarsXdf)$wt_kg,
                      expected = mtcars$wt * 0.454,
    failure_msg = "Your conversion of pounds to kilograms doesn't seem right - take another look, and be sure you're using 0.454 as your conversion ratio."
    )
})

success_msg("Well done! Hopefully these simple transformations are starting to feel a bit more natural.")

```

*** =skills
1, 3


--- type:VideoExercise key:9dc3f8dfba

## Multiple Transformations 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220971
```

--- type:NormalExercise xp:50 key:0ac5f9cf6c

## Exercise 3: Multiple Transformations
  
Convert several variables to metric


*** =instructions

Most of the time, if you're converting one variable to metric, you'll want to
convert several. In this exercise, convert:

 - `mpg` from miles per gallon to liters per 100 kilometers,
 - `disp` from cubic inches to liters, and
 - `wt` from thousands of pounds to thousands of kilograms. In base R, that would
entail these three expressions:

 - `lp100km = 235.215 / mpg`
 - `disp_liters = disp * 0.016`
 - `wt_kg = wt * 0.454`
 
 The MRS syntax is nearly the same as a single transformation; just separate
 each variable with a comma inside the transforms `list`.
 
 Add these new variables to the existing `mtcarsXdf` file.


*** =hint
All your transformations should be in one list, like this: 

    list(lp100km = 235.215 / ____,
         disp_liters = _____ * 0.016, 
         wt_kg = _____ * 0.454)

... which you pass to `rxDataStep`'s `transforms` argument.

*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Convert mtcars to XDF
mtcarsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = cbind(model = rownames(mtcars), mtcars),
         outFile = mtcarsXdf)

# Create a second copy to see if the file is modified at all
mtcarsXdfOriginal <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = cbind(model = rownames(mtcars), mtcars),
         outFile = mtcarsXdfOriginal)

```

*** =sample_code

```{r,eval=FALSE}
# Examine the variables on the dataset
rxGetVarInfo(mtcarsXdf)
```

*** =solution

```{r,eval=FALSE}
rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsXdf,
           transforms = list(lp100km = 235.215 / mpg,
                             disp_liters = disp * 0.016,
                             wt_kg = wt * 0.454
           ),
           overwrite = TRUE
)

```


*** =sct

```{r,eval=FALSE}
test_error()

test_student_typed("rxDataStep", 
                             not_typed_msg = "You'll have to use the function rxDataStep to make this transformation - start there.")

test_student_typed("mtcarsXdf",
                   not_typed_msg = "Make sure you're reading the data from mtcarsXdf, and writing the modified data back to mtcarsXdf.")

test_student_typed("transforms",
                   not_typed_msg = "Make sure to use the 'transforms' argument in rxDataStep.")

test_student_typed("transforms = list(",
                   not_typed_msg = "Remember: the transforms argument takes a list formatted like: 'list(name_of_new_var = old_var * 20)'.")

test_student_typed("overwrite = TRUE",
                   not_typed_msg = "Be sure to set the argument overwrite to TRUE - otherwise, your transformation won't be written to the XDF file.")


test_that("mtcarsXdf has been modified in some way, even if not totally correct", {
    expect_false(identical(rxDataStep(mtcarsXdf), rxDataStep(mtcarsXdfOriginal)),
                 failure_msg = "The mtcarsXdf dataset hasn't been changed - be sure to set 'overwrite = TRUE'.")
})

test_that("lp100km = 235.215 / mpg", {
    expect_equivalent(object = rxDataStep(mtcarsXdf)$lp100km,
                      expected = 235.215 / mtcars$mpg,
                      failure_msg = "Your conversion of miles-per-gallon to liters/100km doesn't seem right - take another look."
    )
})

test_that("disp_liters = disp * 0.016", {
    expect_equivalent(object = rxDataStep(mtcarsXdf)$disp_liters, 
                      expected = mtcars$disp * 0.016,
                      failure_msg = "Your conversion of disp from cubic inches to liters doesn't seem right - take another look."
    )
})

test_that("wt_kg = wt * 0.454", {
    expect_equivalent(object = rxDataStep(mtcarsXdf)$wt_kg,
                      expected = mtcars$wt * 0.454,
                      failure_msg = "Your pounds-to-kilograms conversion doesn't seem right - take another look."
    )
})
  
success_msg("Well done! You've got the hang of multiple transformations.")
```

*** =skills
1, 3



--- type:VideoExercise key:b6b08a6163

## Dropping Variables 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220977
```


--- type:NormalExercise xp:50 key:72cb160561

## Exercise 4: Dropping Variables
  
Drop the variables that use imperial measures


*** =instructions

Now that you've got the metric variables, create a new XDF file called
`mtcarsMetric` and drop the three imperial variables we've replaced
(`mpg`, `wt`, and `disp`).
In `rxDataStep`, you can use the argument `varsToDrop` to name 
specific variables to be dropped from the dataset, or use the argument
`varsToKeep` to drop all the variables you *don't* name.

*** =hint

You need to pass a vector of the variable names you want to drop to the
argument `varsToDrop` - it'll look something like this:

    varsToDrop = c("var1", "var2")



*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Convert mtcars to XDF
mtcarsXdf <- RxXdfData(tempfile(fileext = ".xdf"))
rxImport(inData = cbind(model = rownames(mtcars), mtcars),
         outFile = mtcarsXdf)

# Create a second copy to see if the file is modified at all
mtcarsXdfOriginal <- RxXdfData(tempfile(fileext = ".xdf"))
rxImport(inData = cbind(model = rownames(mtcars), mtcars), 
         outFile = mtcarsXdfOriginal)


# Convert to metric
rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsXdf,
           transforms = list(lp100km = 235.215 / mpg,
                             disp_liters = disp * 0.016,
                             wt_kg = wt * 0.454
           ),
           overwrite = TRUE
)

```


*** =sample_code

```{r,eval=FALSE}
# Create a new XDF for the subset. The tempfile() function is a good way to do this.
mtcarsMetric <- tempfile(fileext = ".xdf")
```

*** =solution

```{r,eval=FALSE}
# Create a new XDF for the subset. The tempfile() function is a good way to do this.
mtcarsMetric <- tempfile(fileext = ".xdf")

rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsMetric,
           varsToDrop = c("mpg", "wt", "disp")
)

```


*** =sct

```{r,eval=FALSE}


test_error()


test_student_typed("rxDataStep", 
                   not_typed_msg = "You'll have to use the function rxDataStep to drop the variables - start there.")

test_student_typed("mtcarsXdf",
                   not_typed_msg = "Make sure you're reading the data from mtcarsXdf.")

test_student_typed("mtcarsMetric",
                   not_typed_msg = "Make sure you're writing the subset of variables to mtcarsMetric.")


test_student_typed("varsToDrop",
                   not_typed_msg = "Make sure to use the 'varsToDrop' argument in rxDataStep.")


test_that("The imperial variables have been removed.", {
    expect_false(any(c("mpg", "wt", "disp") %in% rxGetVarNames(mtcarsMetric)),
                 failure_msg = "At least one of mpg, wt, and disp is still on mtcarsMetric. Check your code and try again."
        )
})

test_that("The metric variables haven't been removed.", {
    expect_true(all(c("lp100km", "wt_kg", "disp_liters") %in% rxGetVarNames(mtcarsMetric)),
                failure_msg = "It looks like you dropped one of the metric variables (lp100km, wt_kg, disp_liters) by accident! Check your code and try again."
        )
})

  
success_msg("Success! You can drop unnecessary variables. The argument varsToKeep works in basically the same way, except it retains the named variables. Use whichever is most convenient.")
```

*** =skills
1, 3




--- type:VideoExercise key:67eba8f335

## Subsetting

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220972
```


--- type:NormalExercise xp:50 key:f576918108

## Exercise 5: Subsetting Rows
  
Subset to efficient cars with four gears


*** =instructions

Now imagine that a customer is looking for a relatively efficient car with four gears.
Subset `mtcarsXdf` to include just cars with four gears and that can go 100km
on less than 10 liters of gas (remember, this dataset is from 1974). In open-source
R, you might use the `subset` function:

    subset(mtcars, 
           gear == 4 & lp100km < 10)

In Microsoft R Server, you can use `rxDataStep` to subset an XDF file.
Use `rxDataStep`'s argument `rowSelection` to select the five cars in `mtcarsXdf`
that meet these criteria.

*** =hint

Remember to use the `&` operator to link the two comparisons - `gear == 4` and
`lp100km < 10` - into one expression.

*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Convert mtcars to XDF
mtcarsXdf <- RxXdfData(tempfile(fileext = ".xdf"))
rxImport(inData = cbind(model = rownames(mtcars), mtcars), 
         outFile = mtcarsXdf)

# Create a second copy to see if the file is modified at all
mtcarsXdfOriginal <- RxXdfData(tempfile(fileext = ".xdf"))
rxImport(inData = cbind(model = rownames(mtcars), mtcars),
         outFile = mtcarsXdfOriginal)


# Convert to metric
rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsXdf,
           transforms = list(lp100km = 235.215 / mpg,
                             disp_liters = disp * 0.016,
                             wt_kg = wt * 0.454
           ),
           overwrite = TRUE
)

```


*** =sample_code

```{r,eval=FALSE}
# Examine the variables on the dataset
rxGetVarInfo(mtcarsXdf)
```


*** =solution

```{r,eval=FALSE}

rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsXdf,
           rowSelection = gear == 4 & lp100km < 10,
           overwrite = TRUE
)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxDataStep", 
                             not_typed_msg = "You'll have to use the function rxDataStep to make this transformation - start there.")

test_student_typed("mtcarsXdf",
                   not_typed_msg = "Make sure you're reading the data from mtcarsXdf, and writing the modified data back to mtcarsXdf.")

test_student_typed("rowSelection",
                   not_typed_msg = "Make sure to use the 'rowSelection' argument in rxDataStep.")


test_student_typed("overwrite = TRUE",
                   not_typed_msg = "Be sure to set the argument overwrite to TRUE - otherwise, your transformation won't be written to the XDF file.")


test_expression_output(rxGetInfo(mtcarsXdf)$numRows < 32,
                       incorrect_msg = "mtcarsXdf still has all 32 records - are you sure you set overwrite = TRUE?")


test_expression_output(rxDataStep(mtcarsXdf)$lp100km < 10,
                incorrect_msg = "Some of the cars in your subset have lp100km values greater than 10 - check your code and try again.")

test_expression_output(rxDataStep(mtcarsXdf)$gear == 4,
                incorrect_msg = "Some of the cars in your subset have three or five gears - check your code and try again.")


  
success_msg("Well done! You've successfully transformed variables, dropped variables, and subset variables. On to the next lesson!")

```

*** =skills
1, 3


