---
title       : Simple Imports
description : Create an XDF file using the rxImport function
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained

--- type:VideoExercise

## Configuring rxImport 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141145807
```









--- type:VideoExercise

## colClasses and colInfo

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141145804
```






--- type:NormalExercise xp:50


## Exercise 5: rxImport Options

You can specify various options when importing data with rxImport, such as how 
to encode variables, what their names and factor levels should be, and whether 
to keep or drop whole variables in the XDF file.

*** =instructions

Use the `rxImport` function as before, but this time you will add a few 
arguments to specify how you want the data imported:  

 - The `varsToKeep` argument takes a vector of variable names that you want to 
   import into the xdf file. *Use `varsToKeep` to import the `creditScore`, 
   `year`, and `default` variables.*

 - The `colClasses` argument takes a vector which specifies the column types to
   use when importing the data. *The `default` variable describes whether a 
   person defaulted on their mortgage or not, so you should import it as a 
   `logical` data type.*

 - The `colInfo` argument lets you specify many import options for the variables
   in your dataset. It looks something like this:
   
    colInfo = list(month = list(type = "factor", newName = "default_month")),
   
`colInfo` takes a `list` of variable names (`colInfo = list(month = `), and 
each variable gets another list of options for importing that variable. In
the example above, we're using the original `month` variable to create a new 
`factor` variable called `"default_month"`. *Use `colInfo` to create a new
`factor` out of the `year` variable called `default_year`*.

 - The `overwrite` argument indicates whether an existing dataset should be 
   overwritten. Be careful when setting this to TRUE! If TRUE, a data set that 
   has the same name as **outFile** will be overwritten with the new imported 
   file. *Overwrite the "mortDefaultSmall2009.xdf" file you created in the last 
   exercise with the new version created by setting `varsToKeep`, `colClasses`,
   and `colInfo`*.


After you finish, use `rxGetInfo` to look at the new `mortDefaultXdf` data set.


*** =hint

Remember to enter the `varsToKeep` and `colClasses` arguments as character 
vectors.

 - `varsToKeep` takes a simple character vector of variable names, like this:

    varsToKeep = c("creditScore", "year", "default")

 - `colClasses` takes a *named* character vector, with the name of the variable
   on the left and the class it should be on the right, in quotes. For example:
 
    colClasses = c(default = "logical")

 - The `colInfo` argument takes a nested list. The first list is of each variable
   you want to specify (e.g., `year`); then each of those variables has its own list of 
   details (e.g., `type` and `newName`).
   It should look like this:

    list(year = list(type = "factor", newName = "default_year"))



*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Create the XDF from the last exercise so there's something to overwrite
file.remove("mortDefaultSmall2009.xdf")
mortDefaultCsv <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall2009.csv")
mortDefaultXdf <- "mortDefaultSmall2009.xdf"
rxImport(inData = mortDefaultCsv, outFile = mortDefaultXdf)
```

*** =solution
```{r, eval=FALSE}
# Store the file paths for the csv and xdf files as objects
mortDefaultCsv <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall2009.csv")
mortDefaultXdf <- "mortDefaultSmall2009.xdf"

# Use the varsToKeep, colClasses, colInfo, and overwrite arguments in rxImport
rxImport(inData = mortDefaultCsv, outFile = mortDefaultXdf,
         varsToKeep = c("creditScore", "year", "default"),
         colClasses = c(default = "logical"),
         colInfo = list(year = list(type = "factor", newName = "default_year")),
         overwrite = TRUE
)

# Look at the new default data file
rxGetInfo(mortDefaultXdf, getVarInfo = TRUE)
```

*** =sample_code
```{r, eval = F}
# Store the file paths for the csv and xdf files as objects
mortDefaultCsv <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall2009.csv")
mortDefaultXdf <- "mortDefaultSmall2009.xdf"

# Use varsToKeep, colClasses, colInfo, and overwrite arguments in rxImport

rxImport(inData = mortDefaultCsv, outFile = mortDefaultXdf,
         varsToKeep = ______________________,
         colClasses = ______________________,
         colInfo = ______________________,
         overwrite = TRUE
)

# Look at the new default data file
rxGetInfo(mortDefaultXdf, getVarInfo = TRUE)
```

*** =sct

```{r,eval=FALSE}

test_error()


# Basic checks for the function and arguments
test_student_typed("rxImport",
                   not_typed_msg = "Be sure to call the 'rxImport' function!")

test_student_typed("varsToKeep",
                   not_typed_msg = "Be sure to use the 'varsToKeep' argument to select just the 'creditScore', 'year', and 'default' variables.")

test_student_typed("colClasses",
                   not_typed_msg = "Use the 'colClasses' argument to convert the 'default' variable from an integer into a logical variable.")

test_student_typed("colInfo",
                   not_typed_msg = 'Use the \'colInfo\' argument to convert the \'year\' variable into a factor. It takes a list with another list inside, like this: list(year = list(type = "factor", newName = "default_year"))')

test_student_typed('rxGetInfo',
                   not_typed_msg = "This is a complicated set of changes, so be sure to use rxGetInfo on mortDefaultXdf to check the results.")


# Check the results
test_expression_output(rxGetVarNames(mortDefaultXdf),
                       incorrect_msg = "Something\'s wrong - your final dataset should have three variables: creditScore, default_year, and default. Try using 'rxGetVarInfo(mortDefaultXdf)' to see which variables are on your dataset.")

test_expression_output(rxGetVarInfo(mortDefaultXdf)$default$varType,
                       incorrect_msg = "It looks like you didn\'t convert 'default' to be a logical variable - check your 'colClasses' argument.")

test_expression_output(rxGetVarInfo(mortDefaultXdf)$default_year$varType,
                       incorrect_msg = "Be sure you're converting 'year' into a factor using the 'colInfo' argument.")

test_expression_output(rxGetInfo(mortDefaultXdf),
                       incorrect_msg = "Your 'inner_joined' data.frame doesn't seem to have the right number of rows - check your code again.")


success_msg("Well done! This exercise only scratches the surface of the options in rxImport, so be sure to explore its documentation a little more.")

```



