---
title_meta  : Chapter 3
title       : Factors and Conclusion
description : Create and modify factors with `rxFactors`
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise key:fc7029784a

## rxFactors 

*** =video_link

```{r,eval=FALSE}  
//player.vimeo.com/video/141147219
```

--- type:NormalExercise xp:50 key:035acc9ee0

## Exercise 5: Creating Factors with rxFactors

Use `rxFactors` to convert the `Year` variable into a factor.


*** =instructions

There are a couple of ways to create factors in MRS, but the `rxFactors` function
is the most powerful (and often the easiest). To simply create a factor from an
existing variable, you can use the `factorInfo` argument.

`factorInfo` takes a list. Each element of that list should be another list that
specifies the properties of the factor you're creating. Here's a simple example:

    factorInfo = list(newVariableName = list(varName = "variableToConvert"))

 - Use `rxFactors` to create a new variable called `FactorYear` from the existing
   variable `Year`.



*** =hint

 - Replace `newVariableName` with the name of the variable you want to create
   (in this case, `FactorYear`)

 - Replace `variableToConvert` with the name of the variable you want to make into
   a factor (in this case, `Year`)

 - Note that `newVariableName` isn't quoted, but `variableToConvert` is. This is
   just a quirk of R's list syntax, but it's important to remember.



*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
set.seed(2015)

djia_source <- RxXdfData( file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf") )

# Add a tempfile for sorting, and one for dropping the sorting variable
djia_temp <- tempfile(fileext = ".xdf")
djia_scrambled <- tempfile(fileext = ".xdf")

# Add a random uniform
rxDataStep(inData = djia_source,
           outFile = djia_temp,
           transforms = list(sortvar = runif(.rxNumRows))
)

# Sort by sortvar
rxSort(inData = djia_temp, 
       outFile = djia_temp,
       sortByVars = "sortvar",
       overwrite = TRUE)

# Drop sortvar
rxDataStep(inData = djia_temp,
           outFile = djia_scrambled,
           varsToDrop = "sortvar",
           overwrite = TRUE)

# Cleanup
file.remove(djia_temp)

```


*** =sample_code

```{r,eval=FALSE}

# Create a path to a new XDF file for the output
djia_factor_year <- tempfile(fileext = ".xdf")


# Examine the variable types in the current dataset
rxGetVarInfo(djia_scrambled)


# Create a factor from Year variable
rxFactors(inData = djia_scrambled, 
          outFile = djia_factor_year, 
          factorInfo = ______________
)


# Examine the output
rxGetVarInfo(djia_factor_year)

```

*** =solution

```{r,eval=FALSE}

# Create a path to a new XDF file for the output
djia_factor_year <- tempfile(fileext = ".xdf")

# View current dataset
rxGetVarInfo(djia_scrambled)

# 
# Create factor from Year variable
rxFactors(inData = djia_scrambled, 
          outFile = djia_factor_year, 
          factorInfo = list(FactorYear = list(varName = "Year"))
)


# View new factor variable
rxGetVarInfo(djia_factor_year)

```


*** =sct

```{r,eval=FALSE}



test_error()

test_student_typed("rxFactors",
                   not_typed_msg = "It doesn't appear that you made a call to rxFactors - start there!")

test_student_typed("djia_scrambled",
                   not_typed_msg = "Be sure to use 'djia_scrambled' as your dataset; it's already been created for you.")

test_student_typed("djia_factor_year",
                   not_typed_msg = "Be sure to pass 'djia_factor_year' to the outFile argument to store your results.")

test_student_typed("factorInfo = ",
                   not_typed_msg = "You'll need to use rxFactors' factorInfo argument.")

test_student_typed("factorInfo = list(FactorYear = ",
                   not_typed_msg = "The factorInfo argument takes a list - double-check your syntax.")

test_student_typed("FactorYear = list(varName = 'Year')",
                   not_typed_msg = 'You need to use a second list to let rxFactors know which variable to convert - like this: factorInfo = list(FactorYear = list(varName = "Year"))')


# Check the results
test_expression_result("FactorYear" %in% rxGetVarNames(djia_factor_year),
                       incorrect_msg = "It doesn't look like the variable FactorYear was created on the djia_factor_year dataset - try again.")

test_expression_output(rxGetVarInfo(djia_factor_year)[["FactorYear"]][["varType"]],
                       incorrect_msg = "FactorYear was created, but it doesn't seem to be a factor - check your code again.")


success_msg("Good work. That's the simplest way to create factors from other variables in MRS.")




```

*** =skills
1, 3






--- type:NormalExercise xp:50 key:6bcfe6fe36

## Exercise 6: Change a Factor's Sort Order

Use `rxFactors` to change the sort order of a factor's levels


*** =instructions

By default, the levels in a factor created using `rxFactors` are created in the
same order they appear in the data, which can result in strange level orders. For
example, using the `FactorYear` variable you created in the last exercise puts
1974 before 1973, followed by 2010, 2006, 1929... 

Because the correct order
of the values of `FactorYear` is their alphanumeric sort order, it's easiest to
use the `sortLevels` option in `rxFactors` to automatically sort the levels.
For example, we could sort the levels of `exampleFactor` like this:

    factorInfo = list( exampleFactor = list(sortLevels = TRUE) )


The next exercise will walk you through how to create custom sort orders.

 - Use `rxFactors` with the `sortLevels` argument to put the levels of `FactorYear`
   in the correct order.


*** =hint

 - Since there are 83 levels in `FactorYear` (1929-2010), and because the years
   should sort correctly, `sortLevels` is probably the best choice.
 - If you need to use `newLevels`, remember that R has a wide variety of commands
   that make creating vectors easier (try `1929:2010`).



*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
set.seed(2015)

djia_source <- RxXdfData( file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf") )

# Add a tempfile for sorting, and one for dropping the sorting variable
djia_temp <- tempfile(fileext = ".xdf")
djia_factor_year <- tempfile(fileext = ".xdf")

# Add a random uniform
rxDataStep(inData = djia_source,
           outFile = djia_temp,
           transforms = list(sortvar = runif(.rxNumRows))
)

# Sort by sortvar
rxSort(inData = djia_temp, 
       outFile = djia_temp,
       sortByVars = "sortvar",
       overwrite = TRUE
)


# Create factor from Year variable
rxFactors(inData = djia_temp, 
          outFile = djia_factor_year, 
          varsToDrop = "sortvar",
          factorInfo = list(FactorYear = list(varName = "Year"))
)

# Cleanup
file.remove(djia_temp)
```


*** =sample_code

```{r,eval=FALSE}

# As you can see, the levels in the FactorYear variable aren't sorted correctly
rxGetVarInfo(djia_factor_year)

# And it has real consequences for analysis - these results are all out of order:
rxSummary(Close ~ FactorYear, data = djia_factor_year)

# So - use rxFactors to sort the levels of FactorYear
rxFactors(inData = djia_factor_year, 
          outFile = djia_factor_year, 
          factorInfo = list( FactorYear = ___________,
          overwrite = TRUE)


# Check the results; 1929 should now be the first result, and 2010 the last:
rxSummary(Close ~ FactorYear, data = djia_factor_year)

```

*** =solution

```{r,eval=FALSE}

# As you can see, the levels in the FactorYear variable aren't sorted correctly
rxGetVarInfo(djia_factor_year)

# And it has real consequences for analysis - these results are all out of order:
rxSummary(Close ~ FactorYear, data = djia_factor_year)

# So - use rxFactors to sort the levels of FactorYear
rxFactors(inData = djia_factor_year, 
          outFile = djia_factor_year, 
          factorInfo = list( FactorYear = list(sortLevels = TRUE) ), 
          overwrite = TRUE)


# Check the results; 1929 should now be the first result, and 2010 the last:
rxSummary(Close ~ FactorYear, data = djia_factor_year)

```


*** =sct

```{r,eval=FALSE}



test_error()

test_student_typed("rxFactors",
                   not_typed_msg = "It doesn't appear that you made a call to rxFactors - start there!")

test_student_typed("djia_factor_year",
                   not_typed_msg = "Be sure to pass 'djia_factor_year' to the outFile argument to store your results.")

test_student_typed("factorInfo = ",
                   not_typed_msg = "You'll need to use rxFactors' factorInfo argument.")

test_student_typed("factorInfo = list(FactorYear = ",
                   not_typed_msg = "The factorInfo argument takes a list - double-check your syntax.")


# Check the results
test_expression_result("FactorYear" %in% rxGetVarNames(djia_factor_year),
                       incorrect_msg = "It doesn't look like the variable FactorYear was created on the djia_factor_year dataset - try again.")

test_expression_output(rxGetVarInfo(djia_factor_year)[["FactorYear"]][["varType"]],
                       incorrect_msg = "FactorYear was created, but it doesn't seem to be a factor - check your code again.")

test_expression_output(rxGetVarInfo(djia_factor_year)[["FactorYear"]][["varType"]][["levels"]],
                       incorrect_msg = "Your factor levels don't seem to be in the right order. Try again.")




success_msg("Nicely done.")


```

*** =skills
1, 3








--- type:NormalExercise xp:50 key:b2f95abf6b

## Exercise 7: Custom Sorting Factors

Use the `newLevels` argument to reverse the levels of `FactorYear`.



*** =instructions

When you need your levels in anything other than alphanumeric order, use the
`newLevels` argument to set the order. `newLevels` takes a character vector that
specifies *all* of the levels in the factor, in the desired order.

In this exercise, use `newLevels` to reverse the order of the `FactorYear` levels,
so that 2010 is the first level and 1928 is the last.


 - `newLevels` will only take a *character* vector, so you may need to convert
   your vector using `as.character`
 - For factors with a large number of levels, it can be tedious to type out every one,
so be sure to take advantage of R's tools for working with vectors (`seq()` will
be particularly useful to you here).


*** =hint

 - To generate the full list of levels, use the function 
 `seq(from = 2010, to = 1928)` - or its shortcut syntax, `2010:1928`


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

djia_source <- RxXdfData( file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf") )
djia_scrambled <- tempfile(fileext = ".xdf")

rxSort(inData = djia_source, outFile = djia_scrambled,
       sortByVars = "Volume",
       overwrite = TRUE)

djia_factor_year <- tempfile(fileext = ".xdf")

# Create factors from Year variable
rxFactors(inData = djia_scrambled, outFile = djia_factor_year, 
          factorInfo = list(FactorYear = list(varName = "Year")), 
          overwrite = TRUE)
```


*** =sample_code

```{r,eval=FALSE}

# Create the custom sort order
reversed_years <- __________

# Sort using the custom order
rxFactors(inData = djia_factor_year, outFile = djia_factor_year,
          factorInfo = list( FactorYear = list(__________ = __________) ),
          overwrite = TRUE)

# Check the results
rxGetVarInfo(djia_factor_year)

```



*** =solution

```{r,eval=FALSE}

# Create the custom sort order
reversed_years <- as.character(2010:1928)

# Sort using the custom order
rxFactors(inData = djia_factor_year, outFile = djia_factor_year,
          factorInfo = list( FactorYear = list(newLevels = reversed_years) ),
          overwrite = TRUE)

# View the data
rxGetVarInfo(djia_factor_year)

```


*** =sct

```{r,eval=FALSE}



test_error()

test_student_typed("rxFactors",
                   not_typed_msg = "It doesn't appear that you made a call to rxFactors - start there!")

test_student_typed("djia_factor_year",
                   not_typed_msg = "Be sure to pass 'djia_factor_year' to the outFile argument to store your results.")

test_student_typed("factorInfo = ",
                   not_typed_msg = "You'll need to use rxFactors' factorInfo argument.")

test_student_typed("factorInfo = list(FactorYear = ",
                   not_typed_msg = "The factorInfo argument takes a list - double-check your syntax.")



# Check the results
test_expression_result("FactorYear" %in% rxGetVarNames(djia_factor_year),
                       incorrect_msg = "It doesn't look like the variable FactorYear was created on the djia_factor_year dataset - try again.")

test_expression_output(rxGetVarInfo(djia_factor_year)[["FactorYear"]][["varType"]],
                       incorrect_msg = "FactorYear was created, but it doesn't seem to be a factor - check your code again.")

test_expression_output(rxGetVarInfo(djia_factor_year)[["FactorYear"]][["varType"]][["levels"]],
                       incorrect_msg = "Your factor levels don't seem to be in the right order. Try again.")




success_msg("Well done. These lists can get complex, but you can always build them up one piece at a time.")

```

*** =skills
1, 3




--- type:NormalExercise xp:50 key:1e326c710f

## Exercise 8: Recoding Factors

Use `newLevels` to combine Tuesday, Wednesday, and Thursday into a single level

*** =instructions

In the last exercise, you used `newLevels` to specify all of the levels in a
factor. Now, use `newLevels` to *combine* factors.

Imagine that you want to analyze the Dow Jones Industrial Average results by day
of week, but you think that Mondays and Fridays deserve special attention, while
Tuesday, Wednesday, and Thursday can all be lumped together. You can create a
new factor with just three levels to help you: Monday, Midweek, and Friday.

You'll use the `newLevels` argument again, but with a new syntax. Instead of a 
character vector, you'll pass a *named list* to `newLevels`. The *name* of each
element in the list should be the new level you want to create, and the *value*
of each element in the list should be the *old* level (or levels) that you want to
map to the *new* level.

For example, if I wanted to recode a month variable into a new quarter variable,
the syntax would look like this:

    quarter = list(varName = month,
                   newLevels = list(Q1 = c("Jan", "Feb", "Mar"),
                                    Q2 = c("Apr", "May", "Jun"),
                                    Q3 = c("Jul", "Aug" "Sep"),
                                    Q4 = c("Oct", "Nov", "Dec"))
    )

                                         
The `quarter` variable will have one level for each element in `newLevels` 
(`Q1`, `Q2`, `Q3`, `Q4`), each of which combines three months into one quarter.

 - **Do the same for the `DayOfWeek` variable**, mapping Monday to `Monday`,
Tuesday, Wednesday, and Thursday to a new level `Midweek`, and Friday to `Friday`.



*** =hint

 - Be sure that the new levels (on the left-hand side of the equation) are *not*
   quoted, while the old levels (on the right-hand side) are. For example:
   `Monday = "Monday"`
   
 - You'll need a total of three entries in the `newLevels` list


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

djia_source <- RxXdfData( file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf") )


djia_factor_year <- tempfile(fileext = ".xdf")

# Create factors from Year variable
rxFactors(inData = djia_source, outFile = djia_factor_year, 
          factorInfo = list(FactorYear = list(varName = "Year")), 
          overwrite = TRUE)
```


*** =sample_code

```{r,eval=FALSE}

# Note the levels of DayOfWeek: Monday, Tuesday, Wednesday, Thursday, Friday
rxGetVarInfo(djia_factor_year)

# Use the blank space to recode Monday to Monday,
# Tuesday, Wednesday, and Thursday to a new level called Midweek
# and Friday to Friday
rxFactors(inData = djia_factor_year, outFile = djia_factor_year, 
          factorInfo = list(
              DayRecoded = list(varName = "DayOfWeek",
                                newLevels = list(__________))),
          overwrite = TRUE)


# View new factor levels
rxGetVarInfo(djia_factor_year)

```

*** =solution

```{r,eval=FALSE}

# Note the levels of DayOfWeek: Monday, Tuesday, Wednesday, Thursday, Friday
rxGetVarInfo(djia_factor_year)

# Use rxFactors to create a recoded DayOfWeek factor called DayRecoded
rxFactors(inData = djia_factor_year, outFile = djia_factor_year, 
          factorInfo = list(
              DayRecoded = list(varName = "DayOfWeek",
                                newLevels = list(Monday = "Monday",
                                                 Midweek = c("Tuesday", "Wednesday", "Thursday"),
                                                 Friday = "Friday"))),
          overwrite = TRUE)


# View new factor levels
rxGetVarInfo(djia_factor_year)

```


*** =sct

```{r,eval=FALSE}



test_error()

test_student_typed("rxFactors",
                   not_typed_msg = "It doesn't appear that you made a call to rxFactors - start there!")

test_student_typed("djia_factor_year",
                   not_typed_msg = "Be sure to pass 'djia_factor_year' to the outFile argument to store your results.")

test_student_typed("factorInfo = ",
                   not_typed_msg = "You'll need to use rxFactors' factorInfo argument.")

test_student_typed("factorInfo = list(",
                   not_typed_msg = "The factorInfo argument takes a list - double-check your syntax.")



# Check the results
test_expression_result("FactorYear" %in% rxGetVarNames(djia_factor_year),
                       incorrect_msg = "It doesn't look like the variable FactorYear was created on the djia_factor_year dataset - try again.")

test_expression_output(rxGetVarInfo(djia_factor_year)[["FactorYear"]][["varType"]],
                       incorrect_msg = "FactorYear was created, but it doesn't seem to be a factor - check your code again.")

test_expression_output(rxGetVarInfo(djia_factor_year)[["FactorYear"]][["varType"]][["levels"]],
                       incorrect_msg = "Your factor levels don't seem to be in the right order. Try again.")




success_msg("Well done! Hopefully factors don't seem as intimidating now.")


```

*** =skills
1, 3


--- type:VideoExercise key:df07d46b0d

## Conclusion 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141147215
```

