---
title       : Importing Data
description : Introduction to the eXternal Data Frame (XDF) file format.
attachments :
  slides_link : https://s3.amazonaws.com/assets.datacamp.com/course/teach/slides_example.pdf
mode        : selfcontained



--- type:VideoExercise key:5b7c7478a3

## Introduction to the XDF File Format 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141145803
```






--- type:MultipleChoiceExercise contains_graph:FALSE key:c49f768bfa

## Exercise 1: R and Big Data

Why does open-source R have a hard time analyzing big data sets?

*** =instructions

- R analyzes data from the hard drive, and large datasets can be many terabytes in size.
- In R, data resides in memory by default, and large datasets are too big for the memory capacity of many computers.
- R relies on other scripting languages to be able to perform any sort of computation on big datasets. 
- R has so many system configuration settings that it becomes a complicated affair to adjust it to handle big datasets. 

*** =hint

Review the course video and choose the selection that most closely matches the addressed topic.

*** =pre_exercise_code

```{r,eval=FALSE}
```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 2, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Incorrect - R doesn't store data on the hard drive by default.",
                          "Correct!",
                          "Incorrect - R does sometimes make calls to other programming languages, but that usually *improves* R's performance.",
                          "Incorrect - R usually needs very little special configuration."
))

```






--- type:MultipleChoiceExercise contains_graph:FALSE key:cda00447a8

## Exercise 2: Advantages of PEMAs

What are some of the advantages of utilizing PEMAs for data processing and 
analysis? Pick the answer with all the true statements.

- a. PEMAs allow you to analyze data in chunks that are managable with R's memory limitations.
- b. PEMAs consolidate all data in a central location so that it's easy to find and access. 
- c. PEMAs store entire data sets in your computer's memory, allowing you to quickly access and analyze data.
- d. PEMAs can run in parallel, allowing you to distribute complex computations among many computing resources.


*** =instructions
- a, b, d
- b, c
- b, c, d
- a, d

*** =hint

Review the course video and choose the selection that most closely matches the addressed topic.

*** =pre_exercise_code

```{r,eval=FALSE}
```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 4, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Incorrect - PEMAs work well with distributed data.",
                          "Incorrect - Open-Source R stores all data in memory, but PEMAs work regardless of where the data is stored.",
                          "Incorrect - PEMAs work with distributed data, whether in memory or on hard drives.",
                          "Correct - PEMAs let users surpass R's memory limits and work in parallel."
))

```






--- type:MultipleChoiceExercise contains_graph:FALSE key:afe0cbc507

## Exercise 3: Advantages of Column Orientation

Why is it beneficial for XDF files to be column-oriented rather than row-oriented?


*** =instructions

- All the values of a single variable are stored together on-disk, which makes 
  them faster to query.
- All the values relating to one record/observation are stored together on-disk, 
  which makes them faster to query.
- Column orientation ensures that a variable is never split across chunks.
- The most widely-used databases (SQL Server, MySQL, etc.) are column-oriented


*** =hint

Review the course video and choose the selection that most closely matches the addressed topic.

*** =pre_exercise_code

```{r,eval=FALSE}
```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 1, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Correct!",
                          "Incorrect - this is the case for row-oriented data.",
                          "Incorrect - variables must be split across chunks in order to take advantage of distributed computation.",
                          "Incorrect - the most widely-used databases are all row-oriented (although column-oriented databases are becoming increasingly popular)"
))

```








--- type:NormalExercise xp:50 key:79c81a7847

## Exercise 4: Import Data with rxImport

Use rxImport to import data from a csv file to xdf format.

*** =instructions

`rxImport` is the key function for importing data into an XDF file. In this
exercise, use `rxImport` to convert a CSV file (`mortDefaultCsv`) into an XDF
file (`mortDefaultXdf`), and then check the results using `rxGetInfo`.


*** =hint

You need two arguments in `rxImport`: `inData` should be the pointer to the CSV
file, and `outFile` should be the point to the XDF file you want to create.

`rxGetInfo` just needs the path to that XDF file - although you might also want
to set `getVarInfo` to `TRUE` to see some information about the variables, or
`numRows` to 5 to see the first few rows of the new XDF file.



*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
file.remove("mortDefaultSmall2009.xdf")
```

*** =solution
```{r, eval=FALSE}
# Create a pointer to the CSV file
mortDefaultCsv <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall2009.csv")

# Create a pointer to the XDF file you need to create
mortDefaultXdf <- "mortDefaultSmall2009.xdf"

# Use rxImport to import the data into xdf format
rxImport(inData = mortDefaultCsv, outFile = mortDefaultXdf, overwrite = TRUE)

# Use rxGetInfo to check the new xdf file
rxGetInfo(mortDefaultXdf, getVarInfo=TRUE, numRows = 5)
```

*** =sample_code
```{r, eval=FALSE}
# Create a pointer to the CSV file
mortDefaultCsv <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall2009.csv")

# Create a pointer to the XDF file you need to create
mortDefaultXdf <- "mortDefaultSmall2009.xdf"


# Use rxImport to import the data into xdf format


# Use rxGetInfo to check the new xdf file



```

*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxImport",
                   not_typed_msg = "Be sure to call the 'rxImport' function!")

test_student_typed("inData = mortDefaultCsv",
                   not_typed_msg = "Make sure you use the 'mortDefaultCsv' object as your 'inData' argument. It's a pointer to the CSV file we're trying to import.")


test_student_typed("outFile = mortDefaultXdf",
                   not_typed_msg = "Make sure you use the 'mortDefaultXdf' object as your 'outFile' argument. It's just the path to the XDF file we're trying to create.")

test_student_typed('rxGetInfo',
                   not_typed_msg = "It's always a good idea to check that your file was imported correctly, so use the function 'rxGetInfo' on your new XDF file (mortDefaultXdf).")



test_expression_output(rxGetInfo("mortDefaultSmall2009.xdf"),
                       incorrect_msg = "Your 'inner_joined' data.frame doesn't seem to have the right number of rows - check your code again.")


success_msg("Off to a good start!")


```




















