---
title_meta  : Chapter 4
title       : Merging In-Database and Out-of-Database Tables
description : Using RRE to combine data from multiple sources
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Merging In-Database and Out-of-Database Tables

*** =video_link

```{r, eval = FALSE}
//player.vimeo.com/video/147564263
```








--- type:NormalExercise xp:50

## Exercise 9: Merge with Renamed Variables

Add information about each flight's `Origin` airport to the dataset.

*** =instructions

You've been provided with the `flights` query from the previous exercises and a
CSV file containing information about the airports. 

 - Examine the `airports` dataset to see which variable corresponds to the `Origin`
   variable on `flights`.
 - Use the `newVarNames2` argument in `rxMerge` to rename that variable to 
   `Origin`, so the variables can be matched.
 - Merge the two datasets into a new XDF file.



*** =hint

 - You need to rename the variable `iata` on `airports` to `Origin`, to match that
   variable on `flights`
 - The `newVarNames2` syntax is a named character vector, like this: `c(iata = "Origin")`


*** =pre_exercise_code
```{r, eval = FALSE}
options(stringsAsFactors = FALSE)
library(RevoScaleR)
rxOptions(reportProgress=0)


```

*** =solution
```{r, eval = FALSE}

# The connection string stays the same as in the previous exercises
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Query stays the same, too
flights <- RxOdbcData(sqlQuery = "
    SELECT flight.Year, flight.Month, flight.DayofMonth, 
           flight.FlightNum, flight.Origin,
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code",
    connectionString = conn_string
)


# Pull the flights data into an XDF; rxMerge can't work directly with RxOdbcData objects
flights_xdf <- tempfile(fileext = ".xdf")

rxImport(inData = flights, outFile = flights_xdf)


# Read in the CSV file with airport details
airports <- read.csv(file.path("data", "airports.csv"))


# Set up a new XDF for the merged data
flights_airports <- tempfile(fileext = ".xdf")


# Merge the two datasets and output to flights_airports
rxMerge(inData1 = flights_xdf,
        inData2 = airports,
        outFile = flights_airports,
        matchVars = "Origin",
        newVarNames2 = c(iata = "Origin")
)


# Check the results
rxDataStep(flights_airports, numRows = 10)


```

*** =sample_code
```{r, eval = FALSE}

# The connection string stays the same as in the previous exercises
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Query stays the same, too
flights <- RxOdbcData(sqlQuery = "
    SELECT flight.Year, flight.Month, flight.DayofMonth, 
           flight.FlightNum, flight.Origin,
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code",
    connectionString = conn_string
)


# Use rxImport to pull the flights data into an XDF; rxMerge can't work 
# directly with RxOdbcData objects
flights_xdf <- tempfile(fileext = ".xdf")




# Read in the CSV file with airport details
airports <- read.csv(file.path("data", "airports.csv"))


# Set up a new XDF for the merged data
flights_airports <- tempfile(fileext = ".xdf")


# Use rxMerge to join the two datasets and output to flights_airports



# Check the results
rxDataStep(flights_airports, numRows = 10)


```

*** =sct
```{r, eval = FALSE}

test_error()


test_function("RxOdbcData", 
              args = c("sqlQuery", "connectionString", "colClasses"),
              incorrect_msg = "Be sure to use both the 'connectionString', 'sqlQuery', and 'colClasses' arguments to set up your data source.",
              not_called_msg = "It doesn't see like you called RxOdbcData - try again.")

test_function("rxImport",
              args = c("inData", "outFile"),
              incorrect_msg = "Be sure to use the rxImport function to pull the query results into an XDF file.",
              not_called_msg = "It doesn't seem like you called rxImport - you'll need to do that to create the XDF file.")

test_function("rxMerge",
              args = c("inData1", "inData2", "outFile", "matchVars"),
              incorrect_msg = "Be sure to use the rxMerge function to merge tables together. You'll need to specify the two input datasets, where to store the output dataset, and the variables to match on.",
              not_called_msg = "It doesn't seem like you called rxMerge - you'll need to do that to create the XDF file.")

test_student_typed('newVarNames2 = c(iata = "Origin")',
                   not_typed_msg = "Be sure to use the newVarNames2 argument in rxMerge to rename iata to 'Origin'; otherwise, you won't be able to merge the two tables together.")


test_expression_result(rxDataStep(flights_airports),
                       incorrect_msg = "Your results don't look quite right - try double-checking your query.")


```





--- type:VideoExercise

## Conclusion

*** =video_link

```{r, eval = FALSE}
//player.vimeo.com/video/147564264
```




