---
title_meta  : Chapter 1
title       : Introduction to Summary Statistics with Microsoft R Server
description : Functions for describing and aggregating data
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise key:df70737089

## Introduction to Summary Statistics

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141148546
```


--- type:VideoExercise key:14e2e5d0d6

## Summarizing Numeric Variables 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141148550
```

--- type:NormalExercise xp:50 key:c556e044a9

## Exercise 1: Summarize a Numeric Variable
  

*** =instructions

Use `rxSummary` to calculate descriptive statistics for one of the numeric
variables on the dataset `diamondsXdf`.

 - First, use `rxGetVarInfo` to identify a numeric variable,
 - then use `rxSummary` to summarize it.


*** =hint
The basic syntax you need is: `rxSummary( ~ some_variable, data = diamondsXdf)`.


*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
# Examine the variables on the diamonds dataset
rxGetVarInfo(diamondsXdf)
```

*** =solution

```{r,eval=FALSE}
# Examine the variables on the diamonds dataset
rxGetVarInfo(diamondsXdf)

rxSummary( ~ depth, data = diamondsXdf)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxSummary",
                   not_typed_msg = "It doesn't appear that you made a call to rxSummary - be sure to call that function.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

# Identifythe numeric variables
diamonds_numerics <- c("carat", "depth", "table", "price", "x", "y", "z")

test_student_typed(paste("~", diamonds_numerics),
                   not_typed_msg = "Your formula doesn't seem quite right - be sure to put one of the numeric variables on the right-hand side of the tilde (~).")


# Catch-all
test_output_contains(expr = "Summary Statistics Results for: ~",
                     incorrect_msg = "Your output doesn't seem quite right - try again.")



success_msg("Good work! That's all it takes to summarize a variable.")


```

*** =skills
1, 3


--- type:NormalExercise xp:50 key:4fbcab093f

## Exercise 2: Summarize a Factor Variable
  

*** =instructions

You can also use `rxSummary` to summarize factor variables. The syntax is the
same, but the output is a frequency table instead of the quantitative descriptive
statistics. In this exercise:

 - Pick one of the factors on `diamondsXdf` and
summarize it.



*** =hint
The basic syntax you need is: `rxSummary( ~ some_variable, data = diamondsXdf)`.


*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
# Use `rxGetVarInfo` to pick a factor
rxGetVarInfo(diamondsXdf)
```

*** =solution

```{r,eval=FALSE}
# Use `rxGetVarInfo` to pick a factor
rxGetVarInfo(diamondsXdf)

rxSummary( ~ cut, data = diamondsXdf)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxSummary",
                   not_typed_msg = "It doesn't appear that you made a call to rxSummary - be sure to call that function.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

# Identify the factor variables
diamonds_factors <- c("cut", "color", "clarity")

test_student_typed(paste("~", diamonds_factors),
                   not_typed_msg = "Your formula doesn't seem quite right - be sure to put one of the factor variables on the right-hand side of the tilde (~).")


# Catch-all
test_output_contains(expr = "Category Counts for",
                     incorrect_msg = "Your output doesn't seem quite right - try again.")


success_msg("Nice work. Take a moment to see how the output changes for a factor.")


```

*** =skills
1, 3

--- type:NormalExercise xp:50 key:c7b5f11645

## Exercise 3: Summarize Multiple Variables
  

*** =instructions

If you want to describe one variable, chances are you want to describe several.

 - Use `rxSummary` to summarize `carat`, `cut`, and `price` on the `diamondsXdf`
datset.


*** =hint
The basic syntax you need is: `rxSummary( ~ some_variable + another + some_other, data = diamondsXdf)`


*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
# Use rxGetVarInfo to identify the factors
rxGetVarInfo(diamondsXdf)
```

*** =solution

```{r,eval=FALSE}
# Use rxGetVarInfo to identify the factors
rxGetVarInfo(diamondsXdf)

rxSummary( ~ carat + cut + price, data = diamondsXdf)

```


*** =sct

```{r,eval=FALSE}


test_error()

test_student_typed("rxSummary",
                   not_typed_msg = "It doesn't appear that you made a call to rxSummary - be sure to call that function.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

test_student_typed("carat",
                   not_typed_msg = "You missed the variable 'carat' - try again.")

test_student_typed("cut",
                   not_typed_msg = "You missed the variable 'cut' - try again.")

test_student_typed("price",
                   not_typed_msg = "You missed the variable 'price' - try again.")

test_output_contains("Category Counts for cut",
                     incorrect_msg = "It doesn't look like you summarized the factor variable 'cut' correctly - try again.")

test_output_contains("carat.*084551",
                     incorrect_msg = "It doesn't look like you summarized the variable 'carat' correctly - try again.")

test_output_contains("price.*3406",
                     incorrect_msg = "It doesn't look like you summarized the variable 'price' correctly - try again.")

success_msg("Excellent - you've learned the basics of rxSummary.")


```

*** =skills
1, 3



--- type:VideoExercise key:ba7fad7458

## Groupwise Summaries

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141148547
```

--- type:NormalExercise xp:50 key:9485f95e2f

## Exercise 4: Groupwise Summaries



*** =instructions

`rxSummary` makes comparing the distribution of variables between groups easy by
taking advantage of R's formula syntax. In the last few exercises, you summarized
variables by naming them on the right-hand side of the formula (` ~ carat`); to
perform the same summaries for the records in each level of a factor, we move
the outcome/dependent variable to the left-hand side of the function, and put
the grouping variable on the right-hand side: `carat ~ cut`

 - Use `rxSummary` to summarize one of the numeric variables on `diamondsXdf` 
for each level of one of the factors.


*** =hint
Make sure to put one of the numeric variables (`carat`, `depth`, `table`, `price`,
`x`, `y`, `z`) on the left-hand side of the tilde (~), and one of the factors
(`cut`, `color`, `clarity`) on the right-hand side.

*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
# Use rxGetVarInfo to pick one numeric variable and one factor
rxGetVarInfo(diamondsXdf)
```

*** =solution

```{r,eval=FALSE}
# Use rxGetVarInfo to pick one numeric variable and one factor
rxGetVarInfo(diamondsXdf)

rxSummary(depth ~ clarity, data = diamondsXdf)
```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxSummary",
                   not_typed_msg = "It doesn't appear that you made a call to rxSummary - be sure to call that function.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

# Identify the numeric and factor variable names
diamonds_numerics <- c("carat", "depth", "table", "price", "x", "y", "z")
diamonds_factors <- c("cut", "color", "clarity")

test_student_typed(
    paste0("rxSummary(", diamonds_numerics, " ~"),
    not_typed_msg = "Be sure to put one of the numeric variables on the left-hand side of the formula.")


# Test for the validity of the right-hand side (and, consequently, the whole formula)
diamonds_combo <- expand.grid(num = diamonds_numerics, fac = diamonds_factors)

permissible_formulas <- apply(diamonds_combo, 
                              MARGIN = 1, 
                              FUN = function(x) {
                                  paste("rxSummary(", x["num"], "~", x["fac"])
                              }  
)

test_student_typed(permissible_formulas,
                   not_typed_msg = "Be sure to put one of the factor vectors (cut, color, clarity) on the right-hand side of the formula.")


test_output_contains(expr = "Summary Statistics Results for:",
                     incorrect_msg = "Your output doesn't look quite right - try again.")
                     


success_msg("Excellent - you've learned the basics of rxSummary.")

```

*** =skills
1, 3


--- type:NormalExercise xp:50 key:1f73cf9683

## Exercise 5: Multiple Grouping Variables
  
Calculate summary statistics using two grouping variables


*** =instructions

 - Now, try summarizing a numeric variable using *two* factors. 

Your numeric
variable should still be on the left-hand side of the equation, but now you'll
have two factors on the right-hand side, joined by a `:`. It should look something
like this:

    someVar ~ factor1 : factor2
    


*** =hint
The basic syntax you need is `rxSummary(numericVar ~ factor1 : factor2, data = diamondsXdf`

*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
# Use rxGetVarInfo to pick one numeric variable and two factors
rxGetVarInfo(diamondsXdf)
```

*** =solution

```{r,eval=FALSE}
# Use rxGetVarInfo to pick one numeric variable and two factors
rxGetVarInfo(diamondsXdf)

rxSummary(depth ~ cut : color, data = diamondsXdf)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxSummary",
                   not_typed_msg = "It doesn't appear that you made a call to rxSummary - be sure to call that function.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")


# Identify the numeric and factor variable names
diamonds_numerics <- c("carat", "depth", "table", "price", "x", "y", "z")
diamonds_factors <- c("cut", "color", "clarity")

test_student_typed(
    paste0("rxSummary(", diamonds_numerics, " ~"),
    not_typed_msg = "Be sure to put one of the numeric variables on the left-hand side of the formula.")


# Test for the validity of the right-hand side (and, consequently, the whole formula)
diamonds_combo <- expand.grid(num = diamonds_numerics, 
                              fac1 = diamonds_factors, 
                              fac2 = diamonds_factors)

permissible_formulas <- apply(diamonds_combo, 
                              MARGIN = 1, 
                              FUN = function(x) {
                                  paste("rxSummary(", x["num"], "~", x["fac1"], ":", x["fac2"])
                              }  
)

test_student_typed(permissible_formulas,
                   not_typed_msg = "Be sure to put two of the factor vectors (cut, color, clarity) on the right-hand side of the formula, separated by a colon.")


test_output_contains(expr = "Summary Statistics Results for:",
                     incorrect_msg = "Your output doesn't look quite right - try again.")

success_msg("Good job - hopefully the formula syntax is becoming more familiar.")
```

*** =skills
1, 3



--- type:VideoExercise key:05fd8707fc

## Calculating Quantiles

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141148551
```


--- type:NormalExercise xp:50 key:5bd1aeb7dc

## Exercise 6: Quantiles

Calculate the quantiles of the `price` variable


*** =instructions

For variables with a long-tailed distribution like the prices of the diamonds in
`diamondsXdf`, quantiles can be more descriptive than the mean and standard
deviation.

 - Use `rxQuantile` to summarize the `price` variable from `diamondsdXdf.
 
The basic syntax you need is `rxQuantile(varName = "variable", data = dataset)`.

*** =hint
Be sure that you've put quotes around `price` when you pass it to `varName`.

*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
# This histogram shows how long-tailed and non-normal the price variable is
rxHistogram( ~ price, data = diamondsXdf)
```

*** =solution

```{r,eval=FALSE}
# This histogram shows how long-tailed and non-normal the price variable is
rxHistogram( ~ price, data = diamondsXdf)

rxQuantile(varName = "price", data = diamondsXdf)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxQuantile",
                   not_typed_msg = "It doesn't appear that you made a call to rxQuantile - be sure to call that function.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

test_student_typed("rxQuantile('price'",
                   not_typed_msg = "Make sure that you put quotes around the name of the price variable: \"price\"")

test_output_contains(expr = "0%   25%   50%   75%  100%",
                     incorrect_msg = "Your output doesn't look quite right - be sure to just let rxQuantile calculate the default percentiles.")

success_msg("Good work!")
```

*** =skills
1, 3



--- type:NormalExercise xp:50 key:0949948d3f

## Exercise 7: Non-Default Quantiles
  
Calculate non-standard quantiles for the `price` variable.


*** =instructions

The default quantiles are useful, but sometimes you may be interested in others.

 - Use the `probs` argument to calculate the 0.01 and 0.99 quantiles.


*** =hint
Pass the two desired quantiles to the `probs` argument as a vector: `c(0.01, 0.99)`

*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
# This histogram shows how long-tailed and non-normal the price variable is
rxHistogram( ~ price, data = diamondsXdf)
```

*** =solution

```{r,eval=FALSE}
# This histogram shows how long-tailed and non-normal the price variable is
rxHistogram( ~ price, data = diamondsXdf)

rxQuantile(varName = "price", data = diamondsXdf, probs = c(0.01, 0.99))

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxQuantile",
                   not_typed_msg = "It doesn't appear that you made a call to rxQuantile - be sure to call that function.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

test_student_typed("rxQuantile('price'",
                   not_typed_msg = "Make sure that you put quotes around the name of the price variable: \"price\"")


test_student_typed("rxQuantile.*probs = .*",
                   not_typed_msg = "Be sure to call the probs argument to set the desired quantiles.")


test_output_contains(expr = "1%   99%",
                     incorrect_msg = "Your output doesn't look quite right - be sure you've set the probs argument to c(0.01, 0.99)."

success_msg("Excellent! You've learned the essential parts of rxQuantile's functionality.")
```

*** =skills
1, 3

