---
title_meta  : Chapter 1
title       : Introduction to Merging
description : Introduction to join terminology and normalized tables
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Introduction to Merging

*** =video_link

```{r, eval = FALSE}
//player.vimeo.com/video/147564262
```




--- type:MultipleChoiceExercise contains_graph:FALSE

## Exercise 1: Keys, IDs, and Matching Variables

The variable (or variables) we use to match records from one table to another
are called keys, IDs, or matching variables.

 - Examine the two tables loaded into your workspace by typing their names into
   the console: `authors` and `organizations`

 - Which variables would you use to join these two tables?


*** =instructions


 - `affiliation` on both tables
 - `name` on both tables
 - `affiliation` on `authors`, and `name` on organizations`
 - This is a natural join - no key necessary
  
*** =hint

The matching variables don't necessarily have the same name in both tables.

*** =pre_exercise_code

```{r,eval=FALSE}

options(stringsAsFactors = FALSE)

authors <- data.frame(name = c("Ross Ihaka", "Robert Gentleman", 
                               "Peter Dalgaard", "Kurt Hornik"),
                     affiliation = c("University of Auckland", "23andMe", 
                                     "University of Copenhagen", NA))

organizations <- data.frame(name = c("University of Auckland", "23andMe", 
                                     "University of Copenhagen"),
                            location = c("New Zealand", "USA", "Denmark"))


```


*** =sample_code
```{r, eval = FALSE}

# Examine these two tables
authors

organizations


```



*** =sct

```{r,eval=FALSE}

test_mc(correct = 3, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Incorrect - affiliation isn't a variable on organizations.",
                          "Not quite - both tables have a variable called 'name', but it doesn't mean the same thing on both tables.",
                          "Correct!",
                          "Incorrect - the keys needed for this join have different names, so a natural join won't work."
))

```








--- type:MultipleChoiceExercise contains_graph:FALSE

## Exercise 2: Different Types of Joins

You may have noticed that one of the authors has a missing affiliation (it should 
be "Wirtschaftsuniversitat Wien"). This has significant implications for merging
the two tables: some types of joins will drop that author, while others will create
a combined table with missing values.

 - Examine the two tables loaded into your workspace by typing their names into
   the console: `authors` and `organizations`

 - To create a table that *keeps* the author with the missing value, which type of
join would you use?


*** =instructions

 - An inner join
 - A left outer join
 - A union join
 - A retaining join
  
*** =hint

You can examine the tables by typing their names into the console: `authors`;
`organizations`

*** =pre_exercise_code

```{r,eval=FALSE}

options(stringsAsFactors = FALSE)

authors <- data.frame(name = c("Ross Ihaka", "Robert Gentleman", 
                               "Peter Dalgaard", "Kurt Hornik"),
                     affiliation = c("University of Auckland", "23andMe", 
                                     "University of Copenhagen", NA))

organizations <- data.frame(name = c("University of Auckland", "23andMe", 
                                     "University of Copenhagen"),
                            location = c("New Zealand", "USA", "Denmark"))

```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 2, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Incorrect - inner joins will drop any table that doesn't have a matching key in both tables.",
                          "Correct! As long as authors is the 'left' table.",
                          "Incorrect - unions stack tables on top of one another, rather than linking records between two tables.",
                          "Incorrect - I just made retaining joins up!"
))

```







