---
title_meta  : Chapter 3
title       : Working with Databases
description : Connect to databases and make in-database joins
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise key:58bb8d0c98

## Working with Databases

*** =video_link

```{r, eval = FALSE}
//player.vimeo.com/video/147564265
```





--- type:NormalExercise xp:50 key:423b09a4c5

## Exercise 6: Set Up a Database Connection

Use `RxOdbcData` to connect to the SQLite database

*** =instructions

In this exercise, use the `RxOdbcData` function to connect to the `flights` table
in the given `sqlite` database.

`RxOdbcData` takes two parameters: a connection string (provided for you) and an
SQL query to run on the databas. In this case, the query we want is:

    SELECT * FROM flight

This is a crucial point: `RxOdbcData` provides access to a **specific SQL query**
in a database. Depending on the analysis, you might need several `RxOdbcData`
connections - one for each table, query, or join you use.



*** =hint

 - Be sure to put quotes around your query: `"SELECT * FROM flight"`
 - Pass the `conn_string` object to the `connectionString` argument (not the other way around)


*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
```

*** =solution
```{r, eval = FALSE}

# First, create the database connection string
# The first part of the string specifies the database driver;
# the second part points to the SQLite file.
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Set up the connection using RxOdbcData
flights <- RxOdbcData(sqlQuery = "SELECT * FROM flight", 
                     connectionString = conn_string)


# Pull a few rows to check the results
head(flights)

```

*** =sample_code
```{r, eval = FALSE}

# First, create the database connection string
# The first part of the string specifies the database driver;
# the second part points to the SQLite file.
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Set up the connection using RxOdbcData
flights <- RxOdbcData(


# Pull a few rows to check the results
head(flights)



```

*** =sct
```{r, eval = FALSE}

test_error()

test_function("RxOdbcData", 
              args = c("sqlQuery", "connectionString"),
              incorrect_msg = "Be sure to use both the 'sqlQuery' and 'connectionString' arguments to set up your data source.",
              not_called_msg = "It doesn't see like you called RxOdbcData - try again.")

test_expression_output(head(flights),
                       incorrect_msg = "Your results don't look quite right - try double-checking your query.")


test_object("flights",
            undefined_msg = "It seems like the 'flights' object hasn't been created; check your code again.",
            incorrect_msg = "Your 'flights' doesn't look quite right - check it again.")

```







--- type:NormalExercise xp:50 key:f5894d53dc

## Exercise 7: Join Two Tables In-Database

Use `RxOdbcData` to join two tables inside the database

*** =instructions

Now, use `RxOdbcData` to create a query that joins two tables in the database.
This left outer join will just add the full carrier name (`carrier.Description`)
to the flight data:

    SELECT flight.Year, flight.Month, flight.DayofMonth, flight.FlightNum, 
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code

Set up an `RxOdbcData` connection using this query, and check the results.


*** =hint

 - You can simply copy the above query into your answer - but be sure to put it in quotes.



*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
```

*** =solution
```{r, eval = FALSE}

# The connection string stays the same as in the previous exercise
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Use the new query
flights <- RxOdbcData(sqlQuery = "
    SELECT flight.Year, flight.Month, flight.DayofMonth, flight.FlightNum, 
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code",

    connectionString = conn_string)


# Pull a few rows to check the results
head(flights)

```

*** =sample_code
```{r, eval = FALSE}

# The connection string stays the same as in the previous exercise
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Use the new query
flights <- RxOdbcData(


# Pull a few rows to check the results
head(flights)



```

*** =sct
```{r, eval = FALSE}

test_error()

test_function("RxOdbcData", 
              args = c("sqlQuery", "connectionString"),
              incorrect_msg = "Be sure to use both the 'sqlQuery' and 'connectionString' arguments to set up your data source.",
              not_called_msg = "It doesn't see like you called RxOdbcData - try again.")

test_expression_output(head(flights),
                       incorrect_msg = "Your results don't look quite right - try double-checking your query.")


test_object("flights",
            undefined_msg = "It seems like the 'flights' object hasn't been created; check your code again.",
            incorrect_msg = "Your 'flights' doesn't look quite right - check it again.")

```






--- type:NormalExercise xp:50 key:019e300c3d

## Exercise 8: Import from Database to XDF

Use `rxImport` to pull your join into an XDF file

*** =instructions

Use `rxImport` to pull the previous exercise's query:

    SELECT flight.Year, flight.Month, flight.DayofMonth, flight.FlightNum, 
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code

... into a new XDF file.


*** =hint

 - Pass the `RxOdbcData` object you create - `flights` - to the `inData` argument in `rxImport`
 - Pass `flights_xdf` to the `outFile` argument in `rxImport`


*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
```

*** =solution
```{r, eval = FALSE}

# Use the previous connection string
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Previous query
flights <- RxOdbcData(sqlQuery = "
    SELECT flight.Year, flight.Month, flight.DayofMonth, flight.FlightNum, 
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code",

    connectionString = conn_string)


# Create a path to a new XDF file
flights_xdf <- tempfile(fileext = ".xdf")


# Use rxImport to pull the flights query results into the flights_xdf file
rxImport(inData = flights,
         outFile = flights_xdf)



# Pull a few rows to check the results
rxDataStep(flights_xdf, numRows = 10)

```

*** =sample_code
```{r, eval = FALSE}

# Use the previous connection string
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Previous query
flights <- RxOdbcData(sqlQuery = "
    SELECT flight.Year, flight.Month, flight.DayofMonth, flight.FlightNum, 
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code",
    connectionString = conn_string)


# Create a path to a new XDF file
flights_xdf <- tempfile(fileext = ".xdf")


# Use rxImport to pull the flights query results into the flights_xdf file




# Pull a few rows to check the results
head(flights_xdf)


```

*** =sct
```{r, eval = FALSE}

test_error()


test_function("RxOdbcData", 
              args = c("sqlQuery", "connectionString"),
              incorrect_msg = "Be sure to use both the 'sqlQuery' and 'connectionString' arguments to set up your data source.",
              not_called_msg = "It doesn't see like you called RxOdbcData - try again.")

test_function("rxImport",
              args = c("inData", "outFile"),
              incorrect_msg = "Be sure to use the rxImport function to pull the query results into an XDF file.",
              not_called_msg = "It doesn't seem like you called rxImport - you'll need to do that to create the XDF file.")

test_expression_result(rxDataStep(flights_xdf),
                       incorrect_msg = "Your results don't look quite right - try double-checking your query.")


```










--- type:NormalExercise xp:50 key:2bfaff6ad4

## Exercise 9: Set Factors on Import

Use `rxImport` to convert the carrier details to a factor

*** =instructions

You might have noticed that the XDF created in the last exercise stored 
UniqueCarrier and Description as character variables. But there are many situations
in which they'd be more useful as factors.

In this exercise, use the `RxOdbcData` argument `colClasses` to convert 
UniqueCarrier and Description into factors upon import.

Use a named vector like this one:

    c(first_variable = "character", second_variable = "integer")

... to specify each variable's type in `colClasses`.


*** =hint

Be sure to specify the type of each variable in `RxOdbcData`.


*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)


```

*** =solution
```{r, eval = FALSE}

# Use the previous connection string
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Here's the previous query. You'll need to fill out the colClasses argument
# to convert the variables to factors.
flights <- RxOdbcData(sqlQuery = "
    SELECT flight.Year, flight.Month, flight.DayofMonth, flight.FlightNum, 
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code",
    
    connectionString = conn_string,
    
    colClasses = c(Year = "integer",  Month = "integer",  DayofMonth = "integer", FlightNum = "integer",
                   UniqueCarrier = "factor", Description = "factor")
)


# Create a path to a new XDF file
flights_xdf <- tempfile(fileext = ".xdf")



# Use rxImport to pull the flights query results into the flights_xdf file
rxImport(inData = flights,
         outFile = flights_xdf
)


# Check the data types
rxGetVarInfo(flights_xdf)

```

*** =sample_code
```{r, eval = FALSE}

# Use the previous connection string
conn_string <- paste0("Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3")


# Here's the previous query. You'll need to fill out the colClasses argument
# to convert the variables to factors.
flights <- RxOdbcData(sqlQuery = "
    SELECT flight.Year, flight.Month, flight.DayofMonth, flight.FlightNum, 
           flight.UniqueCarrier, carrier.Description
    FROM flight
    LEFT OUTER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code",
    connectionString = conn_string,
    colClasses = ___________
)



# Create a path to a new XDF file
flights_xdf <- tempfile(fileext = ".xdf")

# Use rxImport to pull the flights query results into the flights_xdf file
rxImport(inData = flights, 
         outFile = flights_xdf)


# Check the data types
rxGetVarInfo(flights_xdf)

```

*** =sct
```{r, eval = FALSE}

test_error()


test_function("RxOdbcData", 
              args = c("sqlQuery", "connectionString", "colClasses"),
              incorrect_msg = "Be sure to use both the 'connectionString', 'sqlQuery', and 'colClasses' arguments to set up your data source.",
              not_called_msg = "It doesn't see like you called RxOdbcData - try again.")

test_function("rxImport",
              args = c("inData", "outFile"),
              incorrect_msg = "Be sure to use the rxImport function to pull the query results into an XDF file.",
              not_called_msg = "It doesn't seem like you called rxImport - you'll need to do that to create the XDF file.")


test_expression_output(rxGetVarInfo(flights_xdf)$UniqueCarrier,
                       incorrect_msg = 'Be sure that you specify UniqueCarrier = "factor" in the colClasses argument.')

test_expression_output(rxGetVarInfo(flights_xdf)$Description,
                       incorrect_msg = 'Be sure that you specify Description = "factor" in the colClasses argument.')

test_expression_result(rxDataStep(flights_xdf),
                       incorrect_msg = "Your results don't look quite right - try double-checking your query.")


```









