---
title_meta  : Chapter 1
title       : Sorting and Deduplicating Data with Revolution R Enterprise
description : Sort big datasets and remove duplicate records
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Introduction to Sorting and Deduplication

*** =video_link

```{r, eval = FALSE}
//player.vimeo.com/video/141511132
```




--- type:VideoExercise

## The rxSort() Function

*** =video_link

```{r, eval = FALSE}
//player.vimeo.com/video/141511131
```




--- type:NormalExercise xp:50

## Exercise 1: Sorting Data with rxSort

Use `rxSort` to sort an XDF file

*** =instructions

You've been provided with the path to an unsorted XDF file (`xdf_source`).

 - Use `rxSort` to sort that dataset by the variable `creditScore`, and write the
   results to a new XDF file called `xdf_sorted`.

 - Use `rxDataStep` with `numRows = 10` to check the first ten rows.

*** =hint

When you pass a variable name to `sortByVars`, be sure to quote it: `"var_name"`


*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
```

*** =solution
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Use rxDataStep to look at the first few rows of the original file
rxDataStep(xdf_source, numRows = 10)

# Create an output file for the sorted data
xdf_sorted <- tempfile(fileext = ".xdf")

# Use rxSort to sort the mortgage data by the variable creditScore,
# and write the results to xdf_sorted
rxSort(inData = xdf_source,
       outFile = xdf_sorted,
       sortByVars = "creditScore"
)

# Use rxDataStep once again to view the results
rxDataStep(xdf_sorted, numRows = 10)

```

*** =sample_code
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Use rxDataStep to look at the first few rows of the original file
rxDataStep(xdf_source, numRows = 10)

# Create a path for the sorted data
xdf_sorted <- tempfile(fileext = ".xdf")


# Use rxSort to sort the mortgage data by the variable creditScore,
# and write the results to xdf_sorted



# Use rxDataStep once again to view the results


```

*** =sct
```{r, eval = FALSE}

test_error()

test_student_typed("rxSort",
                   not_typed_msg = "It doesn't appear that you made a call to rxSort - start there!")

test_student_typed("inData = xdf_source",
                   not_typed_msg = "Make sure you're specifying 'xdf_source' as the inData argument for rxSort.")

test_student_typed("outFile = xdf_sorted",
                   not_typed_msg = "Make sure you're specifying 'xdf_sorted' as the outFile argument for rxSort.")

test_student_typed("sortByVars = 'creditScore'",
                   not_typed_msg = "It doesn't appear that you made a call to rxSort - start there!")

test_expression_result(rxDataStep(xdf_sorted),
                       incorrect_msg = "Your xdf_sorted file doesn't seem quite right - make sure that you're sorting by the variable 'creditScore'.")


success_msg("Well done! It gets a little more complicated than this, but not much.")


```




--- type:NormalExercise xp:50

## Exercise 2: Decreasing Sorts

Use `rxSort` to sort a dataset in **decreasing** order

*** =instructions

 - This time, use `rxSort` to sort `xdf_source` by **decreasing** `creditScore`.

 - Use `rxDataStep` with `numRows = 10` to check the first ten rows.


*** =hint

Set the `decreasing` argument to `TRUE`.


*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
```

*** =solution
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Use rxDataStep to look at the first few rows of the original file
rxDataStep(xdf_source, numRows = 10)

# Create an output file for the sorted data
xdf_sorted <- tempfile(fileext = ".xdf")

# Use rxSort to sort the mortgage data by the variable creditScore,
# and write the results to xdf_sorted
rxSort(inData = xdf_source,
       outFile = xdf_sorted,
       sortByVars = "creditScore",
       decreasing = TRUE
)

# Use rxDataStep with numRows = 10 once again to view the results
rxDataStep(xdf_sorted, numRows = 10)

```

*** =sample_code
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Create an output file for the sorted data
xdf_sorted <- "mortDefaultSmall_sorted.xdf"

# Use rxSort to sort the mortgage data by decreasing creditScore,
# and write the results to xdf_sorted


# Use rxDataStep with numRows = 10 once again to view the results



```

*** =sct
```{r, eval = FALSE}

test_error()

test_student_typed("rxSort",
                   not_typed_msg = "It doesn't appear that you made a call to rxSort - start there!")

test_student_typed("inData = xdf_source",
                   not_typed_msg = "Make sure you're specifying 'xdf_source' as the inData argument for rxSort.")

test_student_typed("outFile = xdf_sorted",
                   not_typed_msg = "Make sure you're specifying 'xdf_sorted' as the outFile argument for rxSort.")

test_student_typed("sortByVars = 'creditScore'",
                   not_typed_msg = "Be sure to pass the variable creditScore to the argument sortByVars, and remember that it needs to be quoted: 'creditScore'")

test_student_typed("decreasing = TRUE",
                   not_typed_msg = "Be sure to set the argument decreasing to TRUE.")

test_expression_result(rxDataStep(xdf_sorted),
                       incorrect_msg = "Your xdf_sorted file doesn't seem quite right - make sure that you've set the argument decreasing to TRUE.")



success_msg("Nicely done.")

```






--- type:NormalExercise xp:50

## Exercise 3: Sorting by Multiple Variables

Sort the same dataset using two variables: **decreasing** `creditScore` and 
**increasing** `yearsEmploy`

*** =instructions

Sort `xdf_source`

 - by **decreasing** `creditScore`, **and**
 - by **increasing** `yearsEmploy`.
 - Write the results to `xdf_sorted`.
 - View the results using `rxDataStep` with `numRows = 10`.


*** =hint

 - Sort by two variables by passing a vector of both variable names to the
   `sortByVars` argument: `c(variable1, variable2)`
   
 - Sort multliple variables in different directions by passing a vector to the
   `decreasing` argument. That vector should have one
   logical value (`TRUE` or `FALSE`) for each variable passed to `sortByVars`:
   `c(TRUE, FALSE)`


*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
```

*** =solution
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Create an output file for the sorted data
xdf_sorted <- tempfile(fileext = ".xdf")

# Sort in ascending order on one variable and descending order on another variable
rxSort(inData = xdf_source,
       outFile = xdf_sorted,
       sortByVars = c("creditScore", "yearsEmploy"),
       decreasing = c(TRUE, FALSE),
       overwrite = TRUE
)

# Use rxDataStep with numRows = 10 once again to view the results
rxDataStep(xdf_sorted, numRows = 10)
```

*** =sample_code
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Create an output file for the sorted data
xdf_sorted <- "mortDefaultSmall_sorted.xdf"

# Use rxSort to sort xdf_source by *ascending* creditScore and *descending* yearsEmploy;
# write the results to xdf_sorted.


# Use rxDataStep with numRows = 10 once again to view the results


```

*** =sct
```{r, eval = FALSE}

test_error()

test_student_typed("rxSort",
                   not_typed_msg = "It doesn't appear that you made a call to rxSort - start there!")

test_student_typed("inData = xdf_source",
                   not_typed_msg = "Make sure you're specifying 'xdf_source' as the inData argument for rxSort.")

test_student_typed("outFile = xdf_sorted",
                   not_typed_msg = "Make sure you're specifying 'xdf_sorted' as the outFile argument for rxSort.")

test_student_typed('sortByVars = c("creditScore", "yearsEmploy")',
                   not_typed_msg = 'Be sure to pass the variables creditScore and yearsEmploy to the argument sortByVars as a character vector: c("creditScore", "yearsEmploy")')

test_student_typed('decreasing = c(TRUE, FALSE)',
                   not_typed_msg = "The 'decreasing' argument needs two values: one TRUE for creditScore, and one FALSE for yearsEmploy. Pass them as a logical vector: c(TRUE, FALSE)")

test_expression_result(rxDataStep(xdf_sorted),
                       incorrect_msg = "Your xdf_sorted file doesn't seem quite right - make sure that you've set the argument decreasing to TRUE.")



success_msg("Well done! With the sortByVars and decreasing arguments, your sort can be as complex as you need it to be.")


```






--- type:NormalExercise xp:50

## Exercise 4: Deduplicating a Data Set

Use `rxSort` to remove duplicate records from a dataset.


*** =instructions

 - Use `rxSort` and the `removeDupKeys` argument to get all of the unique
   combinations of the `default` and `year` variables from `xdf_source`. Write
   the combinations to the `xdf_nodup` file.

 - Use the `dupFreqVar` argument to create a variable that indicates how many
   records existed for each combination of `default` and `year`.
   
 - Use `rxDataStep` to print **all** of the unique combinations to the console
   (don't set `numRows` this time).



*** =hint

 - The `removeDupKeys` argument takes TRUE or FALSE.
 
 - The `dupFreqVar` argument takes a single *character* value - the name of the
   column you want to create.

*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
```

*** =solution
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Create an output file for the deduplicated data
xdf_nodup <- tempfile(fileext = ".xdf")


# Get all the unique combinations of the default and year variables.
# Set dupFreqVar to get a count of how many records existed for each combination.
# Write the results to xdf_nodup.
rxSort(inData = xdf_source,
       outFile = xdf_nodup,
       sortByVars = c("default", "year"),
       removeDupKeys = TRUE
)


# Use rxDataStep to view the results
rxDataStep(xdf_sorted)

```

*** =sample_code
```{r, eval = FALSE}
# Path to the original XDF file
xdf_source <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Create an output file for the deduplicated data
xdf_nodup <- tempfile(fileext = ".xdf")


# Get all the unique combinations of the default and year variables.
# Set dupFreqVar to get a count of how many records existed for each combination.
# Write the results to xdf_nodup.



# Use rxDataStep to view the results
rxDataStep(xdf_sorted)

```

*** =sct
```{r, eval = FALSE}

test_error()

test_student_typed("rxSort",
                   not_typed_msg = "It doesn't appear that you made a call to rxSort - start there!")

test_student_typed("inData = xdf_source",
                   not_typed_msg = "Make sure you're specifying 'xdf_source' as the inData argument for rxSort.")

test_student_typed("outFile = xdf_nodup",
                   not_typed_msg = "Make sure you're specifying 'xdf_nodup' as the outFile argument for rxSort.")

test_student_typed('sortByVars = c("default", "year")',
                   not_typed_msg = 'Be sure to pass the variables creditScore and yearsEmploy to the argument sortByVars as a character vector: c("creditScore", "yearsEmploy")')

test_student_typed('removeDupKeys = TRUE',
                   not_typed_msg = "The 'decreasing' argument needs two values: one TRUE for creditScore, and one FALSE for yearsEmploy. Pass them as a logical vector: c(TRUE, FALSE)")

test_expression_result(rxDataStep(xdf_nodup),
                       incorrect_msg = "Your xdf_nodup file doesn't seem quite right - make sure that you've set the argument decreasing to TRUE.")



success_msg("That's right! As you can see, deduplication is a simple extension of sorting.")


```





--- type:VideoExercise

## Other Arguments 

*** =video_link

```{r, eval = FALSE}
## Data Management with RRE Sorting Deduplicating 05 Other Arguments
//player.vimeo.com/video/
```




