---
title       : Getting Started With MRS
description : Introduction to Microsoft R Server
---




--- type:VideoExercise

## Getting Started with MRS 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141380221
```




--- type:MultipleChoiceExercise contains_graph:FALSE

## Exercise 1

Which of these is true about DevelopR? 


*** =instructions

- DevelopR comes with the Linux installation of MRS
- DevelopR provides easier command line functionality
- DevelopR is a web services software development kit for integration analytics
- DevelopR is a visual development environment
- None of the above
  
*** =hint

Review the course video and choose the selection that most closely matches the 
addressed topic.

*** =pre_exercise_code

```{r,eval=FALSE}
```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 4, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Incorrect - DevelopR is part of Windows MRS installations.",
                          "Incorrect - try again.",
                          "Incorrect - that's the DeployR component.",
                          "Correct!",
                          "Incorrect - one of the above is true."
))

```


--- type:MultipleChoiceExercise contains_graph:FALSE

## Exercise 2

"rxLinMod()" is part of: 


*** =instructions

- DistributedR
- DevelopR
- DeployR
- ScaleR
- ConnectR
  
*** =hint

Review the course video and choose the selection that most closely matches the 
addressed topic.

*** =pre_exercise_code

```{r,eval=FALSE}
```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 4, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Incorrect - DistributedR is the MRS component for managing distributed computing contexts.",
                          "Incorrect - DevelopR is the visual development environment included with MRS.",
                          "Incorrect - DeployR is the MRS component for setting up web services.",
                          "Correct! ScaleR is the MRS component that includes statistical and machine learning algorithms like linear models.",
                          "Incorrect - ConnectR is the MRS component that enables connections to different data sources."
))

```



--- type:MultipleChoiceExercise contains_graph:FALSE

## Exercise 3

True or False: Specifying the platform is the only change a user needs to make 
to move a solution from one platform to another



*** =instructions

- True
- False
  
*** =hint

Review the course video and choose the selection that most closely matches the 
addressed topic.

*** =pre_exercise_code

```{r,eval=FALSE}
```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 1, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Correct! In most cases, MRS code can be moved to a new platform by simply specifying a different 'compute context'.",
                          "Incorrect - in most cases, MRS code can be moved to a new platform by simply specifying a different 'compute context'."
))
```





--- type:MultipleChoiceExercise contains_graph:FALSE

## Exercise 4

What is the main difference between a function with the prefix "rx" versus a function with the prefix "Rx"?



*** =instructions

- "rx" functions are a part of ScaleR while "Rx" functions are a part of ConnectR
- "Rx" functions aren't intended for use by analysts and programmers, while "rx" functions are
- "rx" functions are meant for regression problems while "Rx" functions are meant for classification problems
- There are no functional differences, and it is implemented randomly.  
  
*** =hint

Review the course video and choose the selection that most closely matches the 
addressed topic.

*** =pre_exercise_code

```{r,eval=FALSE}
```

*** =sct

```{r,eval=FALSE}

test_mc(correct = 1, 
        no_selection_msg = "Please make a selection!", 
        feedback_msgs = c("Correct! 'Rx' functions connect to data sources, while 'rx' functions include functions for data manipulation, aggregation, and building models.",
                          "Incorrect - 'Rx' functions are crucial when setting up data sources.",
                          "Incorrect - both regression and classification functions are part of the ScaleR component and therefore have the 'rx' prefix.",
                          "Incorrect - there's definitely a difference!"
))
```





