---
title_meta  : Chapter 2
title       : rxCube and rxCrossTabs
description : Functions for cross-tabulation and group-wise aggregation
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Categorical Variables

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141148548
```


--- type:NormalExercise xp:50

## Exercise 8: Crosstabulation
  
Crosstabulate with rxCrossTabs and rxCube, and compare the outputs

*** =instructions

It's intuitive that diamonds with higher `clarity` (that is, few natural blemishes)
would be likely to get a higher-grade `cut`. A good first step towarding 
investigating that idea would be to crosstabulate those two factors and see if
there's a noticeable relationship.

RRE includes two functions for crosstabs: `rxCrossTabs`, which produces the
classic frequency table, and `rxCube`, which produces a row for every unique
combination of its input variables. They both use the same formula syntax, with
a `:` separating the input variables:

    rxCrossTabs( ~ factor1 : factor2, data = somedata)
    rxCube( ~ factor1 : factor2, data = somedata)
    
 - Use **both** of these functions to crosstabulate the `cut` and `clarity` variables
on the `diamondsXdf` dataset, and take a moment to compare the differences in
their outputs.


Note: "I1" are the lowest-quality diamonds in this data, and "IF" are the highest.


*** =hint
All the code you need is in the instructions above - just replace `factor1`,
`factor2`, and `somedata` with the appropriate values. Be sure to run **both**
`rxCrossTabs` and `rxCube`, so you can compare the differences.

*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}
```

*** =solution

```{r,eval=FALSE}

# Compare the output of rxCrossTabs
rxCrossTabs( ~ clarity : cut, data = diamondsXdf)

# ... with the output of rxCube
rxCube( ~ clarity : cut, data = diamondsXdf)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxCrossTabs",
                   not_typed_msg = "It doesn't appear that you made a call to rxCrossTabs - be sure to call both functions.")

test_student_typed("rxCube",
                   not_typed_msg = "It doesn't appear that you made a call to rxCube - be sure to call both functions.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

test_student_typed(c(" ~ clarity : cut", " ~ cut : clarity"),
                   not_typed_msg = "Be sure to use the formula syntax, and put both clarity and cut on the right-hand side of the formula.")



test_output_contains("Cross Tabulation Results for: ~",
                     incorrect_msg = "Your rxCrossTabs results don't seem quite right - take a closer look.")

test_output_contains("Cube Results for: ~",
                     incorrect_msg = "Your rxCube results don't seem quite right - take a closer look.")


success_msg("Good work! That's all it takes to summarize a variable.")

```

*** =skills
1, 3

--- type:NormalExercise xp:50

## Exercise 9: Crosstabs for Three Variables
  
Add a third variable to your crosstabs and compare results


*** =instructions

The outputs of `rxCrossTabs` and `rxCube` are fairly similar for two variables,
but they look quite distinct for three. 

 - Add `color` to the previous crosstabs and compare the outputs.
 - Be sure to call both `rxCrossTabs` and `rxCube`.

*** =hint
Your formula should now look something like ` ~ factor1 : factor2 : factor3`

*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)

# Use the diamonds dataset from ggplot2
library(ggplot2)

# Create a path and a data source
diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))

# Select only a subsample of diamonds, to keep it fast
set.seed(2015)
rxImport(inData = diamonds[sample(nrow(diamonds), size = 1e4), ], outFile = diamondsXdf)


```


*** =sample_code

```{r,eval=FALSE}

```

*** =solution

```{r,eval=FALSE}

# Compare the output of rxCrossTabs
rxCrossTabs( ~ clarity : cut : color, data = diamondsXdf)

# ... with the output of rxCube
rxCube( ~ clarity : cut : color, data = diamondsXdf)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxCrossTabs",
                   not_typed_msg = "It doesn't appear that you made a call to rxCrossTabs - be sure to call both functions.")

test_student_typed("rxCube",
                   not_typed_msg = "It doesn't appear that you made a call to rxCube - be sure to call both functions.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use 'diamondsXdf' as your dataset; it's already been created for you.")

test_student_typed(c(" ~ clarity : cut : color", " ~ clarity : color : cut",
                     " ~ cut : clarity : color", " ~ cut : color : clarity",
                     " ~ color : cut : clarity", " ~ color : clarity : clarity"),
                   not_typed_msg = "Be sure to use the formula syntax, and put both clarity and cut on the right-hand side of the formula.")


test_output_contains("Cross Tabulation Results for: ~",
                     incorrect_msg = "Your rxCrossTabs results don't seem quite right - take a closer look.")

test_output_contains("Cube Results for: ~",
                     incorrect_msg = "Your rxCube results don't seem quite right - take a closer look.")

success_msg("Good work! That's all it takes to summarize a variable.")

```

*** =skills
1, 3


