---
title_meta  : Chapter 3
title       : Advanced Imports
description : Import data from a variety of sources
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Importing from Other Sources

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141145805
```







--- type:NormalExercise xp:50



## Exercise 6: Import Data from Multiple Files

Sometimes data is split across multiple files. You can use `rxImport` to combine
several source files into one XDF file.

*** =instructions

So far, we've only worked with one year of mortgage default data, but MRS comes
with ten years of that data. Each year has its own CSV file. In this exercise,
you'll combine all ten into a single XDF file.

The general framework has already been set up for you, so your task is to *write
the `rxImport` that will read in each CSV and append it to the XDF*. A couple
of hints:

 - You'll need to use the `inData`, `outFile`, and `append` arguments in `rxImport`
 - `append` has to be set to `FALSE` for the first CSV, and `TRUE` for every 
   CSV after the first. The `file.exists` function is one way to do that (see
   `?file.exists` for more details)


*** =hint

 - Inside of the `lapply()` function, the current CSV file path is stored as
   `x`. Pass that to the `inData` argument.
 - If you set `append = file.exists(mortgageXdf)`, `append` will be `FALSE` for
   the first CSV (because `rxImport` hasn't created the XDF yet) and `TRUE` for
   every subsequent CSV.



*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
file.remove("mortgage2000_2009.xdf")
```

*** =solution
```{r, eval=FALSE}
# Create a vector of all the mortgage default CSVs
mortgageCsvPaths <- list.files(path = rxGetOption("sampleDataDir"), 
                               pattern = "mortDefaultSmall\\d*.csv", 
                               full.names = TRUE)

# Create a pointer to a new XDF that will combine all the CSVs
mortgageXdf <- "mortgage2000_2009.xdf"

# Use lapply to loop through the files and create one large xdf
lapply(mortgageCsvPaths, FUN = function(x) {
  rxImport(inData = x,
           outFile = mortgageXdf,
           append = file.exists(mortgageXdf))
    }
)


# Use rxGetInfo to check your results.
# There should be a total of 10,000 files (1e+05, in R's scientific notation)
rxGetInfo(mortgageXdf)

```

*** =sample_code
```{r, eval=FALSE}
# Create a vector of all the mortgage default CSVs
mortgageCsvPaths <- list.files(path = rxGetOption("sampleDataDir"), 
                               pattern = "mortDefaultSmall\\d*.csv", 
                               full.names = TRUE)

# Create a pointer to a new XDF that will combine all the CSVs
mortgageXdf <- "mortgage2000_2009.xdf"

# Use lapply to loop through the files and create one large xdf
lapply(mortgageCsvPaths, FUN = function(x) {

    # Insert an rxImport function that reads in each CSV and
    # writes to the new XDF file.
    
    
})


# Use rxGetInfo to check your results.
# There should be a total of 10,000 files (1e+05, in R's scientific notation)
rxGetInfo(mortgageXdf)

```

*** =sct
```{r, eval=FALSE}

test_error()

test_student_typed("rxImport",
                   not_typed_msg = "Be sure to call the 'rxImport' function'!")

test_student_typed("inData = x",
                   not_typed_msg = "Be sure to pass x to your inData argument. Inside of lapply, x will stand in for each of the CSV files we're importing - one after the other.")

test_student_typed("outFile = mortgageXdf",
                   not_typed_msg = "Be sure to pass mortgageXdf to your outFile argument.")

test_student_typed("append = file.exists(mortgageXdf)",
                   not_typed_msg = "We can't simply set append to TRUE or FALSE in this case. To import the first file, append has to be FALSE - there's nothing to append to. For all subsequent CSV files, append has to be TRUE. The trick to this is to use the file.exists() ")



test_student_typed('rxGetInfo(mortgageXdf)',
                   not_typed_msg = "It's always a good idea to check that your file was imported correctly, so use the function 'rxGetInfo' on your new XDF file (mortgageXdf).")



test_expression_output(rxGetInfo(mortgageXdf)$numRows,
                       incorrect_msg = "Your XDF doesn't seem to have the right number of records - check your results again.")


success_msg("Looks good! This is a good example of combining functions from both open-source R and MRS to make your analysis simpler.")

```





--- type:NormalExercise xp:50


## Exercise 7: Import with a Data Source

MRS connects to different data sources using "data source" functions, but sometimes
you'll want to create a data source directly.

*** =instructions

In the previous exercises, `rxImport` has automatically created the "data source" 
wrappers for the CSV and XDF files behind the scenes. But creating data sources
manually can be useful for a variety of reasons:

 - Greater control over how data is imported to XDF (setting delimiters, SAS 
   import options, etc.)
 - The ability to use MRS functions directly on source data files if XDF files
   can't be used
 - The option to use some Open-Source R functions directly on files (CSV, SAS,
   XDF, etc.)
   
In this exercise, you'll examine the differences between three ways of calling
a dataset:

 - Working with the path to an XDF file (as we've done in exercises so far)
 - Working with an XDF data source object created by `RxXdfData`
 - Working with a text data source object created by `RxTextData`
 

Here's what you need to do:

 - Use `RxTextData` to create a text data source called `claimsCsvSource`.
 - Use `rxImport` to create an XDF file.
 - Use `rxGetInfo` on all three sources: `claimsXdf`, `claimsXdfSource`, and
   `claimsXdfSource`, to see how they differ.
 - Use Open-Source R's `head` function on all three sources to see what happens.
   You can try other Open-Source R functions, too: `tail`, `nrow`, `summary`, etc.





*** =hint

To create a data source, just pass the appropriate file path to the data source
creator function (`claimsCsv` to `RxTextData`, `claimsXdf` to `RxXdfData`).


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
file.remove("claims.xdf")
```

*** =solution
```{r, eval=FALSE}

# Here's the path to the source CSV
claimsCsv <- file.path(rxGetOption("sampleDataDir"), "claims.txt")


# 1. Use rxImport to import the CSV to claimsXdf
# As far as R knows, claimsXdf is just a character vector
claimsXdf <- "claims.xdf"

rxImport(inData = claimsCsv, outFile = claimsXdf, overwrite = TRUE)


# 2. Create an XDF data source with RxXdfData
# This creates an in-memory object that will work with some R functions
# (but it's still just a pointer to the XDF file)
claimsXdfSource <- RxXdfData(claimsXdf)

# 3. Create a text data source with RxTextData
# This also creates an in-memory data source, but it points
# directly to the CSV file
claimsCsvSource <- RxTextData(claimsCsv)


# View the imported data with rxGetInfo
rxGetInfo(claimsXdf, getVarInfo = TRUE)
rxGetInfo(claimsXdfSource, getVarInfo = TRUE)
rxGetInfo(claimsCsvSource, getVarInfo = TRUE)


# Use the Open-Source R function head() to view the first few rows.
# You might also try tail(), nrow(), and summary().
# Not every function will work for every source.
head(claimsXdf)
head(claimsXdfSource)
head(claimsCsvSource)


```

*** =sample_code
```{r, eval = FALSE}

# Here's the path to the source CSV
claimsCsv <- file.path(rxGetOption("sampleDataDir"), "claims.txt")


# 1. Use rxImport to import the CSV to claimsXdf
# As far as R knows, claimsXdf is just a character vector
claimsXdf <- "claims.xdf"




# 2. Create an XDF data source with RxXdfData
# This creates an in-memory object that will work with some R functions
# (but it's still just a pointer to the XDF file)


# 3. Create a text data source with RxTextData
# This also creates an in-memory data source, but it points
# directly to the CSV file



# View the imported data with rxGetInfo
rxGetInfo(claimsXdf, getVarInfo = TRUE)
rxGetInfo(claimsXdfSource, getVarInfo = TRUE)
rxGetInfo(claimsCsvSource, getVarInfo = TRUE)


# Use the Open-Source R function head() to view the first few rows.
# You might also try tail(), nrow(), and summary().
# Not every function will work for every source.
head(claimsXdf)
head(claimsXdfSource)
head(claimsCsvSource)

```

*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxImport",
                   not_typed_msg = "Be sure to call the 'rxImport' function!")

test_student_typed("RxXdfData",
                   not_typed_msg = "Be sure to call the 'RxXdfData' function to create your XDF data source.")

test_student_typed("RxTextData",
                   not_typed_msg = "Be sure to call the 'RxTextData' function to create your CSV data source.")






test_expression_output(rxGetInfo(claimsXdf),
                       incorrect_msg = "It seems like your XDF file wasn't created correctly - try examining it with the rxGetInfo function.")

test_expression_output(rxGetInfo(claimsXdfSource),
                       incorrect_msg = "It seems like your XDF data source wasn't created correctly - try examining it with the rxGetInfo function.")

test_expression_output(rxGetInfo(claimsCsvSource),
                       incorrect_msg = "It seems like your CSV data source wasn't created correctly - try examining it with the rxGetInfo function.")



success_msg("Looks good! This is a good example of combining functions from both open-source R and MRS to make your analysis simpler.")

```




--- type:VideoExercise

## Conclusion

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141145806
```



