---
title_meta  : Chapter 2
title       : More Complex Transformations
description : Transformations that need full-data access
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Scaling a Variable 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220974
```


--- type:NormalExercise xp:50

## Exercise 6: Standardizing a Variable
  
Standardize a variable using its mean and standard deviation

*** =instructions

In the video, I scaled a variable so that its lowest value was set to zero
and its highest value was one. Another common transformation is to standardize
a variable - to center it on its mean and then scale it by its standard deviation. This kind of transformation is essential for analyses that are
sensitive to the scale of the input variables, like principal components 
analysis.

In open-source R, you could standardize `x` like this: `(x - mean(x)) / sd(x)`.
In Revolution R Enterprise, the idea is the same, but we have to calculate
statistics like the mean and standard deviation outside of `rxDataStep`,
then subsequently pass them into `rxDataStep` using the `transformObjects`
argument.

For this exercise, use the `mortDefault` dataset to create a new variable called
`creditScoreStandardized`:

 - Use `rxSummary` to calculate the mean of `creditScore`
 - Use `rxSummary` to calculate the standard deviation of `creditScore`
 - Pass the mean and standard deviation to `rxDataStep` with the `transformObjects` argument
 - Calculate `creditScoreStandardized` using the `transforms` argument


*** =hint
To calculate the standardized credit score, use 
    (creditScore - varMean) /
    varSD


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Create a temporary XDF to work with
mortDefaultOriginal <- file.path(rxGetOption("sampleDataDir"),
                                 "mortDefaultSmall.xdf")

mortDefault <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = mortDefaultOriginal,
         outFile = mortDefault)

```


*** =sample_code

```{r,eval=FALSE}

# Use rxSummary to calculate the mean
creditScoreMean <- rxSummary( ~ creditScore, data = mortDefault)$sDataFrame$Mean

# And once more to calculate the standard deviation
creditScoreSD <- rxSummary( ~ creditScore, data = mortDefault)$sDataFrame$StdDev

# Create the standardized variable
rxDataStep(inData = mortDefault,
           outFile = mortDefault,
           transformObjects = list(varMean = __________,
                                   varSD = __________),
           transforms = list(creditScoreStandardized = __________ ), 
           overwrite = TRUE
)

```

*** =solution

```{r,eval=FALSE}

# Use rxSummary to calculate the mean
creditScoreMean <- rxSummary( ~ creditScore, data = mortDefault)$sDataFrame$Mean

# And once more to calculate the standard deviation
creditScoreSD <- rxSummary( ~ creditScore, data = mortDefault)$sDataFrame$StdDev

rxDataStep(inData = mortDefault,
           outFile = mortDefault,
           transformObjects = list(varMean = creditScoreMean,
                                   varSD = creditScoreSD),
           transforms = list(creditScoreStandardized = (creditScore - varMean) / varSD),
           overwrite = TRUE
)

```


*** =sct

```{r,eval=FALSE}
test_error()

test_student_typed("rxSummary",
                   not_typed_msg = "You'll need to use rxSummary to calculate the mean and standard deviation of creditScore.")

test_student_typed("rxDataStep",
                   not_typed_msg = "Use rxDataStep to create the standardized credit score on mortDefault.")

test_student_typed("transformObjects",
                   not_typed_msg = "Remember to pass creditScoreMean and creditScoreSD to rxDataStep using the transformObjects argument.")

test_student_typed("transformObjects = list",
                   not_typed_msg = "Remember to pass creditScoreMean and creditScoreSD to transformObjects inside of a list().")

test_student_typed("transforms = list(",
                   not_typed_msg = "Use the transforms argument to create the scald credit score.")

test_student_typed("varMean = creditScoreMean",
                   not_typed_msg = "In your transformObjects list, be sure to pass creditScoreMean with the name 'varMean', like this: varMean = creditScoreMean.")

test_student_typed("varSD = creditScoreSD",
                   not_typed_msg = "In your transformObjects list, be sure to pass creditScoreSD with the name 'varSD', like this: varSD = creditScoreSD.")

test_student_typed("overwrite = TRUE",
                   not_typed_msg = "Be sure to set overwrite = TRUE so that your changes are written to the XDF.")



test_that("The mortDefault XDF has been modified in some way, even if not totally correct", {
    expect_false(identical(rxDataStep(mortDefaultOriginal), 
                           rxDataStep(mortDefault)),
                 failure_msg = "The mortDefault XDF hasn't been changed - be sure to set 'overwrite = TRUE'.")
})

test_that("creditScoreStandardized has been created", {
    expect_true("creditScoreStandardized" %in% rxGetVarNames(mortDefault),
                failure_msg = "It looks like you haven't created a variable called 'creditScoreStandardized' on the mortDefault dataset. Try rxGetVarInfo(mortDefault) to see what variables exist on the dataset.")
})


test_expression_output(rxDataStep(mortDefault)$creditScoreStandardized,
                       incorrect_msg = "Your creditScoreStandardized doesn't have the expected values - check your code and try again.")


success_msg("Well done! You've made a simple transformation.")

```

*** =skills
1, 3
