---
title_meta  : Chapter 2
title       : Cross-Chunk Communication
description : Learn to communicate across chunks by lagging a variable to calculate change in price. 
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Lagging a Variable 

*** =video_link

```{r,eval = FALSE}
//player.vimeo.com/video/141147217
```


--- type:NormalExercise xp:50

## Exercise 4: Calculate Percent Change From Previous Day
  
For each trading day, calculate the percent change in closing price 
from the previous close. 


*** =instructions

In the previous exercise, you used `transformObjects` to distribute pre-calculated
statistics (the global minimum and maximum) simultaneously to all the XDF chunks.
But any time 
a computation requires passing results from one chunk to another, we have to use
a different set of tools: a custom `transformFunc` that uses the utility functions
`.rxSet`, `.rxGet`, `.rxStartRow`, and `.rxNumRows`.

In this exercise, your objective is to calculate the day-to-day change in closing
prices as a percentage of the previous day's price.

 - Complete the `price_change` function and use `rxDataStep` to apply it to the
`djia_small_chunks` dataset. 


*** =hint

 - `.rxSet` takes two arguments: a character string to use as a variable name
   and a value to assign to that variable. 
 - `.rxGet` takes a single argument: the name of a variable previously assigned
   by `.rxSet`.
 - To calculate the percent change between the `Close` and `previousClose` 
   columns, use the expression:
   
`100 * (dataList$Close - previousClose) / previousClose`

 - To apply `calc_price_change` to the XDF file, pass it to the `rxDataStep`
   argument `transformFunc`.


*** =pre_exercise_code
```{r,eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress = 0)

# Create a small subset of the DJIA data
djia_subset <- tempfile(fileext = ".xdf")

rxDataStep(inData = file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf"),
           outFile = djia_subset, 
           varsToKeep = "Close",
           transforms = list(Date = as.Date(Date)),
           rowSelection = (Date >= as.Date("2008-01-01")) & (Date <= as.Date("2008-01-31"))
)



# Cut it into 5-row chunks
djia_small_chunks <- RxXdfData( tempfile(fileext=".xdf") )

rxDataStep(inData = djia_subset,
           outFile = djia_small_chunks, 
           rowsPerRead = 5
)

file.remove(djia_subset)
rm(djia_subset)

```

*** =solution

```{r,eval = FALSE}

# Function for calculating day-to-day price changes. Remember, dataList will be
# a list containing all the variables of the input dataset in rxDataStep
calc_price_change <- function(dataList) {

    # For the very first day in the dataset, there's no way to calculate change,
    # so the value should be NA.
    if(.rxStartRow == 1) { val1 <- NA } else {
        
        # Otherwise, the first value for *this* chunk should be the last value
        # of the previous chunk. Use .rxGet to retrieve "last_value"; we'll create it
        # in a few more lines.
        val1 <- .rxGet("last_value")
    }
	
    
    # Shift all the prices ahead one day, and insert val1 at the top
	previousClose <- c(val1, dataList$Close[-.rxNumRows])
	
	# Calculate the percentage change between Close and previousClose
	dataList$pct_change <- 100 * (dataList$Close - previousClose) / previousClose
	
	# Make the last price from this chunk available for the next chunk with .rxSet
	.rxSet("last_value", dataList$Close[.rxNumRows])
	
	return(dataList)
	
}


# Make a file to store the transformed XDF output
djia_pct_change <- RxXdfData( tempfile(fileext = ".xdf") )

# Apply the new, improved transform to chunked data
rxDataStep(inData = djia_small_chunks, 
           outFile = djia_pct_change, 
           transformFunc = calc_price_change,
           transformVars = "Close"
)

# Check that it worked
head(djia_pct_change)

```


*** =sample_code

```{r,eval = FALSE}

# Function for calculating day-to-day price changes. Remember, dataList will be
# a list containing all the variables of the input dataset in rxDataStep
calc_price_change <- function(dataList) {

    # For the very first day in the dataset, there's no way to calculate change,
    # so the value should be NA.
    if(.rxStartRow == 1) { val1 <- NA } else {
        
        # Otherwise, the first value for *this* chunk should be the last value
        # of the previous chunk. Use .rxGet to retrieve "last_value"; we'll set it
        # in a few more lines.
        val1 <- _________
        
    
    # Shift all the prices ahead one day, and insert val1 at the top
	previousClose <- c(val1, dataList$Close[-.rxNumRows])
	
	# Calculate the percentage change between Close and previousClose.
	# Calculate the difference between Close and previousClose, then divide by
	# previousClose. Don't forget to multiply by 100 to make it a proper percentage.
	dataList$pct_change <- 100 * (dataList$Close - ________) / ________
	
	# Make the last price from this chunk available for the next chunk with .rxSet
	.rxSet("last_value", _________)
	
	return( dataList )
}

# Make a file to store the transformed XDF output
djia_pct_change <- RxXdfData( tempfile(fileext = ".xdf") )

# Apply the new, improved transform to chunked data
rxDataStep(inData = djia_small_chunks, 
           outFile = djia_pct_change, 
           transformFunc = ________
)

# Check that it worked
head(djia_pct_change)

```


*** =sct

```{r,eval = FALSE}

test_error()


# Check that the student typed...
test_student_typed("djia_small_chunks",
                   not_typed_msg = "Be sure to pass 'djia_small_chunks' to the inData argument in rxDataStep.")


# Check that the student typed...
test_student_typed("djia_pct_change",
                   not_typed_msg = "Be sure to pass 'djia_pct_change' to the outFile argument in rxDataStep to store your results.")

test_student_typed("transformFunc = calc_price_change",
                   not_typed_msg = "Be sure to use the transformFunc argument to pass calc_price_change to rxDataStep.")





# Check that the student's function is correctly defined
# Not sure how to do this yet





# Check that the student's output is correct
solution_df <- structure(list(Close = c(13043.9599609375, 13056.7197265625, 
12800.1796875, 12827.490234375, 12589.0703125, 12735.3095703125, 
12853.08984375, 12606.2998046875, 12778.150390625, 12501.1103515625, 
12466.16015625, 12159.2099609375, 12099.2998046875, 11971.1904296875, 
12270.169921875, 12378.6103515625, 12207.169921875, 12383.8896484375, 
12480.2998046875, 12442.830078125, 12650.3603515625), Date = structure(c(13880, 
13881, 13882, 13885, 13886, 13887, 13888, 13889, 13892, 13893, 
13894, 13895, 13896, 13900, 13901, 13902, 13903, 13906, 13907, 
13908, 13909), class = "Date"), pct_change = c(NA, 0.097821257219521, 
-1.96481232985799, 0.213360652285765, -1.85866383461423, 1.16163667516652, 
0.924832433693325, -1.92008335787449, 1.36321195434048, -2.16807621285908, 
-0.279576728223438, -2.46226738197815, -0.492714217802526, -1.05881643622359, 
2.4974917402204, 0.883772844043298, -1.38497314979997, 1.44767155445114, 
0.778512720857168, -0.300230981217508, 1.66787034890356)), .Names = c("Close", 
"Date", "pct_change"), class = "data.frame", row.names = c(NA, 
-21L))

    
test_what(
    expect_equivalent(object = rxDataStep(djia_pct_change),
                      expected = solution_df),
    feedback_msg = "Your output doesn't look quite right - try again."
)


success_msg("Well done. Passing data between chunks isn't easy, but it is essential for some operations.")

```

*** =skills
1, 3







--- type:VideoExercise


## Lagging by Merging 

*** =video_link

```{r,eval = FALSE}
//player.vimeo.com/video/141147220
```


