---
title_meta  : Chapter 2
title       : Types of Merges
description : Using rxMerge to join tables
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise key:12ee8e260f

## Types of Merges

*** =video_link

```{r, eval = FALSE}
//player.vimeo.com/video/147564261
```



--- type:NormalExercise xp:50 key:9203aeb223

## Exercise 3: Merge Two Datasets

Use `rxMerge` to join the datasets `customer` and `order` together using an
inner join.

*** =instructions

You've been provided with two `data.frames`: `customer`, which contains the names and
IDs of ten customers; and `order`, which contains the order ID, product name, 
and ID of the customer who placed the order.

 - Use the `nrow` function to see how many records are in each dataset. How many
rows do you think the joined dataset will have?

 - Use the `rxMerge` function to join these two datasets together into a single
new `data.frame`, using an inner join (`rxMerge` works on XDF files, too)

 - Use the `nrow` function to count how many rows the new `data.frame` has. Is it
as many as you expected?



*** =hint

 - You can specify the join type by setting `type = "inner"`, but the default join type for `rxMerge` is an inner join - so this step is optional.
 
 - Be sure to name the variable that the two datasets have in common, using
the `matchVars` argument.


*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
set.seed(100)

customer <- data.frame(customer_id = 1:10,
                       customer_name = c("Lin", "Karl", "Lupe", "Deshawn", "Colin",
                                          "Sushma", "Annie", "Muhammad", "Hannah", "Rajesh")
)

order <- data.frame(order_id = seq(238, length.out = 15),
                    customer_id = sample(1:10, size = 15, replace = TRUE),
                    order_date = sample(seq(Sys.Date(), length.out = 30, by = -1),
                                        size = 15, replace = TRUE),
                    product = sample(LETTERS[1:5], size = 15, replace = TRUE)
)


```


*** =solution
```{r, eval = FALSE}

# Examine the datasets
customer
order


# How many rows does each dataset have?
nrow(customer)
nrow(order)


# Join them together using rxMerge
inner_joined <- rxMerge(inData1 = customer, 
                        inData2 = order, 
                        matchVars = "customer_id")
    

    
# How many rows does the joined dataset have?
nrow(inner_joined)


```

*** =sample_code
```{r, eval = FALSE}

# Examine the datasets
customer
order


# How many rows does each dataset have?


# Join them together using rxMerge
inner_joined <- rxMerge(
    

    
# How many rows does the joined dataset have?





```

*** =sct
```{r, eval = FALSE}


test_error()

test_student_typed("rxMerge",
                   not_typed_msg = "There are many ways to do joins in R, but be sure to use the rxMerge() function for now.")

test_student_typed('matchVars = "customer_id"',
                   not_typed_msg = 'Make sure you set the matchVars argument to "customer_id"')


test_expression_output(nrow(inner_joined),
                       incorrect_msg = "Your 'inner_joined' data.frame doesn't seem to have the right number of rows - check your code again.")


test_object("inner_joined",
            undefined_msg = "It seems like the 'inner_joined' object hasn't been created; check your code again.",
            incorrect_msg = "Your 'inner_joined' doesn't look quite right - check it again.")

```






--- type:NormalExercise xp:50 key:f9f40cad41

## Exercise 4: Keep All Customer Records

Join `customer` and `order` using a left outer join

*** =instructions

In the last exercise, you used an inner join to merge `customer` and `order`.
Inner joins only keep the records that exist in both tables - so if a customer
hasn't placed any orders, an inner join would not include that customer in its
output. A left outer join will. 

 - Set `type = "left"` and merge the tables, then compare the results with the inner join.



*** =hint




*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
set.seed(100)

customer <- data.frame(customer_id = 1:10,
                       customer_name = c("Lin", "Karl", "Lupe", "Deshawn", "Colin",
                                          "Sushma", "Annie", "Muhammad", "Hannah", "Rajesh")
)

order <- data.frame(order_id = seq(238, length.out = 15),
                    customer_id = sample(1:10, size = 15, replace = TRUE),
                    order_date = sample(seq(Sys.Date(), length.out = 30, by = -1),
                                        size = 15, replace = TRUE),
                    product = sample(LETTERS[1:5], size = 15, replace = TRUE)
)


```


*** =solution
```{r, eval = FALSE}

# Here's the inner join from the previous exercise
inner_joined <- rxMerge(inData1 = customer, 
                        inData2 = order, 
                        matchVars = "customer_id")


# Now create a left outer join using the same datasets
left_joined <- rxMerge(inData1 = customer, 
                       inData2 = order, 
                       matchVars = "customer_id",
                       type = "left")


# Compare the results
inner_joined
left_joined

```

*** =sample_code
```{r, eval = FALSE}

# Here's the inner join from the previous exercise
inner_joined <- rxMerge(inData1 = customer, 
                        inData2 = order, 
                        matchVars = "customer_id")


# Now create a left outer join using the same datasets
left_joined <- rxMerge(

    


# Compare the results
inner_joined
left_joined


```

*** =sct
```{r, eval = FALSE}

test_error()

# test_student_typed("rxMerge",
#                    not_typed_msg = "There are many ways to do joins in R, but be sure to use the rxMerge() function for now.")

test_function("rxMerge", 
              args = c("matchVars", "type"), 
              index = 2,
              incorrect_msg = "Be sure to use both the 'matchVars' and 'type' arguments to construct your left join.",
              not_called_msg = "You'll need to call rxMerge at least once!")

test_student_typed('matchVars = "customer_id"',
                   not_typed_msg = 'Make sure you set the matchVars argument to "customer_id"')

test_student_typed('type = "left"',
                   not_typed_msg = 'Make sure you set the type argument to "left" to create a left outer join.')
                 

test_expression_output(nrow(left_joined),
                       incorrect_msg = "Your 'left_joined' data.frame doesn't seem to have the right number of rows - check your code again.")


test_object("left_joined",
            undefined_msg = "It seems like the 'left_joined' object hasn't been created; check your code again.",
            incorrect_msg = "Your 'left_joined' doesn't look quite right - check it again.")

```







--- type:NormalExercise xp:50 key:74a2ee991c

## Exercise 5: Stack Two Tables

Join `customer` and `new_customer` using a union join

*** =instructions

In this exercise, imagine your company has acquired a competitor. Now you've
got two tables with customer data in them: `customer` from the previous exercises,
and `new_customer`. Use `rxMerge` with `type = "union"` to combine these two 
tables into a new table called `all_customers`.



*** =hint




*** =pre_exercise_code
```{r, eval = FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
set.seed(100)

customer <- data.frame(customer_id = 1:10,
                       customer_name = c("Lin", "Karl", "Lupe", "Deshawn", "Colin",
                                          "Sushma", "Annie", "Muhammad", "Hannah", "Rajesh")
)

new_customer <- data.frame(customer_id = 11:20,
                           name = c("Jose", "Yuri", "Neil", "Fatma", "Sophia",
                                    "Tamar", "Brian", "Sarah", "Mei", "Joseph")
)

```


*** =solution
```{r, eval = FALSE}

# Union the two tables together
all_customers <- rxMerge(inData1 = customer, 
                         inData2 = new_customer, 
                         type = "union")

# Review the results
all_customers

```

*** =sample_code
```{r, eval = FALSE}

# Union the two tables together
all_customers <- rxMerge(


# Review the results
all_customers


```

*** =sct
```{r, eval = FALSE}

test_error()

test_function("rxMerge", 
              args = "type",
              incorrect_msg = "Be sure to use the 'type' argument to specify that you want a union join.",
              not_called_msg = "You'll need to call rxMerge to run the union query.")

test_student_typed('type = "union"',
                   not_typed_msg = 'Make sure you set the type argument to "union".')

test_expression_output(nrow(all_customers),
                       incorrect_msg = "Your 'all_customers' data.frame doesn't seem to have the right number of rows - check your code again.")


test_object("all_customers",
            undefined_msg = "It seems like the 'all_customers' object hasn't been created; check your code again.",
            incorrect_msg = "Your 'all_customers' doesn't look quite right - check it again.")

```









