---
title_meta  : Chapter 3
title       : Weighted Computation and Conclusion
description : Adapting existing functions for use with weighted data
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise key:24fee5e68c

## Weighted Data

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141148549
```

--- type:NormalExercise xp:50 key:fb28c2e31f

## Exercise 10: Weighted Means, Etc.

Calculate weighted descriptive statistics.


*** =instructions

 - Use `rxSummary` to calculate the average number of weeks worked in the
`censusData` dataset.

The "weeks worked" variable is named `wkswork1`; the weight variable is `perwt`.
Be sure to put quotation marks around `perwt` when you pass it to `rxSummary`: `"perwt"`


*** =hint

 - `rxSummary` takes a formula - something like ` ~ variable`
 - Because `perwt` is a count, it's a frequency weight - so use the `fweights` argument


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Create a pointer to the claims data
censusData <- RxXdfData(file.path(rxGetOption("sampleDataDir"), "CensusWorkers.xdf"))

```


*** =sample_code

```{r,eval=FALSE}
```

*** =solution

```{r,eval=FALSE}

rxSummary( ~ wkswork1, data = censusData, fweights = "perwt")

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxSummary",
                   not_typed_msg = "It doesn't appear that you made a call to rxSummary.")

test_student_typed("censusData",
                   not_typed_msg = "Be sure to use 'censusData' as your dataset; it's already been created for you.")

test_student_typed(" ~ wkswork1",
                   not_typed_msg = "Be sure to use the formula syntax, with 'wkswork1' on the right-hand side of the formula.")


test_output_contains("Summary Statistics Results for: ~wkswork1",
                     incorrect_msg = "Your rxCube results don't seem quite right - take a closer look.")

test_output_contains("Frequency weights: perwt",
                     incorrect_msg = "Your rxCube results don't seem quite right - take a closer look.")


success_msg("Good work! As you can see, using weighted data is straightforward.")
```

*** =skills
1, 3

--- type:NormalExercise xp:50 key:927ac7de8f

## Exercise 11: Weighted Crosstabs
  
Compute a frequency table from weighted data


*** =instructions

 - Using either `rxCrossTabs` or `rxCube`, create a table of `sex` and `state`
using `censusData`.


*** =hint

 - Both `rxCrossTabs` and `rxCube` take a formula, which should look something like: ` ~ factor1 : factor2`
 

*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Create a pointer to the claims data
censusData <- RxXdfData(file.path(rxGetOption("sampleDataDir"), "CensusWorkers.xdf"))

```


*** =sample_code

```{r,eval=FALSE}
```

*** =solution

```{r,eval=FALSE}

rxCrossTabs( ~ state : sex, data = censusData, fweights = "perwt")

rxCube( ~ state : sex, data = censusData, fweights = "perwt")

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed(c("rxCrossTabs", "rxCube"),
                   not_typed_msg = "It doesn't appear that you made a call to rxCrossTabs or rxCube.")

test_student_typed("censusData",
                   not_typed_msg = "Be sure to use 'censusData' as your dataset; it's already been created for you.")

test_student_typed(c(" ~ state:sex", " ~ sex:state"),
                   not_typed_msg = "Be sure to use the formula syntax, with state and sex on the right-hand side, separated by a colon (:).")



test_output_contains("Results for: ~",
                     incorrect_msg = "Your rxCube results don't seem quite right - take a closer look.")

test_output_contains("Frequency weights: perwt",
                     incorrect_msg = "Your rxCube results don't seem quite right - take a closer look.")

success_msg("Correct!")
```

*** =skills
1, 3




--- type:VideoExercise key:7844c21a17

## Conclusion

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141148553
```

