---
title_meta  : Chapter 1
title       : Advanced Transforms and the transformFunc Argument
description : We introduce advanced transforms that use functions rather than simple expressions to manipulate and generate variables in a dataset
attachments :
slides_link: 
framework   : datacamp
mode        : selfcontained


--- type:VideoExercise

## Introduction to Advanced Transformations

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141147218
```






--- type:VideoExercise

## Scaling Multiple Variables

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/141147216
```






--- type:NormalExercise xp:50

## Exercise 1: Scale Multiple Variables with transforms

In this exercise, you'll manually scale five variables from the Dow Jones 
Industrial Average dataset using the `transforms` argument in `rxDataStep`.
In the next exercise, you'll learn how to write a function to automate this 

Scaling is is the process of mapping values from their natural range (for
example, 50 to 150) to another range (in this exercise, 0 to 1).

*** =instructions

The Dow Jones dataset has been loaded for you; use the `djia_2008` object to
access it. Start by using `rxDataStep` to take a look at the dataset.

Write your scaled results to a new XDF file called `djia_scaled`; the code for
creating a temporary file has been provided for you.

Next, use `rxGetVarInfo` to get information about the global minimum and maximum 
values for the variables from the dataset, and store the results as an object
called `DJinfo`.

Now, use `rxDataStep` to scale the price columns - `Open`, `High`, `Low`, 
`Close` - as well as `Volume`:

 - Using the `transformObjects` argument, assign the `DJinfo` to the `var_info` 
   variable for use in the transforms expressions below.  
 - Using the `transforms` argument, write the R code to scale each variable.
   `Open` has already been scaled for you.


Finally, check your new scaled variables by using `rxDataStep` to examine
the first few rows of the modified data.


*** =hint

All the code you need has been provided, so review it carefully. If you need to,
you can reset the code to its original state by clicking the circling arrow
next to the green "Submit Answer" button.


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

djia_source <- RxXdfData( file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf") )
djia_2008 <- RxXdfData( tempfile(fileext=".xdf") )

rxDataStep(djia_source, djia_2008, 
            varsToKeep=c("Date", "Open", "High", "Low", "Close", "Volume"),
            transforms = list(Date = as.Date(Date)),
            rowSelection = (Date >= as.Date("2008-01-01")) & 
                           (Date <= as.Date("2008-12-31")),
            overwrite = TRUE)
```


*** =sample_code

```{r,eval=FALSE}

# Check out the first few lines of the data
rxDataStep(djia_2008, numRows = 5)

# Create a location to hold the scaled dataset
djia_scaled <- tempfile(fileext = ".xdf")

# Use rxGetVarInfo to calculate the minimums and maximums of the 
# variables in djia_2008
DJinfo <- _______

# Use rxDataStep to scale each variable individually
rxDataStep(
  inData = djia_2008,
  outFile = djia_scaled,
  transformObjects = list(_____ = _____),
  transforms = list(
    OpenScaled   = (Open - var_info$Open$low) / (var_info$Open$high - var_info$Open$low),
    HighScaled   = _____,
    LowScaled    = _____,
    CloseScaled  = _____,
    VolumeScaled = _____
  ),
  overwrite = TRUE
)

# Check your new variables
rxDataStep(djia_scaled, numRows = 5)
```

*** =solution

```{r,eval=FALSE}

# Use rxDataStep to see the first few lies of input data
rxDataStep(djia_2008, numRows = 5)

# Create a location to hold the scaled dataset
djia_scaled <- tempfile(fileext = ".xdf")

# Use rxGetVarInfo to calculate the minimums and maximums of the 
# variables in djia_2008
DJinfo <- rxGetVarInfo(djia_2008)

# Use rxDataStep to scale each variable individually
rxDataStep(
    inData = djia_2008,
    outFile = djia_scaled,
    
    # Pass the variable summary info using transformObjects
    transformObjects = list(var_info = DJinfo),
    
    # Use transforms to create the new variables
    transforms = list(
        OpenScaled   = (Open - var_info$Open$low) / (var_info$Open$high - var_info$Open$low),
        HighScaled   = (High - var_info$High$low) / (var_info$High$high - var_info$High$low),
        LowScaled    = (Low  - var_info$Low$low) / (var_info$Low$high - var_info$Low$low),
        CloseScaled  = (Close - var_info$Close$low) / (var_info$Close$high - var_info$Close$low),
        VolumeScaled = (Volume - var_info$Volume$low) / (var_info$Volume$high - var_info$Volume$low)
  )
)

# Check your new variables
rxDataStep(djia_scaled, numRows = 5)

```


*** =sct

```{r,eval=FALSE}


test_error()

test_student_typed("rxDataStep",
                   not_typed_msg = "It doesn't appear that you made a call to rxDataStep - start there!")

test_student_typed("djia_2008",
                   not_typed_msg = "Be sure to use 'djia_2008' as your dataset; it's already been created for you.")

test_student_typed("djia_scaled",
                   not_typed_msg = "Be sure to pass 'djia_scaled' to the outFile argument to store your results.")

test_student_typed("transformObjects = list(var_info = DJinfo",
                   not_typed_msg = "Be sure to use transformObjects to pass DJinfo - which contains the high and low values - to rxDataStep.")


# Make sure all of the scaled variables were created
scaled_vars_created <- c("OpenScaled", "HighScaled", "LowScaled", "CloseScaled",
                         "VolumeScaled") %in% rxGetVarNames(djia_scaled)

scaled_vars_not_created <- rxGetVarNames(djia_scaled)[!scaled_vars_created]

test_what(
    expect_true(all(scaled_vars_created)),
    feedback_msg = paste("Be sure to create a scaled version of each of the variables. It looks like you're missing:",
                         paste(scaled_vars_not_created, collapse = ", "))
)



# Check the values are correctly scaled
solution_df <- rxDataStep(
    inData = djia_2008,
    
    transformObjects = list(var_info = DJinfo),
    
    transforms = list(
        OpenScaled   = (Open - var_info$Open$low) / (var_info$Open$high - var_info$Open$low),
        HighScaled   = (High - var_info$High$low) / (var_info$High$high - var_info$High$low),
        LowScaled    = (Low  - var_info$Low$low) / (var_info$Low$high - var_info$Low$low),
        CloseScaled  = (Close - var_info$Close$low) / (var_info$Close$high - var_info$Close$low),
        VolumeScaled = (Volume - var_info$Volume$low) / (var_info$Volume$high - var_info$Volume$low)
        
  )
)

    
test_what(
    expect_equivalent(object = rxDataStep(djia_scaled),
                      expected = solution_df),
    feedback_msg = "You've created all of the scaled variables, but their values don't look right - take a closer look at your code, or reset to the sample code if you need to."
)




success_msg("Good work. There are more efficient ways to transform many variables at the same time, but this is the essential approach.")


```

*** =skills
1, 3





--- type:NormalExercise xp:50

## Exercise 2: Scale Multiple Variables with transformFunc

As you've seen, the `transforms` argument makes it easy to do simple 
row-by-row transformations, where the computed value of the new variable only 
depends on values in that same row. But there can be a lot of repetition 
involved in applying similar operations to multiple variables. In this exercise,
you'll use a more sophisticated approach: transform functions.  

Generalize the process of scaling a variable by writing a function. Instead of 
using the `transforms` argument to build new or modified variables from 
expressions, use `transformFunc` to pass a function. The parameters 
`transformVars` and `transformObjects` make variables from the dataset and 
other objects (respectively) available withn the transform function.


*** =instructions

First, review the `scalesVars` function. It's already laid out for you, but
you'll need to fill in the blanks to make it operational.

Next, use `rxGetVarInfo` to get the minimum and maximum values for all of the
variables in `djia_2008`.

Finally, use `rxDataStep` to apply the `scaleVars` function to all of the 
columns in `djia_2008`:

 - Use the `scaleVars` function as the `transformFunc`.
 - Give `transformVars` a character vector specifying which variables will 
   appear in the `dataList`. 
 - Pass `DJinfo` to the `transformObjects` argument to make it available as
   `var_info` inside of the `scaleVars` function.

Finally, examine the results by using `rxDataStep` with `numRows = 5`. You 
see the five new scaled variables in addition to the originals.


*** =hint

 - In the `scaleVars` function, the object `var_name` stands in for the variables
named in `transformVars`

 - To scale the variables, use the expression
`(var_name - low_value) / (high_value - low_value`.


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

djia_source <- RxXdfData( file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf") )
djia_2008 <- RxXdfData( tempfile(fileext=".xdf") )

rxDataStep(djia_source, djia_2008, 
            varsToKeep=c("Date", "Open", "High", "Low", "Close", "Volume"),
            transforms = list(Date = as.Date(Date)),
            rowSelection = (Date >= as.Date("2008-01-01")) & (Date <= as.Date("2008-12-31")),
            overwrite = TRUE)
```


*** =sample_code

```{r,eval=FALSE}
# Write a function to scale individual variables
scaleVars <- function(dataList) {
    
  # Loop over the names of the variables in dataList
  for(var_name in names(dataList)) {
      
    # Create the new variable name by pasting "Scaled" onto the original
  	dataList[[paste0(_____, "Scaled")]] <-
  	    
  	    # Subtract the minimum value from all the values...
  	    (dataList[[_____]] - var_info[[_____]]$low) / 
  	    
  	    # ... then divide by the difference between the min and max values
	        (var_info[[_____]]$high - var_info[[_____]]$low)
  }
    
  # Return dataList with its new variables
  return(dataList)

}

# Use rxGetVarInfo to calculate the minimums and maximums of the 
# variables in djia_2008
DJinfo <- _____

# Create a location to hold the scaled dataset
djia_scaled <- tempfile(fileext = ".xdf")

# Use rxDataStep to apply the scaleVars function 
# to "Open", "High", "Low", "Close", and "Volume"
rxDataStep(inData = djia_2008, 
           outFile = djia_scaled, 
           
           # transformFunc identifies a function you'd like to apply
           transformFunc = _____, 
           
           # transformVars names the variables you want to apply the function to
           # (that is, the variables that will be in dataList)
           transformVars = _____, 
           
           # transformObjects passes any other necessary objects to transformFunc
           transformObjects = list(var_info = ___)

)

# Examine the results
rxDataStep(djia_scaled, numRows = 5)
```

*** =solution

```{r,eval=FALSE}

# Write a function to scale individual variables
scaleVars <- function(dataList) {
    
  # Loop over the names of the variables in dataList
  for(var_name in names(dataList)) {
      
    # Create the new variable name by pasting "Scaled" onto the original
  	dataList[[paste0(var_name, "Scaled")]] <-
  	    
  	    # Subtract the minimum value from all the values...
  	    (dataList[[var_name]] - var_info[[var_name]]$low) / 
  	    
  	    # ... then divide by the difference between the min and max values
	        (var_info[[var_name]]$high - var_info[[var_name]]$low)
  }
    
  # Return dataList with its new variables
  return(dataList)

}

# Use rxGetVarInfo to calculate the minimums and maximums of the 
# variables in djia_2008
DJinfo <- rxGetVarInfo(djia_2008)

# Create a location to hold the scaled dataset
djia_scaled <- tempfile(fileext = ".xdf")

# Use rxDataStep to apply the scaleVars function 
# to "Open", "High", "Low", "Close", and "Volume"
rxDataStep(inData = djia_2008, 
           outFile = djia_scaled, 
           
           # transformFunc identifies a function you'd like to apply
           transformFunc = scaleVars, 
           
           # transformVars names the variables you want to apply the function to
           # (that is, the variables that will be in dataList)
           transformVars = c("Open", "High", "Low", "Close", "Volume"), 
           
           # transformObjects passes any other necessary objects to transformFunc
           transformObjects = list(var_info = DJinfo)

)

# Examine the results
rxDataStep(djia_scaled, numRows = 5)

```


*** =sct

```{r,eval=FALSE}

test_error()


# Check that the student typed...
test_student_typed("DJinfo <- rxGetVarInfo(djia_2008)",
                   not_typed_msg = "Be sure to use rxGetVarInfo on the 'djia_2008' dataset, and store the results as DJinfo.")

test_student_typed("djia_scaled",
                   not_typed_msg = "Be sure to pass 'djia_scaled' to the outFile argument to store your results.")

test_student_typed("transformFunc = scaleVars",
                   not_typed_msg = "Be sure to use transformFunc to pass scaleVars to rxDataStep.")

test_student_typed("transformVars = c(",
                   not_typed_msg = "Be sure to use transformVars to pass the names of the variables you want to transform to rxDataStep. It should be a list of quoted variable names - for example: c('Open', 'High')   ")

test_student_typed("transformObjects = list(var_info = DJinfo",
                   not_typed_msg = "Be sure to use transformObjects to pass DJinfo - which contains the high and low values - to rxDataStep.")




# Check that the student's function is correctly defined
test_function_definition(
    name = "scaleVars",
    test_expression_result({ test_data <- list(test1 = 1:10, test2 = 11:20);
                            var_info <- rxGetVarInfo(test_data);
                            scaleVars(test_data) }),
    undefined_msg = "It looks like you didn't specify the scaleVars function - try again."
)





# Check that the student's output is correct
# Make sure all of the scaled variables were created
scaled_vars_created <- c("OpenScaled", "HighScaled", "LowScaled", "CloseScaled",
                         "VolumeScaled") %in% rxGetVarNames(djia_scaled)

scaled_vars_not_created <- rxGetVarNames(djia_scaled)[!scaled_vars_created]

test_what(
    expect_true(all(scaled_vars_created)),
    feedback_msg = paste("Be sure to create a scaled version of each of the variables. It looks like you're missing:",
                         paste(scaled_vars_not_created, collapse = ", "))
)



# Check the values are correctly scaled
solution_df <- rxDataStep(
    inData = djia_2008,
    transformObjects = list(var_info = DJinfo),
    transforms = list(
        OpenScaled   = (Open - var_info$Open$low) / (var_info$Open$high - var_info$Open$low),
        HighScaled   = (High - var_info$High$low) / (var_info$High$high - var_info$High$low),
        LowScaled    = (Low  - var_info$Low$low) / (var_info$Low$high - var_info$Low$low),
        CloseScaled  = (Close - var_info$Close$low) / (var_info$Close$high - var_info$Close$low),
        VolumeScaled = (Volume - var_info$Volume$low) / (var_info$Volume$high - var_info$Volume$low)
  )
)

    
test_what(
    expect_equivalent(object = rxDataStep(djia_scaled),
                      expected = solution_df),
    feedback_msg = "You've created all of the scaled variables, but their values don't look right - take a closer look at your code, or reset to the sample code if you need to."
)


success_msg("Nicely done! This is one way to perform the same transformation on many variables.")

```

*** =skills
1, 3



--- type:NormalExercise xp:50


## Exercise 3: How to Break transformFunc
  
If you're still wondering why we can't just use `min()` and `max()` inside of
the `transforms` list, this exercise should help. Your goal is to write a bad
scaling function that appears to work - but returns incorrect results.


*** =instructions

Write a transform called `bad_scaleVars` that's similar to `scaleVars` from
the last exercise, with one difference: use `min()` and `max()` to find the
smallest and largest values, instead of calculating them with `rxGetVarInfo()`.

A small dataset, `simple_xdf`, has been provided for you. It contains one variable
that shows on which chunk of the XDF each record is stored, and two variables to
be transformed: `value1` and `value2`.

Use `rxDataStep` to apply `bad_scaleVars` to `value1` and `value2`, then view
the results. You'll see that there's a minimum (scaled value = 0) and maximum
(scaled value = 1) for *every* chunk, even though both variables have unique a
unique minimum and maximum value. This is why it's crucial to use `rxGetVarInfo`
to calculate global statistics when you're working on chunked data.


In short:

 - Create a `bad_scaleVars` function using `min` and `max`
 - Use `rxDataStep` to apply it to the `simple_xdf` XDF file
 - Pass `bad_scaleVars` to the `transformFunc` argument, and
 - Pass `value1` and `value2` to the `transformVars` argument


*** =hint

 - To calculate the minimum inside of the `bad_scaleVars` function, use: 
`min(dataList[[var_name]])`

 - The maximum is calculated the same way, just using `max` instead of `min`.

 - Pass `value1` and `value2` to `transformVars` as a character vector, like this:
 
    transformVars = c("value1", "value2")
    


*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)
set.seed(2015)

# Make a very simple data.frame
simple_df <- data.frame(chunk_id = LETTERS[sort(rep(1:5, 3))],
                        value1 = 1:15, 
                        value2 = sample(1:20, size = 15, replace = FALSE)
)

# Store in small chunks of 3 lines each
simple_xdf <- tempfile(fileext = ".xdf")

rxDataStep(inData = simple_df, 
           outFile = simple_xdf, 
           rowsPerRead = 3,
           overwrite = TRUE)

```


*** =sample_code

```{r,eval=FALSE}


# Examine the sample data
rxDataStep(simple_xdf)


# This "bad" version of scaleVars incorrectly assumes that `min` and `max` can
# be used to find the lowest and highest values of the target variable.
bad_scaleVars <- function(dataList) {
    
  # Loop over the names of the variables in dataList
  for(var_name in names(dataList)) {
      
    # Create the new variable name by pasting "Scaled" onto the original
  	dataList[[paste0(var_name, "Scaled")]] <-
  	    
  	    # Subtract the minimum value from all the values...
  	    (dataList[[var_name]] - __________) /
  	    
  	    # ... then divide by the difference between the min and max values
	        (__________ - __________)
  }
    
  # Return dataList with its new variables
  return(dataList)

}



# Use rxDataStep to apply bad_scaleVars to simple_xdf and
# output the results to scaled_xdf
scaled_xdf <- tempfile(fileext = ".xdf")

rxDataStep(inData = simple_xdf,
           outFile = scaled_xdf,
           transformFunc = __________,
           transformVars = __________,
)



# Check the results
rxDataStep(scaled_xdf)


```

*** =solution

```{r,eval=FALSE}

# Examine the sample data
rxDataStep(simple_xdf)


# This "bad" version of scaleVars incorrectly assumes that `min` and `max` can
# be used to find the lowest and highest values of the target variable.

bad_scaleVars <- function(dataList) {
    
  # Loop over the names of the variables in dataList
  for(var_name in names(dataList)) {
      
    # Create the new variable name by pasting "Scaled" onto the original
  	dataList[[paste0(var_name, "Scaled")]] <-
  	    
  	    # Subtract the minimum value from all the values...
  	    (dataList[[var_name]] - min(dataList[[var_name]])) /
  	    
  	    # ... then divide by the difference between the min and max values
	        (max(dataList[[var_name]]) - min(dataList[[var_name]]))
  }
    
  # Return dataList with its new variables
  return(dataList)

}



# Use rxDataStep to apply bad_scaleVars to simple_xdf and
# output the results to scaled_xdf
scaled_xdf <- tempfile(fileext = ".xdf")

rxDataStep(inData = simple_xdf,
           outFile = scaled_xdf,
           transformFunc = bad_scaleVars, 
           transformVars = c("value1", "value2")
)


# Check the results
rxDataStep(scaled_xdf)



```


*** =sct

```{r,eval=FALSE}


test_error()


# Check that the student typed...
test_student_typed("simple_xdf",
                   not_typed_msg = "Be sure to pass 'simple_xdf' to the outFile argument to store your results.")

test_student_typed("transformFunc = bad_scaleVars",
                   not_typed_msg = "Be sure to use the transformFunc argument to pass bad_scaleVars to rxDataStep.")

test_student_typed("transformVars = c(",
                   not_typed_msg = "Be sure to use transformVars to pass the names of the variables you want to transform to rxDataStep. It should be a list of quoted variable names - for example: c('Open', 'High')   ")




# Check that the student's function is correctly defined
test_function_definition(
    name = "bad_scaleVars",
    test_expression_result({ test_data <- list(test1 = 1:10, test2 = 11:20);
                            var_info <- rxGetVarInfo(test_data);
                            bad_scaleVars(test_data) }),
    undefined_msg = "It looks like you didn't specify the bad_scaleVars function - try again."
)





# Check that the student's output is correct
# Make sure all of the scaled variables were created
scaled_vars_created <- c("value1Scaled", "value2Scaled") %in% rxGetVarNames(scaled_xdf)

scaled_vars_not_created <- rxGetVarNames(scaled_xdf)[!scaled_vars_created]

test_what(
    expect_true(all(scaled_vars_created)),
    feedback_msg = paste("Be sure to create a scaled version of each of the variables. It looks like you're missing:",
                         paste(scaled_vars_not_created, collapse = ", "))
)



# Check the values are correctly (er, incorrectly) scaled
solution_df <- structure(list(
    chunk_id = structure(c(1L, 1L, 1L, 2L, 2L, 2L, 
3L, 3L, 3L, 4L, 4L, 4L, 5L, 5L, 5L), 
               .Label = c("A", "B", "C", "D", "E"), 
               class = "factor"), 
    value1 = 1:15, 
    value2 = c(2L, 16L, 6L, 1L, 3L, 18L, 7L, 20L, 8L, 15L, 12L, 4L, 17L, 19L, 5L),      
    value1Scaled = c(0, 0.5, 1, 0, 0.5, 1, 0, 0.5, 1, 0, 0.5, 1, 0, 0.5, 1), 
    value2Scaled = c(0, 1, 0.285714285714286, 0, 0.117647058823529, 1, 0, 1, 
                     0.0769230769230769, 1, 0.727272727272727, 0, 0.857142857142857, 1, 0)),
    .Names = c("chunk_id", "value1", "value2", "value1Scaled", "value2Scaled"), 
    class = "data.frame", 
    row.names = c(NA, -15L)
)

    
test_what(
    expect_equivalent(object = rxDataStep(scaled_xdf),
                      expected = solution_df),
    feedback_msg = "You've created all of the scaled variables, but their values don't look right - take a closer look at your code, or reset to the sample code if you need to."
)


success_msg("Good job! Notice how the 'maximum' scaled values correspond to many different numbers in the unscaled variables. That's how you can tell that maximums are being calculated within chunks.")

```

*** =skills
1, 3
