---
title_meta  : Chapter 3
title       : Additional Arguments
description : Less frequently-used but essential options for rxDataStep
attachments :
 slides_link: 
framework   : datacamp
mode        : selfcontained



--- type:VideoExercise key:4595c668f6

## transformPackages 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220996
```

--- type:NormalExercise xp:50 key:1548aa6473

## Exercise 7: transformPackages
  
Use a function from a package to extract a word from text


*** =instructions

Create a new variable called `make` on the `mtcarsXdf` dataset by extracting
the first word from the `model` variable. The package `stringr` includes a
variety of functions that make manipulating text data much easier. Use
`rxDataStep` and the expression:

    str_extract(string = model, pattern = "^\\w*")`
    
to create the new variable `make`.

Note:  `^\\w*` is an example of a "regular expression" - if you don't understand
what it means, don't worry. Regular expressions are a great skill to have but
are beyond the scope of this tutorial.


*** =hint
Try copy/pasting the `str_extract` code; it's quite sensitive to typos.

*** =pre_exercise_code
```{r,eval=FALSE}
library(RevoScaleR)
rxOptions(reportProgress=0)

# Convert mtcars to XDF
mtcarsXdf <- RxXdfData(tempfile(fileext = ".xdf"))
rxImport(inData = cbind(model = rownames(mtcars), mtcars), 
         outFile = mtcarsXdf)

# Create a second copy to see if the file is modified at all
mtcarsXdfOriginal <- RxXdfData(tempfile(fileext = ".xdf"))

rxImport(inData = cbind(model = rownames(mtcars), mtcars),
         outFile = mtcarsXdfOriginal)

```


*** =sample_code

```{r,eval=FALSE}

```

*** =solution

```{r,eval=FALSE}
rxDataStep(inData = mtcarsXdf,
           outFile = mtcarsXdf,
           transformPackages = "stringr",
           transforms = list(make = str_extract(string = model, pattern = "^\\w*")),
           overwrite = TRUE
)

```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxDataStep",
                   not_typed_msg = "Start with rxDataStep for this transformation.")

test_student_typed("transformPackages = 'stringr'",
                   not_typed_msg = "Be sure to pass the package name ('stringr') to the transformPackages argument.")

test_student_typed("transforms = list",
                   not_typed_msg = "You'll need to use the transforms argument to create the new variable (and remember, transforms always takes a list).")

test_student_typed("make = str_extract(",
                   not_typed_msg = "Use the function str_extract to create a new variable called make.")




test_that("mtcarsXdf has been modified in some way, even if not totally correct", {
    expect_false(identical(rxDataStep(mtcarsXdfOriginal), 
                           rxDataStep(mtcarsXdf)),
                 failure_msg = "mtcarsXdf hasn't been changed - be sure to set 'overwrite = TRUE'.")
})


test_that("make has been created", {
    expect_true("make" %in% rxGetVarNames(mtcarsXdf),
                failure_msg = "It looks like you haven't created a variable called 'make' on the mtcarsXdf dataset. Try rxGetVarInfo(mtcarsXdf) to see what variables exist on the dataset.")
})


test_expression_output(rxDataStep(mtcarsXdf)$make,
                       incorrect_msg = "Your make variable isn't quite right - take another look.")


success_msg("Well done! That's all it takes to use a package in your transformations.")

```

*** =skills
1, 3



--- type:VideoExercise key:4c5c16901a

## Factors 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220995
```


--- type:NormalExercise xp:50 key:130bd6b1ad

## Exercise 8: Factors
  
Create a simple factor 


*** =instructions

To create a factor using `rxDataStep`, we have to provide the names of all of
the levels of the factor to make sure that the factor is created correctly on
every chunk of data. For variables with only a few levels, this is no problem.

In this exercise, we'll be working with a dataset of the characteristics of
54,000 diamonds (from the package `ggplot2`). There are three character variables
on this dataset: `cut`, `color`, and `clarity`. I've provided the example code for
converting `clarity` to `clarityFactor`; now try to:

 - convert `cut` to `cutFactor`
 - convert `color` to `colorFactor`
 
 The unique values of a character variable become its levels when that
 variable is converted to a factor. To look up the unique values of a character
 variable, we can use the function `rxSort` and ask it to remove duplicates. I've
 provided example code for this to your right (and don't worry, we'll learn the
 details of `rxSort` in a later session).


*** =hint
 - Look up the unique levels of each factor using `rxSort`, then use `rxDataStep`
to create that factor.

 - Be sure to name the two new factors `cutFactor` and `colorFactor`, 
 exactly as shown - otherwise it's hard for us to check your work.


*** =pre_exercise_code
```{r,eval=FALSE}

library(RevoScaleR)
rxOptions(reportProgress=0)


# Use the diamonds dataset from ggplot2
library(ggplot2)

factor_indices <- sapply(diamonds, is.factor)
diamonds[factor_indices] <- lapply(diamonds[factor_indices], as.character)

diamondsXdf <- RxXdfData(tempfile(fileext = ".xdf"))
rxImport(inData = diamonds[1:1e4, ], outFile = diamondsXdf)
rm(diamonds)

```


*** =sample_code

```{r,eval=FALSE}

# Use rxSort to list the unique values of each variable. 
# Here's an example to get you started - just change "clarity" to the name of
# the variable you'd like to see.
rxSort(inData = diamondsXdf, removeDupKeys = TRUE,
       sortByVars = "clarity", varsToKeep = "clarity") 

# Use rxDataStep and the function factor() to create those variables
rxDataStep(inData = diamondsXdf,
           outFile = diamondsXdf,
           transforms = list(
               clarityFactor = factor(clarity,
                                      levels = c("I1", "IF", "SI1", "SI2", "VS1",
                                                 "VS2", "VVS1", "VVS2"))
           ),
           overwrite = TRUE
)

```

*** =solution

```{r,eval=FALSE}

# Use rxCube to identify the unique values of each variable
rxSort(inData = diamondsXdf, removeDupKeys = TRUE,
       sortByVars = "clarity", varsToKeep = "clarity") 
rxSort(inData = diamondsXdf, removeDupKeys = TRUE,
       sortByVars = "cut", varsToKeep = "cut") 
rxSort(inData = diamondsXdf, removeDupKeys = TRUE,
       sortByVars = "color", varsToKeep = "color") 

# Use rxDataStep and the function factor() to create those variables
rxDataStep(inData = diamondsXdf,
           outFile = diamondsXdf,
           transforms = list(
               clarityFactor = factor(clarity,
                                      levels = c("I1", "IF", "SI1", "SI2", "VS1",
                                                 "VS2", "VVS1", "VVS2")),
               cutFactor = factor(cut, levels = c("Fair", "Good", "Ideal",
                                                  "Premium", "Very Good")),
               colorFactor = factor(color, levels = c("D", "E", "F", "G", "H",
                                                      "I", "J"))
           ),
           overwrite = TRUE
)


```


*** =sct

```{r,eval=FALSE}

test_error()

test_student_typed("rxCube",
                   not_typed_msg = "Be sure to use rxCube to identify the levels of each variable")

test_student_typed("rxDataStep",
                   not_typed_msg = "You'll need to use rxDataStep to create the new variable.")

test_student_typed("diamondsXdf",
                   not_typed_msg = "Be sure to use the diamondsXdf dataset.")


test_that("cutFactor was created", {
    expect_true("cutFactor" %in% rxGetVarNames(diamondsXdf),
                failure_msg = "Make sure to create a new variable called 'cutFactor' on the diamondsXdf file."
    )  
})

test_that("colorFactor was created", {
    expect_true("colorFactor" %in% rxGetVarNames(diamondsXdf),
                failure_msg = "Make sure to create a new variable called 'colorFactor' on the diamondsXdf file."
    )  
})

test_expression_output(rxGetVarInfo(diamondsXdf)$cutFactor$varType,
                       incorrect_msg = "You created a variable called 'cutFactor', but it's not a factor - try again.")

test_expression_output(rxGetVarInfo(diamondsXdf)$colorFactor$varType,
                       incorrect_msg = "You created a variable called 'colorFactor', but it's not a factor - try again."
    )

test_expression_output(rxGetVarInfo(diamondsXdf)$cutFactor$levels,
                       incorrect_msg = "It looks like cutFactor is indeed a factor, but it seems to have the wrong levels - or they're in the wrong order. Check your code again.")

test_expression_output(rxGetVarInfo(diamondsXdf)$colorFactor$levels,
                       incorrect_msg = "It looks like colorFactor is indeed a factor, but it seems to have the wrong levels - or they're in the wrong order. Check your code again.")


success_msg("Excellent! You've made a simple factor. We'll cover more complex factors later on.")

```

*** =skills
1, 3





--- type:VideoExercise key:b75c358a50

## Conclusion 

*** =video_link

```{r,eval=FALSE}
//player.vimeo.com/video/137220994
```



