---
title: "Getting Started with RRE"
output: pdf_document
toc: FALSE
knit: "RevoKnitR::rx_compile_rmd"
navigation-symbols: FALSE
aspect-ratio: 169
number-frames: FALSE
fig_caption: no
---


```{r setup, echo = FALSE, results = FALSE}

rxOptions(reportProgress = 0)

opts_chunk$set(tidy = FALSE,
               cache = FALSE,
               output.max = 15
)

# I found more pictures:
# cp ~/../siddarth.ramesh/github/Revolution_Course_Materials/DataCamp/Getting_Started_With_RRE/src/images/*.png .

```


## Learning Objectives
* Packages constituting RRE
* Organization of functions
* Help and documentation


## RRE Overview

\center{
  \Oldincludegraphics[height=5.4cm]{images/rre_components.png}
}



## DevelopR: Windows IDE
\center{
  \Oldincludegraphics[height=5.4cm]{images/RPE_screenshot.png}
}




## DeployR: Web Services SDK
* APIs for Java, Javascript, & .NET 
* Invokes server-side R Scripts from web services calls
* RESTful interface
* Works well with web or mobile apps



## RevoScaleR

\center{
  \Oldincludegraphics[height=5.4cm]{images/RevoScaleR_architecture.png}
}




## ConnectR

Easy access to various data sources

* __SQL ODBC datasources:__ `RxOdBcData`
* __Hadoop HDFS:__ `RxHadoopMR`
* __Teradata databases:__ `RxTeradata`

## ScaleR

Functions for parallelized, scalable analytics

* __Data Preparation:__ `rxDataStep`, `rxFactors`, `rxSort`, `rxMerge`, ...
* __Descriptive Statistics and Summaries:__ `rxSummary`, `rxGetInfo`, `rxCrossTabs`, ...
* __Correlation and Covariance Matrixes:__ `rxCor`, `rxCov`, `rxCovCor`
* __Predictive Modeling:__ `rxLinMod`, `rxLogit`, `rxRoc`, ...
* __Machine Learning:__ `rxKmeans`, `rxDTree`, `rxBTrees`, ...
  
## DistributedR

Distributed computing framework based on compute contexts:

* Workstations
* SMP Servers
* Grid computing
* In-database
* In-Hadoop

## Changing Compute Contexts

```{r eval=FALSE, size="normalsize"}
rxSetComputeContext("localpar")
...
rxSetComputeContext(computeContext = myCluster)
```


## Examples of ScaleR Functions

* `rxImport`: generates XDF files from various data sources
* `rxDataStep`: multipurpose data filtering and cleaning
* `rxSummary`: univariate summaries of data
* `rxLinMod`: linear regression models
* `rxLogit`: logistic regression models
	
## RRE counterparts in R

```{r eval=FALSE, size="normalsize"}
# Open-Source R
lm(DepDelay ~ DayOfWeek, airlines_dataframe)

# RRE
rxLinMod(DepDelay ~ DayOfWeek, "airlines.xdf")
```

## Examples of ConnectR Functions

* `RxHadoopMR`: creates Hadoop Map Reduce compute context.
* `RxOdbcData`: generates ODBC data source object.
* `RxTeradata`: generates Teradata data source object.


## Connecting to an ODBC Data Source
```{r eval=FALSE, size = "small"}
conStr <- paste("Driver={SQL Server}", "Server=win-database01", 
                "Database=TestData", 
                "Uid=mktest", "Pwd=sqlpwd;", sep=";")

claimsDS <- RxOdbcData(sqlQuery="SELECT * FROM claims",
                       connectionString = conStr)

# Load data from the ODBC connection into an XDF file
claimsFile <- RxXdfData("claimsFromODBC.xdf")
rxImport(inData = claimsDS, outFile = claimsFile, overwrite = TRUE)
```


## Resources

* __Revolution Analytics website:__ http://revolutionanalytics.com/resources
* __User Guides:__ http://packages.revolutionanalytics.com/doc
* __Revolutions blog:__ http://blog.revolutionanalytics.com
* __Technical support:__ http://support.revolutionanalytics.com

## Built-in Help

```{r eval=FALSE, size="LARGE"}
help(package="RevoScaleR")
```

## RevoScaleR Package Documentation

\center{
  \Oldincludegraphics[height=5.4cm]{images/RevoScaleR_help.png}
}





## RRE Overview

\center{
  \Oldincludegraphics[height=5.4cm]{images/rre_components.png}
}





## Try this:

```{r eval=FALSE}
grep(x = "Revo.*", pattern = rownames(installed.packages()), value = TRUE)


?rxLinMod
```
