---
title: "Data Management with RRE:\nSimple Transformations"
output: pdf_document
toc: FALSE
knit: "RevoKnitR::rx_compile_rmd"
navigation-symbols: FALSE
aspect-ratio: 169
number-frames: FALSE
---

```{r setup, echo = FALSE, results = FALSE}

rxOptions(reportProgress = 0)

opts_chunk$set(tidy = FALSE,
               output.max = 15
)


```


## Set Up Sample Data

```{r import}

# First, set up pointers to the source XDF file
sourcePath <- file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf")

# And a temporary file
xdfPath <- tempfile(fileext = ".xdf")

# Import from the source file to our temporary one
rxImport(inData = sourcePath,
         outFile = xdfPath)

```

## Set Up Sample Data

```{r examine_data}

# Look at the top several rows
rxDataStep(xdfPath, numRows = 6)


```

## Set Up Sample Data

```{r examine_data2}

# Let's see what types those variables are
rxGetVarInfo(xdfPath)

```

## Converting Character to Date

```{r char_to_date}

rxDataStep(inData = xdfPath,
           outFile = xdfPath,
           transforms = list(Date2 = as.Date(Date)),
           overwrite = TRUE
)

```


## Converting Character to Date

```{r char_to_date_check, output.max = 15}

rxGetVarInfo(xdfPath)


```


## Transforms for Multiple Variables

```{r multitransform}

rxDataStep(inData = xdfPath,
           outFile = xdfPath,
           transforms = list(
               Date2 = as.Date(Date),
               combinedDate = paste(Year, Month, DayOfMonth, sep = "-"),
               Date3 = as.Date(combinedDate)
           ),
           overwrite = TRUE
)

```

## Multiple Transformations

```{r multitransform_check}

rxGetVarInfo(xdfPath)

```



## Dropping Variables


```{r lotta_vars}

rxDataStep(xdfPath, numRows = 6)

```

## Dropping Variables

```{r drop_vars}

xdfSubset <- tempfile(fileext = ".xdf")

rxDataStep(inData = xdfPath,
           outFile = xdfSubset,
           varsToDrop = c("Date", "Year", "Month", "DayOfMonth",
                          "DaysSince1928", "YearFrac", "combinedDate", "Date3")
)

```

## Dropping Variables

```{r drop_vars_check}

rxDataStep(xdfSubset, numRows = 6)


```


## Keeping Variables

```{r keep_vars}

xdfSubset <- tempfile(fileext = ".xdf")

rxDataStep(inData = xdfPath,
           outFile = xdfSubset,
           varsToKeep = c("Date2", "Open", "High", "Low", "Close", 
                          "Volume", "DayOfWeek")
)


```

## Keeping Variables

```{r keep_vars_check}

rxDataStep(xdfSubset, numRows = 6)

```





## Subsetting Rows: By Date

```{r subset_2000}

xdf2000 <- tempfile(fileext = ".xdf")

rxDataStep(inData = xdfSubset,
           outFile = xdf2000,
           rowSelection = Date2 > as.Date("1999-12-31")
)

```

## Subsetting Rows and Columns

```{r subset_MonFri}

xdfMonFri <- tempfile(fileext = ".xdf")

rxDataStep(inData = xdfSubset,
           outFile = xdfMonFri,
           rowSelection = DayOfWeek %in% c("Monday", "Friday"),
           varsToKeep = c("Date2", "DayOfWeek", "Open", "Close")
)

```




```{r delete_tempfiles, echo = FALSE, results = 'hide'}

# Feels like an excess of tempfiles is causing errors later - not sure
file.remove(xdf2000)
file.remove(xdfMonFri)

```



## Scaling (The Wrong Way)

```{r naive_scaling}

# Reasonable but wrong
rxDataStep(inData = xdfSubset,
           outFile = xdfSubset,
           transforms = list(VolumeScaled = 
               (Volume - min(Volume)) / (max(Volume) - min(Volume))
           ),
           overwrite = TRUE
)

```


## Scaling (The Wrong Way)

```{r naive_histogram, echo = FALSE}

library(ggplot2)

# Extract the data into a data.frame
compareVolumes <- rxDataStep(xdfSubset)

# Histogram of VolumeScaled
ggplot(compareVolumes, aes(x = VolumeScaled)) +
    geom_histogram(color = "white") +
    labs(title = "Scaled values between zero and one? Yep")


```


## Scaling (The Wrong Way)

```{r naive_scatter, echo = FALSE}

# Checking for linearity of transformation
ggplot(compareVolumes, aes(x = Volume, y = VolumeScaled)) +
    geom_point(alpha = 0.4) + 
    labs(title = "Scaled volume a linear function of volume? Nope")

```

## Scaling (The Wrong Way)

```{r naive_timeline, echo = FALSE}

# Melt down for comparative plotting
library(reshape2)

compareVolumesMelt <- melt(compareVolumes, 
                           id.var = "Date2",
                           measure.var = c("Volume", "VolumeScaled")
)

# When does the actual volume max happen? The scaled maxes?
volumeMaxes <- rbind(
    data.frame(Date2 = with(compareVolumes, Date2[Volume == max(Volume)]),
               variable = "Volume"),
    data.frame(Date2 = with(compareVolumes, Date2[VolumeScaled == 1]),
               variable = "VolumeScaled")
)


ggplot(compareVolumesMelt, aes(x = Date2, y = value, color = variable)) +
    geom_point() +
    geom_vline(data = volumeMaxes, 
               aes(xintercept = as.numeric(Date2))) +
    facet_wrap( ~ variable, 
               nrow = 2,
               scales = "free_y") +
    labs(title = "Comparing Unscaled and Scaled Maximums") +
    guides(color = FALSE)

```


## One Maximum Per Chunk

```{r chunk_count}
rxGetInfo(xdfSubset)
```


## Scaling (The Right Way)

```{r calc_minmax}

# Min and max values can be extracted from rxGetInfo
str(rxGetInfo(xdfSubset, getVarInfo = TRUE))

```


## Scaling (The Right Way)

```{r assign_minmax}

# Assign them for easy reference
volumeMin <- rxGetInfo(xdfSubset, getVarInfo = TRUE)$varInfo$Volume$low
volumeMax <- rxGetInfo(xdfSubset, getVarInfo = TRUE)$varInfo$Volume$high

```


## Scaling (The Right Way)

```{r tempfix, echo = FALSE, results = 'hide'}

file.remove(xdfSubset)

rxDataStep(inData = xdfPath,
           outFile = xdfSubset,
           varsToKeep = c("Date2", "Open", "High", "Low", "Close", 
                          "Volume", "DayOfWeek")
)

```

```{r scale_with_minmax}

rxDataStep(inData = xdfSubset,
           outFile = xdfSubset,
           transformObjects = list(volumeMin = volumeMin, 
                                   volumeMax = volumeMax),
           transforms = list(VolumeScaled = 
               (Volume - volumeMin) / (volumeMax - volumeMin)),
           overwrite = TRUE
)

```

## Scaling (The Right Way)


```{r correct_scatter, echo = FALSE}
compareVolumes2 <- rxDataStep(xdfSubset)

ggplot(compareVolumes2, aes(x = Volume, y = VolumeScaled)) +
    geom_point(alpha = 0.4) + 
    labs(title = "Scaled volume a linear function of volume? Yes!")
```

## Transform with a Package

```{r extract_date}

rxDataStep(inData = xdfSubset,
           outFile = xdfSubset,
           transformPackages = "lubridate",
           transforms = list(year = year(Date2)),
           overwrite = TRUE
)

```

## Creating Factors

```{r factors}

rxDataStep(inData = xdfSubset,
           outFile = xdfSubset,
           transforms = list(
               DayOfWeekFactor = factor(DayOfWeek,
                                        levels = c("Monday", "Tuesday",
                                                   "Wednesday", "Thursday",
                                                   "Friday"))),
           overwrite = TRUE
)

```


