---
title: "Data Management with RRE:\nAdvanced Transformations"
output: pdf_document
toc: FALSE
knit: "RevoKnitR::rx_compile_rmd"
navigation-symbols: FALSE
aspect-ratio: 169
number-frames: FALSE
---


```{r setup, include = FALSE}

rxOptions(reportProgress = 0)

opts_chunk$set(tidy = FALSE,
               cache = FALSE,
               output.max = 15
)


```



## Simple Transformations Review

Arguments in `rxDataStep`:

* creating and transforming variables: `transforms`
* subset rows: `rowSelection`
* drop variables: `varsToKeep` or `varsToDrop`
* create simple factors: use `factor` function with `transforms` 

## Logging and Exponentiation

```{r logging_exponentiating, eval=FALSE, size = "large", tidy = FALSE}
rxDataStep(myData, 
           transforms = list(ln_x = log(x),
                             exp_x = exp(x))
)
```

## A More Advanced Transformation

```{r scaling_example, eval = FALSE, tidy = FALSE}
djiaScaledXdf <- tempfile(fileext = ".xdf")

# Calculate the minimum and maximum
volumeMin <- rxGetInfo(djiaXdf, getVarInfo = TRUE)$varInfo$Volume$low
volumeMax <- rxGetInfo(djiaXdf, getVarInfo = TRUE)$varInfo$Volume$high

# Make the transformation (and drop a few variables, too)
rxDataStep(inData = djiaXdf, 
           outFile = djiaScaledXdf,
           varsToKeep = c("Date", "Open", "High", "Low", "Close", "Volume"),
           transformObjects = list(volumeMin = volumeMin, 
                                   volumeMax = volumeMax),
           transforms = list(VolumeScaled = 
               (Volume - volumeMin) / (volumeMax - volumeMin))
)


```

## First: Load Example Data

```{r load_DJIA_data, tidy = FALSE}
# First, set up pointers to the source XDF file
sourcePath <- file.path(rxGetOption("sampleDataDir"), "DJIAdaily.xdf")

# And a temporary file
djiaXdf <- tempfile(fileext = ".xdf")

# Import from the source file to our temporary one, convert Date to a Date
rxDataStep(inData = sourcePath,
           outFile = djiaXdf,
           transforms = list(Date = as.Date(Date)),
           overwrite = TRUE
)
```

## Examine XDF Info

```{r examine_local_copy_inf, output.max=12, tidy = FALSE}
rxGetInfo(djiaXdf, getVarInfo = TRUE)
```

## Examine XDF Data

```{r examine_local_copy_data, output.max=12, tidy = FALSE}
rxDataStep(djiaXdf, numRows = 10)
```


## Scale a Single Variable
```{r scale_one, tidy = FALSE}
djiaScaledXdf <- tempfile(fileext = ".xdf")

# Calculate the minimum and maximum
volumeMin <- rxGetInfo(djiaXdf, getVarInfo = TRUE)$varInfo$Volume$low
volumeMax <- rxGetInfo(djiaXdf, getVarInfo = TRUE)$varInfo$Volume$high

# Make the transformation (and drop a few variables, too)
rxDataStep(inData = djiaXdf, 
           outFile = djiaScaledXdf,
           varsToKeep = c("Date", "Open", "High", "Low", "Close", "Volume"),
           transformObjects = list(volumeMin = volumeMin, 
                                   volumeMax = volumeMax),
           transforms = list(VolumeScaled = 
               (Volume - volumeMin) / (volumeMax - volumeMin))
)

```



## Function to Scale Multiple Variables

```{r scale_many, tidy = FALSE}

# Define a custom scaling function.
scaleVars <- function(dataList) {

    # The variables we want to scale are now elements in dataList. 
    for(i in names(dataList)) {

        # For each variable "price", create a new variable "priceScaled"
        dataList[[paste0(i, "Scaled")]] <- 

            # Both dataList and varInfo can be indexed by the variable's
            # name, represented here as i
            (dataList[[i]] - varInfo[[i]]$low) / (varInfo[[i]]$high - varInfo[[i]]$low)
    }
    # Return dataList with the new scaled variables
    dataList
}
```

## Using the Custom Scaling Function

```{r rxGetInfo, tidy = FALSE}
# Collect all of the mins and maxes of the variables
DJinfo <- rxGetVarInfo(djiaXdf)
```

```{r rxGetInfo_output, echo = FALSE}
DJinfo
```


## Using the Custom Scaling Function

```{r use_scaleVars, tidy = FALSE}

# Use rxDataStep to apply the scaleVars function
rxDataStep(inData = djiaXdf, 
           outFile = djiaScaledXdf,
           varsToKeep = c("Date", "Open", "High", "Low", "Close", "Volume"),
           transformFunc = scaleVars,
           transformVars = c("Open", "High", "Low", "Close", "Volume"),
           transformObjects = list(varInfo = DJinfo),
           overwrite = TRUE
)

```

```{r scaleVars_results, echo = FALSE, tidy = FALSE}
rxDataStep(djiaScaledXdf, 
           varsToDrop = c("Date", "Open", "High", "Low", "Close", "Volume"),
           numRows = 6)

```

## Lagging: Example Data

```{r naive_lag_data, echo = FALSE, tidy = FALSE, output.max = 14}

# Generate some sample data
lagTestData <- data.frame(date = seq(as.Date("2015-01-01"), 
                                     by = "1 day",
                                     length.out = 12),
                          chunkID = (0:11 %/% 3) + 1, # Deal with it
                          Price = 1:12)

# Import to an XDF with four chunks
lagTestXdf <- tempfile(fileext = ".xdf")

# Load in chunks of 3 rows at a time
for (i in c(1, 4, 7, 10)) {
    rxImport(inData = lagTestData[i:(i + 2), ],
             outFile = lagTestXdf,
             append = file.exists(lagTestXdf))
}

rxDataStep(lagTestXdf, numRows = 12)

```

## Lagging: Naïvely

```{r naive_lag_demo, tidy = FALSE}
rxDataStep(inData = lagTestXdf, outFile = lagTestXdf,
           transforms = list(
               previousPriceNaive = c(NA, Price[-length(Price)])),
           overwrite = TRUE
)
```

```{r naive_lag_results, output.max = 8, echo = FALSE}
rxDataStep(lagTestXdf)
```


## Lagging: transformFunc

```{r lag_func, tidy = FALSE}

lagVar <- function(dataList) { 

    # Is this the first row in the entire dataset?
    if(.rxStartRow == 1) {
        # If so, the first value should be NA; drop the last value
        dataList[[newName]] <- c(NA, dataList[[varToLag]][-.rxNumRows]) 
    } else {
        # Otherwise, get the last value from the previous chunk; drop the last value
        dataList[[newName]] <- c(.rxGet("lastValue"),
                                 dataList[[varToLag]][-.rxNumRows]) 
    }

    # Make the last value available for the next chunk
    .rxSet("lastValue", dataList[[varToLag]][.rxNumRows])
    # Return dataList with the new variable
    dataList
}

```

## Lagging: transformFunc

```{r sort_lagVar, echo = FALSE}
lagTestSorted <- tempfile(fileext = ".xdf")

rxSort(inData = lagTestXdf,
       outFile = lagTestSorted,
       sortByVars = "date",
       overwrite = TRUE)

```

```{r use_lagVar, tidy = FALSE, output.max = 7}

# Be sure the dataset is sorted chronologically
rxDataStep(inData = lagTestSorted,
           transformObjects = list(
               varToLag = "Price", 
               newName = "previousPrice"), 
           transformFunc = lagVar,
           overwrite = TRUE)
```


## Lagging: By Merging

```{r lag_by_merging, tidy = FALSE, output.max = 15}

# Create a new dataset with every date shifted forward one day
lagTestShifted <- tempfile(fileext = ".xdf")
rxDataStep(inData = lagTestXdf, 
           outFile = lagTestShifted,
           varsToKeep = c("date", "Price"),
           transforms = list(date = date + 1))

# Merge on the original and shifted dates
rxMerge(inData1 = lagTestXdf, 
        inData2 = lagTestShifted,
        outFile = lagTestShifted,
        matchVars = "date", 
        type = "left", 
        duplicateVarExt = c("", "MergeLagged"),
        overwrite = TRUE
)

```



## Lagging: By Merging

```{r lag_by_merging_results, echo = FALSE, tidy = FALSE, output.max = 14}

rxDataStep(lagTestShifted)

```


## Simple Factors

```{r simple_factor, eval = FALSE, tidy = FALSE, output.max = 12}
rxDataStep(inData = djiaXdf,
           outFile = djiaFactor,
           transforms = list(
               DayOfWeekFactor = factor(DayOfWeek,
                  levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")))
)

```

## Create a Factor with rxFactors

```{r create_factor, echo = FALSE, tidy = FALSE}
djiaScrambled <- tempfile(fileext = ".xdf")
```

```{r scramble, tidy = FALSE}

# First, I'm going to scramble the Dow Jones dataset, so that we can see
# how the order of values affects the order of levels
rxSort(inData = djiaXdf,
       outFile = djiaScrambled,
       sortByVars = "Volume")

# Next, extract Year from Date
rxDataStep(inData = djiaScrambled, 
           outFile = djiaScrambled,
           transforms = list(charYear = substr(Date, 1, 4)),
           overwrite = TRUE)

rxFactors(inData = djiaScrambled, 
          outFile = djiaScrambled,
          factorInfo = list( factorYear = list(varName = "charYear") ),
          overwrite = TRUE)

```

## Aggregate by Factor

```{r aggregate_by_factor_fake, eval = FALSE, tidy = FALSE}
rxCube(Open ~ factorYear, data = djiaScrambled)
```

```{r aggregate_by_factor_output, echo = FALSE, tidy = FALSE}
as.data.frame(rxCube(Open ~ factorYear, data = djiaScrambled))
```


## Sort Factor

```{r sort_factor, tidy = FALSE}
rxFactors(inData = djiaScrambled, 
          outFile = djiaScrambled,
          factorInfo = list(
              factorYear = list(sortLevels = TRUE) ),
          overwrite = TRUE)
```

```{r sort_factor_fake_output, eval = FALSE, tidy = FALSE}
rxCube(Open ~ factorYear, data = djiaScrambled)
```

```{r sort_factor_output, echo = FALSE, tidy = FALSE}
as.data.frame(rxCube(Open ~ factorYear, data = djiaScrambled))[1:6, ]
```

## Custom Order

```{r custom_order, tidy = FALSE}
even_odd <- as.character(c(seq(1928, 2010, by = 2), 
                           seq(1929, 2009, by = 2)))

rxFactors(inData = djiaScrambled,
          outFile = djiaScrambled,
          factorInfo = list(factorYear = list(newLevels = even_odd)),
          overwrite = TRUE
)
```

```{r custom_order_fake_output, eval = FALSE}
rxCube(Open ~ factorYear, data = djiaScrambled)
```

```{r customer_order_output, echo = FALSE, tidy = FALSE}
as.data.frame(rxCube(Open ~ factorYear, data = djiaScrambled))[1:6, ]

```

## Recode Levels

```{r recode_newlevels, tidy = FALSE}
djiaRecode <- tempfile(fileext = ".xdf")

rxFactors(inData = djiaXdf, 
          outFile = djiaRecode,
          factorInfo = list(DayCombined = list( 
              varName = "DayOfWeek",
              newLevels = list(Monday = "Monday",
                               Midweek = "Tuesday", Midweek = "Wednesday", Midweek = "Thursday",
                               Friday = "Friday")))
)
```

```{r recode_fake_output, eval = FALSE}
rxCube(Open ~ DayCombined, data = djiaRecode)
```

```{r recode_output, echo = FALSE}
as.data.frame(rxCube(Open ~ DayCombined, data = djiaRecode))
```

## Recode otherLevel

```{r recode_otherlevel, tidy = FALSE}

rxFactors(inData = djiaXdf, 
          outFile = djiaRecode,
          factorInfo = list(
              DayCombined = list(newLevels = list(Monday = "Monday",
                                                  Friday = "Friday"),
                                 otherLevel = "Midweek",
                                 varName = "DayOfWeek")),
          overwrite = TRUE)



```

```{r otherlevel_fake_output, eval = FALSE}
rxCube(Open ~ DayCombined, data = djiaRecode)
```

```{r otherlevel_output, echo = FALSE}
as.data.frame(rxCube(Open ~ DayCombined, data = djiaRecode))
```
