---
title: 'Summary Statistics Slides'
output: pdf_document
toc: FALSE
knit: "msknit::msknit"
navigation-symbols: FALSE
aspect-ratio: 169
number-frames: FALSE
---


```{r setup, include = FALSE}


rxOptions(reportProgress=0)

library(ggplot2)

theme_set(theme_bw())

opts_chunk$set(tidy = FALSE,
               cache = FALSE,
               output.max = 15
)

set.seed(42)

```



## Summarize Credit Score

```{r summarize_one, output.max = 15, eval = FALSE, tidy = FALSE}
# Create a pointer to one of MRS's sample XDFs
defaultData <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Summarize credit score
rxSummary(~ creditScore, data = defaultData)
```

```{r summarize_one_output, echo = FALSE, tidy = FALSE}
defaultData <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

rxSummary(~ creditScore, data = defaultData)$sDataFrame
```


## Summarize Two Variables
```{r summarize_two, eval = FALSE, tidy = FALSE}

rxSummary( ~ creditScore + ccDebt, data = defaultData)

```

```{r summarize_two_output, echo = FALSE, tidy = FALSE}

rxSummary( ~ creditScore + ccDebt, data = defaultData)$sDataFrame

```


## Summarize All Variables
```{r summarize_all, eval = FALSE, tidy = FALSE}

rxSummary( ~ ., data = defaultData)

```

```{r summarize_all_output, echo = FALSE, tidy = FALSE}

rxSummary( ~ ., data = defaultData)$sDataFrame

```



## Data Structure

```{r summary_str, output.max = 12, tidy = FALSE}

# Assign rxSummary to an object
defaultSummary <- rxSummary( ~ ., data = defaultData)

# Examine the structure of the results
str(defaultSummary)

```



## Plot Summary Statistics

```{r plotSummary, tidy = FALSE}

# Extract the data frame
summaryDF <- defaultSummary$sDataFrame


# Plot the means
ggplot(summaryDF, aes(x = Name, y = Mean)) + geom_point(size = 10) +
    labs(x = "Variable", y = "Variable Mean")

```



## Groupwise Summary

```{r summary_by_group, eval = FALSE, tidy = FALSE}

rxSummary(creditScore ~ F(year), data = defaultData)

```

```{r summary_by_group_output, echo = FALSE, tidy = FALSE}

rxSummary(creditScore ~ F(year), data = defaultData)$categorical[[1]]

```



## Extracting a Data Frame

```{r findSummaryCat, eval = FALSE, tidy = FALSE}

scoreByYear <- rxSummary(creditScore ~ F(year), data = defaultData)

str(scoreByYear)

```


```{r findSummaryCat_output, output.max = 14, echo = FALSE}

scoreByYear <- rxSummary(creditScore ~ F(year), data = defaultData)

scoreByYear[[3]] <- data.frame()

str(scoreByYear)

```



## rxQuantile

```{r rxQuantile, tidy = FALSE}

rxQuantile(varName = "creditScore",
           data = defaultData)

```


## Custom Quantiles

```{r nondefault_quantiles, tidy = FALSE}

rxQuantile(varName = "creditScore", 
           probs = c(0.025, 0.975),
           data = defaultData)

```



## Add A Variable

```{r add_cat, tidy = FALSE}

# Create a new path for this modified file
defaultTemp <- tempfile(fileext = ".xdf")

# Import the source data, adding a categorical creditScore variable
rxDataStep(inData = defaultData,
           outFile = defaultTemp,
           transforms = list(creditScoreCat = cut(creditScore,
                                                  breaks = seq(from = 400,
                                                               to = 900,
                                                               by = 100)))
           )


```



## Default Counts

```{r rxcrosstabs_univar, eval = FALSE, tidy = FALSE}

rxCrossTabs( ~ F(default), data = defaultTemp)

```

```{r rxcrosstabs_univar_output, echo = FALSE, tidy = FALSE}
rxCrossTabs( ~ F(default), data = defaultTemp)[["counts"]][[1]]
```

```{r rxcube_univar, eval = FALSE, tidy = FALSE}

rxCube( ~ F(default), data = defaultTemp)

```

```{r rxcube_output, echo = FALSE, tidy = FALSE}
as.data.frame(rxCube( ~ F(default), data = defaultTemp))
```


## Bivariate Tables

```{r bivar_tables_syntax, eval = FALSE, tidy = FALSE}

rxCrossTabs( ~ F(default) : creditScoreCat, data = defaultTemp)

rxCube( ~ F(default) : creditScoreCat, data = defaultTemp)

```


## Bivariate Tables: rxCrossTabs

```{r rxcrosstabs_bivar, eval = FALSE, tidy = FALSE}
rxCrossTabs( ~ F(default) : creditScoreCat, data = defaultTemp)
```

```{r rxcrosstabs_bivar_output, echo = FALSE, tidy = FALSE}
rxCrossTabs( ~ F(default) : creditScoreCat, data = defaultTemp)[["counts"]][[1]]
```


## Bivariate Tables: rxCube

```{r rxcube_bivar, eval = FALSE, tidy = FALSE}
rxCube( ~ F(default) : creditScoreCat, data = defaultTemp)

```

```{r rxcube_bivar_output, echo = FALSE, output.max = 12, tidy = FALSE}
as.data.frame(rxCube( ~ F(default) : creditScoreCat, data = defaultTemp))
```



## Three Variables: rxCrossTabs

```{r rxcrosstabs_trivar, eval = FALSE, tidy = FALSE}
rxCrossTabs( ~ F(default) : creditScoreCat : F(year), data = defaultTemp)
```

```{r rxcrosstabs_trivar_output, echo = FALSE, output.max = 14, tidy = FALSE}
rxCrossTabs( ~ F(default) : creditScoreCat : F(year), data = defaultTemp)$counts
```


## Three Variables: rxCube

```{r rxcube_trivar, eval = FALSE, tidy = FALSE}
rxCube( ~ F(default) : creditScoreCat : F(year), data = defaultTemp)
```

```{r rxcube_trivar_output, echo = FALSE, output.max = 12, tidy = FALSE}
as.data.frame(rxCube( ~ F(default) : creditScoreCat : F(year), data = defaultTemp))
```



## rxCube with ggplot

```{r plot_cube_wrong, error = TRUE, tidy = FALSE}

creditScoreCounts <- rxCube( ~ creditScoreCat, data = defaultTemp)

# It's actually a list
class(creditScoreCounts)

# ggplot doesn't know what to do with a list
ggplot(creditScoreCounts, aes(x = creditScoreCat, weights = Counts)) +
    geom_bar()

```



## rxCube data.frame with ggplot

```{r plot_cube_correct, error=TRUE, tidy = FALSE}

creditScoreCountsDF <- rxCube( ~ creditScoreCat, data = defaultTemp,
                              returnDataFrame = TRUE)

# Now a data.frame
class(creditScoreCountsDF)

```

```{r plot_cube_output, echo = FALSE, tidy = FALSE}
# No problem plotting
ggplot(creditScoreCountsDF, aes(x = creditScoreCat, weights = Counts)) +
    geom_bar()
```



## Weighted Data

```{r get_census, tidy = FALSE}

censusData <- file.path(rxGetOption("sampleDataDir"), "CensusWorkers.xdf")

rxDataStep(censusData, numRows = 10)


```



## Help for Dataset

```{r census_help, eval = FALSE, tidy = FALSE}
?CensusWorkers
```


\center{
  \Oldincludegraphics[height=5.0cm]{images/CensusWorkersHelp.png}
}


## Frequency Weights

```{r fweights, eval = FALSE, tidy = FALSE}

rxCube( ~ state : sex,
       fweights = "perwt",
       data = censusData)

```

```{r fweights_output, echo = FALSE, tidy = FALSE}
as.data.frame(
    rxCube( ~ state : sex,
           fweights = "perwt",
           data = censusData)
)
```



## Income Summary

```{r inc_state_summary, eval = FALSE, tidy = FALSE}

rxSummary(incwage ~ state : sex,
          fweights = "perwt",
          data = censusData)

```

```{r inc_state_summary_output, echo = FALSE, tidy = FALSE, output.max = 15, size = "tiny"}
options(width = 115)
rxSummary(incwage ~ state :  sex,
          fweights = "perwt",
          data = censusData)$categorical[[1]]
```
