---
title: "Data Management with RRE:\nMerging Data"
output: pdf_document
toc: FALSE
knit: "RevoKnitR::rx_compile_rmd"
navigation-symbols: FALSE
aspect-ratio: 169
number-frames: FALSE
---




```{r setup, include = FALSE}

library(knitr)
library(sqldf)

rxOptions(reportProgress = 0)

opts_chunk$set(tidy = FALSE,
               cache = FALSE,
               output.max = 15,
               warning = FALSE,
               message = FALSE
)


# Set the data directory
data_dir <- "data"

```




```{r lib, echo=FALSE}


runSQL <- function(myQuery){
  library(RSQLite)
  db_file <- file.path(data_dir,
                       "top3_airlines_Jan_2000.sqlite3")
  db <- dbConnect(SQLite(), db_file)
  tbl <- dbGetQuery(db, myQuery)
  dbDisconnect(db)
  return(tbl)
}

demoSQL <- function(myQuery, n=6){
  tbl <- runSQL(myQuery)
  cat(myQuery)
  knitr::kable(head(tbl, n=n))
}

SQL_to_kable <- function(myQuery, n=6){
  tbl <- runSQL(myQuery)
  knitr::kable(head(tbl, n=n))
}

sqldf_to_kable <- function(myQuery) kable(sqldf(myQuery))

```




## Airline Data
```{r, echo=FALSE}
flight <- runSQL("SELECT UniqueCarrier, FlightNum, Year, Month, DayOfMonth, Origin, Dest, TailNum, DayOfWeek, ArrTime, DepTime, ArrDelay FROM flight LIMIT 8;")
carrier <- runSQL("SELECT Code, Description FROM carrier WHERE Code IN ('DL','UA','WN','AA','6R');")
airline <- runSQL("SELECT tailnum, model, year FROM plane WHERE model != '' LIMIT 3;")
```

\center{\Oldincludegraphics[height=5.4cm]{images/airline_tables.png}}





## Airline Database

\center{\Oldincludegraphics[height=5.7cm]{images/airlines_ERD_edited_1.png}}





## Composite Key

```{r flights_composite_key, include = FALSE}
# https://en.wikipedia.org/wiki/Flight_number
composite_key_sql <- "
SELECT UniqueCarrier, FlightNum, Year, Month, 
    DayofMonth, DepTime, TailNum, Origin, Dest  
FROM flight
WHERE FlightNum=789 AND Year=2000 AND Month=1 AND DayofMonth=1 
ORDER BY UniqueCarrier, DepTime;"
```

```{r print_composite_key, eval = FALSE, size = "scriptsize"}
SELECT UniqueCarrier, FlightNum, Year, Month, DayofMonth, DepTime, TailNum, Origin, Dest  
FROM flight
WHERE FlightNum = 789 
  AND Year = 2000 
  AND Month = 1
  AND DayofMonth = 1 
ORDER BY UniqueCarrier, DepTime;
```

```{r, echo = FALSE}
SQL_to_kable(composite_key_sql, n=12)
```




## Analysis Table
```{r denormalized_table, echo=FALSE, eval=FALSE}
denormalized_tbl <- runSQL("SELECT UniqueCarrier, carrier.Description Carrier, FlightNum, flight.Year, Month, DayofMonth, Origin, Dest, DayOfWeek, ArrTime, DepTime, plane.model AS plane_model, plane.year as plane_year, ArrDelay
  FROM flight
	INNER JOIN carrier
	ON flight.UniqueCarrier = carrier.Code
	LEFT OUTER JOIN plane
	ON flight.TailNum = plane.tailnum 
	WHERE DayofMonth=1 AND plane.model IS NOT NULL
	LIMIT 10")
```

\center{\Oldincludegraphics[height=4.9cm]{images/denormalized.png}}

## External Table

\center{\Oldincludegraphics[height=5.7cm]{images/airlines_ERD_edited_2.png}}





## Toy Datasets for Merges

```{r toy_merges_setup, echo=FALSE}
uc <- data.frame(id=1:6, letter=LETTERS[1:6])
scrambled_id <- sample(4:9)
lc <- data.frame(id=scrambled_id, letter=letters[scrambled_id])
```

\center{\Oldincludegraphics[height=5.7cm]{images/uc_lc.png}}




## Inner Join
```{r inner_join_sql}
sqldf("SELECT * FROM uc JOIN lc USING (id)")
```




## Inner Joins with merge and rxMerge
```{r toy_merge_inner_join_2}
merge(uc, lc, by="id")
rxMerge(uc, lc, matchVars = "id")
```





## Inner Join is the Default in rxMerge
```{r}
# In rxMerge, the "type" parameter specifies the kind of join.
rxMerge(inData1 = uc, inData2 = lc, matchVars = "id", type="inner")
```




## Inner Join is the Default in SQL
```{r}
# "JOIN" in SQL means "INNER JOIN" unless you say otherwise.
sqldf("SELECT * FROM uc INNER JOIN lc USING (id)")
```




## Left Outer Join
```{r left_outer_joins, tidy=FALSE}
sqldf("SELECT uc.id, uc.letter letter_uc, lc.letter letter_lc 
       FROM uc LEFT OUTER JOIN lc USING (id)")
```




## Left Outer Join
```{r left_outer_join_merge_rxMerge, output.max = 5}
# merge from base R
merge(uc, lc, by="id", all.x=TRUE)

# rxMerge
rxMerge(inData1 = uc, inData2 = lc, matchVars = "id", type="left")
```




## Union

\center{\Oldincludegraphics[height=5.4cm]{images/Three_Unions.png}}





## oneToOne
```{r oneToOne_join, output.max = 5}
rxMerge(inData1 = uc, inData2 = lc, matchVars = "id", type="oneToOne")
cbind(uc, lc)
```




## Merging Multiple Tables

```{r multiple_join, eval = FALSE, tidy = FALSE}

# Joining three tables in RRE
rxMerge(inData = list(xdf1, xdf2, xdf3),
        matchVars = "id")


# Joining three tables in SQL
SELECT * 
FROM table1 
JOIN table2 
    USING (id) 
JOIN table3 
    USING (id)
```




## Merging Airline Tables

```{r airline_join_sql, echo=FALSE, message=FALSE}
sql <- "SELECT flight.*, carrier.Description Carrier, plane.*
    FROM flight 
INNER JOIN carrier
    ON flight.UniqueCarrier = carrier.Code
LEFT OUTER JOIN plane
    ON flight.TailNum = plane.tailnum"

demoSQL(sql)

```




## SQL Query in RxOdbcData Object

```{r rxodbcdata_pre, include = FALSE}
data_dir <- "data"
airlines_db <- file.path(data_dir,"top3_airlines_Jan_2000.sqlite3")
airlines_xdf <- file.path(data_dir, "top3_airlines_Jan_2000_analysis.xdf")
```

```{r connect_odbc, tidy = FALSE, eval = FALSE}
odbc_connection <- "Driver={SQLite3 ODBC Driver};Database=data/top3_airlines_Jan_2000.sqlite3"

odbc_data <-
  RxOdbcData(sqlQuery = "SELECT flight.*, carrier.Description Carrier, plane.* 
                         FROM flight 
                         INNER JOIN carrier
                             ON flight.UniqueCarrier = carrier.Code
                         LEFT OUTER JOIN plane
                        	 ON flight.TailNum = plane.tailnum",
             connectionString = odbc_connection)

head(odbc_data)
```




## Importing to XDF: colClasses and colInfo

```{r import_odbc, eval = FALSE, size = "tiny", tidy = FALSE}

colClasses <- sapply(head(odbc_data, n=1000), class)
colClasses[c("engine_type")] <- "factor"
colClasses[c("Year", "year", "DayOfMonth", "DayOfWeek", "DepTime", "ArrTime")] <- "uint16"

colInfo <- list(
    UniqueCarrier = list(type="factor", 
                         levels=c("DL", "UA", "WN")),
    Carrier = list(type = "factor",
                   levels = c("Delta Air Lines Inc.", "United Air Lines Inc.", "Southwest Airlines Co.")),
    year = list(newName = "aircraft_year")
)

rxImport(inData = odbc_data, 
         outFile = airlines_xdf, 
         colClasses = colClasses, 
         colInfo = colInfo, 
         overwrite = TRUE)

```




## Using an SQL View

```{r sql_view, eval = FALSE, size = "scriptsize"}
# In SQL
CREATE VIEW analysis_view AS 
  SELECT flight.*, 
         carrier.Description Carrier, 
         plane.* FROM flight 
  JOIN carrier
      ON flight.UniqueCarrier = carrier.Code
  LEFT OUTER JOIN plane
      ON flight.TailNum = plane.tailnum
```

\hrulefill

```{r an_RxOdbcData_with_a_view, eval=FALSE}
# In R
view_data <- RxOdbcData(table = "analysis_view", connectionString = odbc_connection)

head(view_data)
```


## External Table: Airports

\center{\Oldincludegraphics[height=6.4cm]{images/airlines_ERD_edited_2.png}}

## Airport Data
```{r read_airports}
airports <- read.csv(file.path(data_dir, "airports.csv"), stringsAsFactors=FALSE)
```

```{r examine_airports, echo=FALSE}
knitr::kable(head(read.csv(file.path(data_dir, "airports.csv"))))
```


## Merging on Airport Details

```{r merge_origin, cache = TRUE, tidy = FALSE, eval = FALSE}
# New XDF path for flights with origin location merged on
airlines_with_location <- file.path(data_dir, "airlines_with_location.xdf")

rxMerge(inData1 = airlines_xdf, inData2 = airports, 
        outFile = airlines_with_location,
        matchVars ="Origin",
        newVarNames2 = c(iata = "Origin",
                         airport = "orig_airport", city = "orig_city", state = "orig_state", 
                         lat = "orig_lat", long = "orig_long")
)

rxMerge(inData1 = airlines_with_location, inData2 = airports,
        outFile = airlines_with_location,
        matchVars = "Dest",
        newVarNames2 = c(iata = "Dest", 
                         airport = "dest_airport", city = "dest_city", state = "dest_state", 
                         lat = "dest_lat", long = "dest_long"),
        overwrite = TRUE)
```



## Map Routes for Carriers
\center{\Oldincludegraphics[height=5.7cm]{images/United_routes.png}}






## Flight Legs with Headings
\center{\Oldincludegraphics[height=5.7cm]{images/WN863.png}}






## Direction-Dependent Delay

Slightly greater delay for planes heading West.

\center{\Oldincludegraphics[height=5.5cm]{images/polar_barplot.png}}






## rxMerge Parameters: Input
  + `inData1`: either
    - a data frame
    - an XDF file name
    - an `RxXdfData` object, or
    - a list of `RxXdfData` objects
  + `inData2`
    - optional second data set
    - only used if `inData1` is not a list





## rxMerge Parameters: Output

 - `outFile`: XDF file for output (returns a dataframe if `NULL`)

 - `overwrite`: whether the output XDF file should be overwritten (if it already exists)

 - `maxRowsByCols`: limit on the total size of the dataframe returned (when `outFile = NULL`)

 - `rowsPerOutputBlock`: controls output data chunk size
    
 - `xdfCompressionLevel`: controls level of compression in the output XDF file.





## Merging

  + `type`
    - inner
    - left
    - right
    - full
    - oneToOne
    - union
  + `matchVars`
    - names columns to use as keys in the merge
    
    
    


## Renaming Columns

  + `newVarNames1`
    - specifies new names for the columns of `inData1`
  + `newVarNames2`
    - specifies new names for the columns of `inData2`
  + `duplicateVarExt`
    - suffixes to be added to any non-unique column names




## Column Selection

  + `varsToKeep1`, `varsToDrop1`
    - vector of column names to keep (or drop) from `inData1`
    - ignored if NULL
    - either use keep or drop, not both
  + `varsToKeep2`, `varsToDrop2`
    - same thing for `inData2`




## Sorting Options

  + `autoSort`
    - whether `rxMerge` should automatically sort the datasets before merging
  + `missingsLow`
    - whether missing values should be sorted as the lowest value
  + `decreasing`
    - whether values should be sorted highest-to-lowest




## Messages

  + `reportProgress`
    - 0: no progress messages
    - 1: show number of processed rows
    - 2: also show timings
    - 3: show it all
  + `verbose`
    - 0: no additional output
    - 1: show summary messages 




## Processing

  + `bufferLimit`
    - integer specifying Mb of RAM for merge buffer
    - default (-1) sets buffer size automatically
  
