

```{r setup, echo = FALSE}

rxOptions(reportProgress = 0)

library(ggplot2)

theme_set(theme_bw())


```


# Data Management with MRS:    
# Summary Statistics


## Intro

Welcome back to Data Management with Microsoft R Server. In our last
session, we learned how to transform multiple variables at once, create
transformation functions that pass results between chunks, and create and
modify complex factors.

Throughout the last few sessions, we've occasionally seen some of MRS's
summary functions in use, like `rxGetInfo` and `rxSummary`.
In this session, we'll take a closer look at MRS's functions for calculating
those summary statistics, as well as tables and correlations.




## Numeric Variables

### rxSummary

First, let's take a closer look at `rxSummary`. This is your workhorse function
for the classic univariate statistics - mean, standard deviation, minimum,
maximum, and the number of valid and missing observations.


```{r summarize_one}

# Create a pointer to one of MRS's sample XDFs
defaultData <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

# Summarize credit score
rxSummary( ~ creditScore, data = defaultData)

```

`rxSummary` uses R's formula syntax to specify which variables should be
summarized. The formula syntax is most commonly seen in regression 
functions like `lm` and `glm`, but it's a useful convention in other situations,
too. To summarize one variable, we just name it on the right-hand side of the
formula.

```{r summarize_two}

rxSummary( ~ creditScore + ccDebt, data = defaultData)

```

We can add a second variable to the summary by separating it from the first
with a plus sign.

```{r summarize_all}

rxSummary( ~ ., data = defaultData)


```

And we can summarize *all* of the variables in a dataset by putting a dot on
the right-hand side of the formula. The dot represents any variable not already
specified in the formula - in this case, that's all of them.

So far, this is straightforward. But often in the course of an analysis, we want
to use these summary statistics as inputs for another process, like plotting.
In that case, we can save the output of `rxSummary` as an R object and pass it
to other functions.

```{r summary_str}

# Assign rxSummary to an object
defaultSummary <- rxSummary( ~ ., data = defaultData)

# Check the class
str(defaultSummary)

```

If we use R's `str` command to look at what `rxsummary` returns, we can see it's
a bit more complex than what it prints to the screen. Most of this stuff is
related to the arguments we gave `rxSummary` and isn't much use. But one
element, `sDataFrame`, looks like it has what we need: a data frame in which
each variable in our data is a row, and each summary statistic is a column.
It's easy to extract this data frame and plot the stats:

```{r plotSummary}

# Extract the data frame
summaryDF <- defaultSummary$sDataFrame


# Plot the means
ggplot(summaryDF, aes(x = Name, y = Mean)) +
    geom_point(size = 10) +
    labs(x = "Variable", y = "Variable Mean")


```

This code generates a simple plot of the means, but we could go on to plot the
minimums and maximums, or to use the standard deviations to plot ranges around
the means, and so on.



### Groupwise Summary

Now, what if we wanted to know how average credit score had changed over
the years represented in this data? We can easily calculate group-wise means 
with `rxSummary`. We just have to make a couple modifications to our code:


```{r summary_by_group}

rxSummary(creditScore ~ F(year), data = defaultData)

```

First, we move `creditScore` to the *left* hand side of the formula - this is
traditionally where the dependent variable would go in a regression, and it's
how we tell `rxSummary` that *these* summary stats will be dependent on some
second variable. On the right-hand side of the formula, we add `year`. If year
were already a factor, that would be enough, but it's actually an integer - so
we use the capital-F function so that `rxSummary` will treat the year variable
as a factor. Normally we'd need to create a proper factor variable on our
dataset, but there's some behind-the-scenes magic when formulas are involved
that makes capital-F work as a shortcut.

In the output of `rxSummary`, each row now represents one category of the
right-hand variable, instead of one whole variable as before. But we can plot
the results just as easily:

```{r plotSummaryCat}

scoreByYear <- rxSummary(creditScore ~ F(year), data = defaultData)

str(scoreByYear)

summaryDFCat <- scoreByYear$categorical[[1]]

ggplot(summaryDFCat, aes(x = F_year, y = Means)) +
    geom_point(size = 10) +
    geom_line(aes(group = 1)) +
    labs(x = "Year",
         y = "Average Credit Score")


```

Accessing the data frame of results is a little different, but using R's `str` 
function again makes it fairly evident that the results have moved from the
`sDataFrame` element down to the `categorical` element. If you ever have trouble
finding the element you need inside of an R object, `str` should help you find
it.




### rxQuantile

Unfortunately, `rxSummary` can only compute the statistics we've seen so far:
mean, standard deviation, etc. What if we want to calculate quantiles, like
the median? The function for quantiles is simply `rxQuantile`. Right now,
`rxQuantile` can only calculate results for one variable at a time, so there's
no formula interface.
The syntax is very straightforward - just name the variable and the dataset:

```{r rxQuantile}

rxQuantile(varName = "creditScore", data = defaultData)

```

`rxQuantile` defaults to the 25th, 50th, and 75th quantiles, as well as the
zeroth and hundredth, which are the same as the min and max.
Other quantiles can be calculated by simply changing the `probs` argument
of `rxQuantile`. Here's an example that returns the upper and lower limits of
a 95 percent interval.

```{r nondefault_quantiles}

rxQuantile(varName = "creditScore", 
           probs = c(0.025, 0.975),
           data = defaultData)

```

Quantiles are a little more challenging to compute on big data because they
depend on all of values of the variable in question at once. Consequently,
`rxQuantile` is less flexible than `rxSummary`.




## Categorical Variables

So far, we've focused on summary statistics for numeric variables, but for
categorical variables, we typically turn to tables. MRS has two functions for 
making tables and crosstabs: `rxCrossTabs` and `rxCube`.
`rxCrossTabs` is the more traditional interface,
generating a table where each unique combination of `x` and `y` is represented
by a cell with a count of how often that combination occurs.

In `rxCube`, `x` and `y` are both columns, such that each *row* 
represents a unique combination of values in `x` and `y`, with a third column
containing the count. We'll compare the output of these two functions - but
first, let's add another categorical variable to our dataset.

```{r add_cat}

# Create a new path for this modified file
defaultTemp <- tempfile(fileext = ".xdf")

# Import the source data, adding a categorical creditScore variable
rxDataStep(inData = defaultData,
           outFile = defaultTemp,
           transforms = list(creditScoreCat = cut(creditScore,
                                                  breaks = seq(from = 400,
                                                               to = 900,
                                                               by = 100)))
)


```

I'm going to split `creditScore` into bins one hundred points wide, ranging
from four hundred to nine hundred, just so we have a third categorical variable
to work with. I'm using open-source R's `cut` function, which returns a factor;
but because I'm specifying all of the levels in advance (from four hundred, to
nine hundred, by one hundred), I don't need to use `rxFactors`.

Okay - let's take a look at the output of `rxCrossTabs` and `rxCube` for a 
single variable. We've been working with mortgage default data, but still don't
have a good idea of how many defaults there are in that data. Let's check that.

```{r table_univar}

rxCrossTabs( ~ F(default),
            data = defaultTemp)

rxCube( ~ F(default),
       data = defaultTemp)

```

`rCube` and `rxCrossTabs` both use the formula syntax mentioned earlier.
So far, the results of the two functions look about the same. With two 
variables, however, they'll start to look distinct:

```{r table_bivar}

rxCrossTabs( ~ F(default) : creditScoreCat,
            data = defaultTemp)

rxCube( ~ F(default) : creditScoreCat,
       data = defaultTemp)


```

I'm still using the formula syntax, but instead of the plus sign,
I'm using a colon to separate the variables. Roughly speaking, that means that
I'm interested in the relationship between these two variables, but that
neither is "dependent" on the other.

Take a look at the `rxCrossTabs` results. The unique values of `default` - 
zero and one - each get a row,
and the unique values of `creditScoreCat` each get a column. The cells of this
table represent each possible unique combination of `default` and 
`creditScoreCat`, and contain the counts of how often that combination occurs.

In `rxCube`, each combination of `default` and `creditScoreCat` gets
one row, and all of the counts are in a third column. At this point, both tables
are about equally easy to read, although the cube version would be easier
to use as an input for a plot or a regression. Let's see what happens when
we add `year` as a third variable.


```{r table_bivar}

rxCrossTabs( ~ F(default) : creditScoreCat : F(year),
            data = defaultTemp)

rxCube( ~ F(default) : creditScoreCat : F(year),
       data = defaultTemp)


```

To accommodate a third dimension, `rxCrossTabs` is generating a complete
crosstabulation of `default` and `creditScoreCat` for each `year`. `rxCube`, on
the other hand, has almost the same format as before: it's just added a new
column for `year`, and many new rows to account for all the new combinations.
For tables with three or more dimensions, `rxCube` 
produces much more manageable results.

There's one catch to using `rxCube` results in another function: you have to
ask `rxCube` to return a data frame. Otherwise, the `rxCube` output is just a 
list that many functions won't know what to do with.
In this example, I'm just trying to plot
the counts of each credit score category. As you can see, ggplot doesn't know
what to do with the list returned by `rxCube`.

```{r plot_cube_wrong}

creditScoreCounts <- rxCube( ~ creditScoreCat,
                            data = defaultTemp)

# It's actually a list
class(creditScoreCounts)

# ggplot doesn't know what to do with a list
ggplot(creditScoreCounts, aes(x = creditScoreCat, weights = Counts)) +
    geom_bar()

```

To solve this problem, I simply set the `returnDataFrame` argument to TRUE.
That allows the output of `rxCube` to be used directly inside of `ggplot`, or
any other function that expects a data frame.


```{r plot_cube_correct}


creditScoreCountsDF <- rxCube( ~ creditScoreCat,
                              data = defaultTemp,
                              returnDataFrame = TRUE)

# Now a data.frame
class(creditScoreCountsDF)

# No problem plotting
ggplot(creditScoreCountsDF, aes(x = creditScoreCat, weights = Counts)) +
    geom_bar()


```



Let's review: `rxCrossTabs` represents unique combinations of two or more 
variables as cells in a table. 
`rxCube` represents those unique combinations as rows in a table. 
Some people prefer the output of `rxCrossTabs` for reading, but
`rxCube` is almost always the format you want if you're going to use the results
in another function - in which case you need to set `returnDataFrame` to TRUE.



## Weighted Data

By default, these two tabulating functions treat every row of your data equally,
as one observation. But one row doesn't always correspond exactly to one 
observation. For example, census data is commonly pre-aggregated, so that
every combination of characteristics is represented by one row, with a column
indicating the frequency of that combination - just like the output of `rxCube`.
Another example is data collected using stratified sampling, where some
groups are oversampled during data collection and then weighted during analysis
to reflect their actual population prevalence.

Fortunately, many functions in MRS, including `rxSummary`, `rxCrossTabs`,
`rxCube`, and most of the statistical models, can adjust their results to
accommodate  weights. Let's look at an example. First, let's load up some US
Census data that's included with MRS:


```{r get_census}

censusData <- file.path(rxGetOption("sampleDataDir"), "CensusWorkers.xdf")

rxDataStep(censusData, numRows = 10)


```

Using `rxDataStep` to check the first few rows, I can see a variable called
"per weight" that's probably my weighting variable - but to be sure, I can 
take a quick look at the help file for this dataset:

```{r census_help}

?CensusWorkers

```

Here in the help file, I can see that each row in this dataset represents an
entire group of people with the same age, sex, location, number of weeks 
worked, and approximate income, and that the "per weight" column is a count of 
how many such people there are in this entire sample.

There are two options for using weights: `fweights` for *frequency* weighting
variables, and `pweights` for *probability* weighting variables. Since
"per weight" is a count of individuals, it's a frequency weight. Both
`rxCrossTabs` and `rxCube` return the same results, so I'll just show the
cube version:


```{r fweights}

rxCube( ~ state : sex,
       fweights = "perwt",
       data = censusData)

```

If I wanted to summarize income by state and sex, I could use `rxSummary`. I'll
just put wage variable on the left-hand side of the same formula, which gives me
a row for each combination of state and sex, with the summary stats for income
for each of those groups. `rxSummary` also returns a "SumOfWeights" variable,
which is the same as the "Counts" column that `rxCube` returned.

```{r income_summary}

rxSummary(incwage ~ state :  sex,
          fweights = "perwt",
          data = censusData)

rxSummary( ~ sex,
          fweights = "perwt",
          data = censusData)


```

As you can see, it's simple to modify your code to account for weighted data.
If you have probability weights, which are usually represented as a proportion
that sums to one across the whole dataset, you can use the function `pweights`
instead of `fweights` to achieve the same effect.



## Conclusion

That's all we have time for in this session. You should now have the tools you
need to create just about any kind of summary statistics you need, whether
your variable is quantitative or categorical, or your dataset is weighted or
unweighted. 

Take care, and see you next time.






















