

```{r setup, echo = FALSE}

rxOptions(reportProgress = 0)


```



# Data Management with RRE:
# Sorting and Deduplicating Datasets

## Intro

Welcome back to Data Management with Revolution R Enterprise. Most of the time
when we're working with big data, it's a good idea to move the data as little
as possible. The XDF file format is designed to make that easier by allowing
chunks of data to be quickly read and distributed for computations, and allowing
new data to be appended to an existing XDF without requiring a re-write of the
whole file. 

But sometimes there's no good way to avoid rearranging the entire dataset. 
For example, in our session on advanced transformations, we created a lagged
variable that required the entire dataset to be in chronological order.

Merging datasets also requires sorting them. The RRE function for merging
datasets will sort them for you automatically, but if you're planning to merge
multiple datasets together, it can be a tremendous time-saver to do all of the
sorting upfront.

Finally, removing duplicate records from a dataset also typically involves 
sorting the data first.

In RRE, sorting and deduplication are all handled by the same function: 
`rxSort`. In this session, we'll cover how to use `rxSort` for all of these
scenarios.




## Sorting

First, let's take a look at our example data. This is mortgage default data
covering ten years from 2000 to 2009. It's already sorted by year, but we'll
mix that up in a second.


```{r load_data}

xdfSource <- file.path(rxGetOption("sampleDataDir"), "mortDefaultSmall.xdf")

rxDataStep(xdfSource, numRows = 10)

```

Now, let's try out `rxSort` for a single variable. The syntax is quite simple:
I just indicate my source data file, the path where I'd like to save the
sorted data, and the name of the variable I'd like to sort on, which I pass to
the argument `sortByVars`.

```{r single_sort}

# A output file for the sorted data
xdfSorted <- tempfile(fileext = ".xdf")

rxSort(inData = xdfSource,
       outFile = xdfSorted,
       sortByVars = "creditScore"
)

rxDataStep(xdfSorted, numRows = 10)

```

As you can see, `rxSort` sorts in ascending order by default. But it's simple
to sort in *descending* order by setting the argument `decreasing` to TRUE.


```{r single_desc}

rxSort(inData = xdfSource,
       outFile = xdfSorted,
       sortByVars = "creditScore",
       decreasing = TRUE,
       overwrite = TRUE
)

rxDataStep(xdfSorted, numRows = 10)

```



### Multi-Variable Sorts

To sort on two or more variables, just name additional variables in 
`sortByVars` by creating a character vector.
This code will sort by `creditScore` first, and then by
`yearsEmploy` any time there's more than one record with the same credit score.

```{r multi_sort}

rxSort(inData = xdfSource,
       outFile = xdfSorted,
       sortByVars = c("creditScore", "yearsEmploy"),
       overwrite = TRUE
)

rxDataStep(xdfSorted, numRows = 10)

```

When we're sorting multiple variables at once, setting `decreasing` to TRUE
will cause all of the sorting variables variables to be sorted in decreasing
order.


```{r multi_desc}

rxSort(inData = xdfSource,
       outFile = xdfSorted,
       sortByVars = c("creditScore", "yearsEmploy"),
       decreasing = TRUE,
       overwrite = TRUE
)

rxDataStep(xdfSorted, numRows = 18)

```

But it's quite common to need to sort ascending on one variable, and descending
on another. To do that, pass a vector of TRUE and FALSE to `decreasing` that's
the same length as the vector you pass to `sortByVars`.
For example, if I wanted to know the highest `creditScore` for
someone with no employment history, I could sort with ascending years of
employment and descending credit score. Since I want my first variable to
ascend and my second to descend, the first value I give to `decreasing` is
FALSE, and the second is TRUE.


```{r multi_mixed}

rxSort(inData = xdfSource,
       outFile = xdfSorted,
       sortByVars = c("yearsEmploy", "creditScore"),
       decreasing = c(FALSE, TRUE),
       overwrite = TRUE
)

rxDataStep(xdfSorted, numRows = 18)

```

Wow - credit scores over 800 with no employment history; must be nice!
In practice, I'd actually use `rxSummary` to answer this question - but it's 
a good illustration of `rxSort`'s flexibility. 



## Deduplication

Besides sorting datasets, `rxSort` is also the function to turn to when you
need to remove duplicate records from a dataset. When we set the argument
`remove dupe keys` to TRUE, `rxSort` will keep only the first record for
each unique combination of the sorting variables. 


```{r default_nodup}

xdfNoDup <- tempfile(fileext = ".xdf")

rxSort(inData = xdfSource,
       outFile = xdfNoDup,
       sortByVars = "default",
       removeDupKeys = TRUE
)


rxDataStep(xdfNoDup)


```

In this example, my "key" is the default variable, which only has two unique
values - zero and one. Accordingly, I only get two rows back: the very first
zero encountered, and the very first one. All of the other records had values
of default that duplicated these - so they were removed. If we try the same
thing with the year variable, we get ten records - the first from each year.



```{r year_nodup}

rxSort(inData = xdfSource,
       outFile = xdfNoDup,
       sortByVars = "year",
       removeDupKeys = TRUE,
       overwrite = TRUE
)

rxDataStep(xdfNoDup)

```


If we add a second variable to `sortByVars`, `rxSort` will return the first
record for each unique combination of values. Since I have two unique values
in default and ten unique values in year, I get a total of twenty records back.

```{r double_nodup}

rxSort(inData = xdfSource,
       outFile = xdfNoDup,
       sortByVars = c("default", "year"),
       removeDupKeys = TRUE,
       overwrite = TRUE
)

rxDataStep(xdfNoDup)

```

Sometimes it can be useful to know how many records have been removed. For
example, if you're subsetting to the latest activity record for a customer, you
might also want to know how many activity records that customer has in total.
If so, you can use the "dupe freq var" argument to name a column that will
contain those counts. You can name the column whatever you like, but "frequency"
makes a reasonable default.

```{r dup_freq}

rxSort(inData = xdfSource,
       outFile = xdfNoDup,
       sortByVars = "year",
       removeDupKeys = TRUE,
       dupFreqVar = "freq",
       overwrite = TRUE
)

rxDataStep(xdfNoDup)

```

This output is actually quite similar to that of the function `rxCube`, which
we discussed in an earlier session, with one big exception: the output of 
`rxSort` includes *all* of the variables of the selected records, whereas
`rxCube` returns only the key and the count.

```{r cube}

rxCube( ~ F(year), data = xdfSource)

```

In most situations, the two functions take about the same amount of time, so 
you can use whichever has the most convenient output.



### Other Arguments

`rxSort` takes two other arguments that you should know about:

`missingsLow` specifies whether a missing value should be counted as the
highest value in a sort order, or the lowest. It defaults to TRUE, which means
that in our previous examples, any person with a missing creditScore would
be sorted as if they had the lowest credit score.

If you're sorting on a character variable, you can also choose whether `rxSort`
should be case-sensitive in its sorting using the `caseSensitive`
argument. By default, `caseSensitive` is set to FALSE - which means that a
capital "A" and a lower-case "A" would have the same sort order.

Be careful, though - it's easy to mistake a factor variable for a character 
if you haven't actually checked the data types.
Factors will always sort based on the order of their levels, not the text of 
their labels. So if you find yourself getting strange results when sorting
a variable, check to make sure it isn't a factor.

Conversely, if you *need* a variable to sort in a non-standard way, make it a
factor! That's what they're for.



## Conclusion

That's all for this session. While the best strategy for sorting is to simply 
avoid it, hopefully you now feel well-equipped to handle whatever sorting and 
deduplication tasks that come your way.

Thanks, and see you next time.





